---
title: "Tier 3 Functional Enrichment"
author: "Carlota Myhre de Gouveia"
date: "2026-02-21"
output:
  html_document:
    toc: true
    toc_float: true
---


## Setup and load required libraries
```{r packages, results='hide'}
library(dplyr)
library(tidyr)
library(stringr)
library(topGO)
library(readr)
library(ggplot2)
library(pheatmap)
library(purrr)
library(methods)
```

## Load EggNOG transcript‑level annotation
```{r load_eggnog_anns}
eggnog <- read.delim(
  "results/annotation/eggnog/Trinity.emapper.annotations",
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = ""
)

colnames(eggnog) <- c(
  "query_name", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
  "max_annot_lvl", "COG_category", "Description", "Preferred_name",
  "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
  "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
  "CAZy", "BiGG_Reaction", "PFAMs"
)

eggnog
```

## Build gene-level GO annotation table from EggNOG output
```{r build_gene_go_mapping}

eggnog_go <- eggnog %>%
  dplyr::filter(GOs != "" & !is.na(GOs)) %>%                 # keep only annotated rows
  dplyr::select(trinity_id = query_name, GOs) %>%            # keep relevant columns
  dplyr::mutate(
    trinity_id = sub("\\.p[0-9]+$", "", trinity_id),         # remove protein suffix
    gene_id    = sub("_i[0-9]+$", "", trinity_id)            # collapse isoforms → gene
  )

## Expand GO terms into one row per gene–GO pair
gene2go <- eggnog_go %>%
  tidyr::separate_rows(GOs, sep = ",|;") %>%
  dplyr::mutate(GOs = stringr::str_trim(GOs)) %>%
  dplyr::filter(GOs != "") %>%
  dplyr::distinct(gene_id, GOs)

## Build topGO universe and mapping
universe_ids <- unique(gene2go$gene_id)

geneList_template <- factor(
  x = rep(0, length(universe_ids)),
  levels = c(0, 1)
)
names(geneList_template) <- universe_ids

geneID2GO <- tapply(gene2go$GOs, gene2go$gene_id, unique)

head(gene2go)
```

## Define Tier 3 DE file paths
```{r define_files}
## Define Tier 3 DE file paths
tier3_files <- list(
  misexpressed_12C   = "results/DE/tier3_DE/misexpressed/misexpressed_12C.tsv",
  misexpressed_15C   = "results/DE/tier3_DE/misexpressed/misexpressed_15C.tsv",
  misexpressed_18C   = "results/DE/tier3_DE/misexpressed/misexpressed_18C.tsv",
  parent_like_S_12C  = "results/DE/tier3_DE/parent_like/parent_like_S_12C.tsv",
  parent_like_S_15C  = "results/DE/tier3_DE/parent_like/parent_like_S_15C.tsv",
  parent_like_S_18C  = "results/DE/tier3_DE/parent_like/parent_like_S_18C.tsv",
  parent_like_W_12C  = "results/DE/tier3_DE/parent_like/parent_like_W_12C.tsv",
  parent_like_W_15C  = "results/DE/tier3_DE/parent_like/parent_like_W_15C.tsv",
  parent_like_W_18C  = "results/DE/tier3_DE/parent_like/parent_like_W_18C.tsv"
)

## Inspect Tier 3 DE files
tier3_inspection <- lapply(names(tier3_files), function(lbl) {
  df <- read.delim(tier3_files[[lbl]], header = FALSE)
  colnames(df) <- "transcript_id"

  list(
    category = lbl,
    n_rows = nrow(df),
    columns = names(df),
    preview = head(df)
  )
})

tier3_inspection
```

## Check annotation coverage
```{r ann_cov}
## Check annotation coverage for Tier 3 categories
annotation_summary <- lapply(names(tier3_files), function(lbl) {
  df <- read.delim(tier3_files[[lbl]], header = FALSE)
  colnames(df) <- "transcript_id"
  ids <- df$transcript_id

  annotated <- sum(ids %in% universe_ids)
  total <- length(ids)

  data.frame(
    category = lbl,
    total_transcripts = total,
    annotated_transcripts = annotated,
    percent_annotated = round(annotated / total * 100, 2)
  )
})

annotation_summary <- bind_rows(annotation_summary)

annotation_summary
```

## GO Enrichment Function
```{r enrich_func}

## Run GO enrichment for a given set of transcripts
run_go_enrichment <- function(transcript_ids, ontology, geneID2GO, geneList_template) {
  geneList <- geneList_template
  geneList[names(geneList) %in% transcript_ids] <- 1L

  GOdata <- new(
    "topGOdata",
    ontology = ontology,
    allGenes = geneList,
    annot = annFUN.gene2GO,
    gene2GO = geneID2GO,
    nodeSize = 10
  )

  result <- runTest(GOdata, algorithm = "weight01", statistic = "fisher")
  table <- GenTable(GOdata, weightFisher = result, topNodes = length(score(result)))

  table
}
```

## Run enrichment
```{r run_enrich}
## Run GO enrichment for all Tier 3 categories
tier3_enrichment <- lapply(names(tier3_files), function(lbl) {
  df <- read.delim(tier3_files[[lbl]], header = FALSE)
  colnames(df) <- "transcript_id"

  bp <- run_go_enrichment(df$transcript_id, "BP", geneID2GO, geneList_template)
  mf <- run_go_enrichment(df$transcript_id, "MF", geneID2GO, geneList_template)
  cc <- run_go_enrichment(df$transcript_id, "CC", geneID2GO, geneList_template)

  list(
    category = lbl,
    BP = bp,
    MF = mf,
    CC = cc
  )
})

tier3_enrichment
```
## Save GO enrichment results
```{r save_GO_enrich}
output_dir <- "results/functional_enrichment/tier_3"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

for (res in tier3_enrichment) {
  lbl <- res$category

  write.csv(res$BP, file = file.path(output_dir, paste0(lbl, "_BP.csv")), row.names = FALSE)
  write.csv(res$MF, file = file.path(output_dir, paste0(lbl, "_MF.csv")), row.names = FALSE)
  write.csv(res$CC, file = file.path(output_dir, paste0(lbl, "_CC.csv")), row.names = FALSE)
}
```

## Filter significant GO terms
```{r}
tier3_significant <- lapply(tier3_enrichment, function(res) {
  bp <- res$BP %>%
    mutate(p_value = as.numeric(weightFisher)) %>%
    filter(p_value < 0.05)

  mf <- res$MF %>%
    mutate(p_value = as.numeric(weightFisher)) %>%
    filter(p_value < 0.05)

  cc <- res$CC %>%
    mutate(p_value = as.numeric(weightFisher)) %>%
    filter(p_value < 0.05)

  list(
    category = res$category,
    BP = bp,
    MF = mf,
    CC = cc
  )
})

tier3_significant
```

## Display significant BP terms in the report
```{r}
for (res in tier3_significant) {
  cat("### Significant BP terms for:", res$category, "\n")
  bp <- res$BP
  if (nrow(bp) > 0) {
    print(
      bp %>%
        arrange(p_value) %>%
        head(20)
    )
  } else {
    cat("No significant BP terms.\n")
  }
  cat("\n\n")
}
```

## Save summary tables of significant GO terms
```{r}
summary_dir <- file.path(output_dir, "summary_tables")
dir.create(summary_dir, recursive = TRUE, showWarnings = FALSE)

for (res in tier3_significant) {
  lbl <- res$category

  write.csv(res$BP, file = file.path(summary_dir, paste0(lbl, "_BP_significant.csv")), row.names = FALSE)
  write.csv(res$MF, file = file.path(summary_dir, paste0(lbl, "_MF_significant.csv")), row.names = FALSE)
  write.csv(res$CC, file = file.path(summary_dir, paste0(lbl, "_CC_significant.csv")), row.names = FALSE)
}
```


## Dotplot function for enriched GO terms
```{r dotplot_func}
plot_go_dotplot <- function(go_table, title) {
  go_table %>%
    arrange(p_value) %>%
    slice_head(n = 10) %>%        # show only top 10 terms
    mutate(
      Term = make.unique(Term),
      Term = factor(Term, levels = rev(Term)),
      log_p = -log10(p_value)
    ) %>%
    ggplot(aes(x = log_p, y = Term)) +
    geom_point(size = 3, color = "#2C7BB6") +
    labs(
      x = "-log10(p-value)",
      y = "",
      title = title
    ) +
    theme_minimal(base_size = 12)
}
```

```{r}
## --- 1. Name the list elements by category ---
names(tier3_significant) <- sapply(tier3_significant, function(x) x$category)


## --- 2. Function to extract top GO terms (BP only) ---
extract_top_terms <- function(sig_list, n = 10) {
  lapply(sig_list, function(res) {
    res$BP %>%
      arrange(p_value) %>%
      slice_head(n = n) %>%
      dplyr::select(GO.ID, Term, p_value)
  })
}


## --- 3. Extract top 10 terms per category ---
top_terms <- extract_top_terms(tier3_significant, n = 10)


## --- 4. Print them nicely in the knitted report ---
for (cat in names(top_terms)) {
  cat("### Top BP terms for:", cat, "\n")
  print(top_terms[[cat]])
  cat("\n\n")
}
```


## Generate dotplots for each Tier 3 category
```{r dotplots_enrich}
dotplot_dir <- file.path(output_dir, "dotplots")
dir.create(dotplot_dir, recursive = TRUE, showWarnings = FALSE)

for (res in tier3_significant) {
  bp <- res$BP
  if (nrow(bp) > 0) {
    p <- plot_go_dotplot(bp, paste("BP Enrichment:", res$category))

    # Display in the knitted report
    print(p)

    # Save to file
    ggsave(
      filename = file.path(dotplot_dir, paste0(res$category, "_BP_dotplot.png")),
      plot = p,
      width = 8,
      height = 6
    )
  }
}
```

## helper to shorten GO term names
```{r}
shorten_terms <- function(x, max = 40) {
  ifelse(nchar(x) > max, paste0(substr(x, 1, max), "..."), x)
}
```


## Heatmap function of shared GO terms across categories
```{r heatmap_func}
build_heatmap_matrix <- function(sig_list, categories, n = 10) {
  mats <- list()

  for (cat in categories) {
    bp <- sig_list[[cat]]$BP %>%
      dplyr::arrange(p_value) %>%
      dplyr::slice_head(n = n) %>%
      dplyr::mutate(Term = shorten_terms(Term))

    df <- bp %>%
      dplyr::select(Term, p_value)

    rownames(df) <- df$Term
    df$Term <- NULL

    mats[[cat]] <- df
  }

  combined <- do.call(cbind, mats)
  colnames(combined) <- categories
  combined
}
```

## Generate heatmaps for misexpressed, parent‑like South, parent‑like West
```{r heatmap_enrich}
heatmap_dir <- file.path(output_dir, "heatmaps")
dir.create(heatmap_dir, recursive = TRUE, showWarnings = FALSE)

misexp_cats <- c("misexpressed_12C", "misexpressed_15C", "misexpressed_18C")
south_cats  <- c("parent_like_S_12C", "parent_like_S_15C", "parent_like_S_18C")
west_cats   <- c("parent_like_W_12C", "parent_like_W_15C", "parent_like_W_18C")

mat_misexp <- build_heatmap_matrix(tier3_significant, misexp_cats)
mat_south  <- build_heatmap_matrix(tier3_significant, south_cats)
mat_west   <- build_heatmap_matrix(tier3_significant, west_cats)
```


```{r}
# Misexpressed
pheatmap(mat_misexp, fontsize_row = 8, fontsize_col = 10)
pheatmap(mat_misexp,
         filename = file.path(heatmap_dir, "misexpressed_heatmap.png"),
         fontsize_row = 8, fontsize_col = 10)
```
```{r}

# Parent-like South
pheatmap(mat_south, fontsize_row = 8, fontsize_col = 10)
pheatmap(mat_south,
         filename = file.path(heatmap_dir, "parent_like_S_heatmap.png"),
         fontsize_row = 8, fontsize_col = 10)

```
```{r}
# Parent-like West
pheatmap(mat_west, fontsize_row = 8, fontsize_col = 10)
pheatmap(mat_west,
         filename = file.path(heatmap_dir, "parent_like_W_heatmap.png"),
         fontsize_row = 8, fontsize_col = 10)
```
```{r}
############################################################
## 1. Helper to shorten GO term names
############################################################

shorten_terms <- function(x, max = 40) {
  ifelse(nchar(x) > max, paste0(substr(x, 1, max), "..."), x)
}


############################################################
## 2. Build a combined dataframe for dotplot
############################################################

build_dotplot_df <- function(sig_list, categories, n = 10) {
  df_list <- list()

  for (cat in categories) {
    bp <- sig_list[[cat]]$BP

    # extract temperature from name (e.g., "misexpressed_12C")
    temp <- sub(".*_(\\d+C)$", "\\1", cat)

    # extract category type (misexpressed, parent_like_S, parent_like_W)
    type <- sub("_(\\d+C)$", "", cat)

    # sort by p-value and take top n
    bp <- bp[order(bp$p_value), ]
    bp <- bp[1:n, ]

    # shorten GO terms
    bp$Term <- shorten_terms(bp$Term)

    # compute gene ratio
    bp$GeneRatio <- bp$Significant / bp$Annotated

    # add metadata
    bp$Category <- type
    bp$Temperature <- temp

    df_list[[cat]] <- bp
  }

  do.call(rbind, df_list)
}


############################################################
## 3. Define categories and build the dotplot dataframe
############################################################

misexp_cats <- c("misexpressed_12C", "misexpressed_15C", "misexpressed_18C")
south_cats  <- c("parent_like_S_12C", "parent_like_S_15C", "parent_like_S_18C")
west_cats   <- c("parent_like_W_12C", "parent_like_W_15C", "parent_like_W_18C")

all_cats <- c(misexp_cats, south_cats, west_cats)

dot_df <- build_dotplot_df(tier3_significant, all_cats, n = 10)


############################################################
## 4. Combined dotplot with facets (the meaningful figure)
############################################################

library(ggplot2)

ggplot(dot_df, aes(x = Category, 
                   y = Term,
                   size = GeneRatio,
                   color = -log10(p_value))) +
  geom_point() +
  scale_color_gradient(low = "lightblue", high = "red") +
  scale_size(range = c(2, 8)) +
  facet_wrap(~ Temperature, ncol = 1) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "GO Enrichment Across Categories and Temperatures",
    x = "Category",
    y = "GO Term",
    color = "-log10(p-value)",
    size = "Gene Ratio"
  )
```
```{r}
############################################################
## 0. Setup: directories and category names
############################################################

output_dir <- "results/functional_enrichment/tier_3"

misexp_cats <- c("misexpressed_12C", "misexpressed_15C", "misexpressed_18C")
south_cats  <- c("parent_like_S_12C", "parent_like_S_15C", "parent_like_S_18C")
west_cats   <- c("parent_like_W_12C", "parent_like_W_15C", "parent_like_W_18C")

all_cats <- c(misexp_cats, south_cats, west_cats)


############################################################
## 1. Reload saved GO enrichment results (FAST)
############################################################

tier3_enrichment <- lapply(all_cats, function(lbl) {
  bp <- read.csv(file.path(output_dir, paste0(lbl, "_BP.csv")))
  mf <- read.csv(file.path(output_dir, paste0(lbl, "_MF.csv")))
  cc <- read.csv(file.path(output_dir, paste0(lbl, "_CC.csv")))

  list(
    category = lbl,
    BP = bp,
    MF = mf,
    CC = cc
  )
})

names(tier3_enrichment) <- all_cats


############################################################
## 2. Filter significant GO terms (FAST)
############################################################

tier3_significant <- lapply(tier3_enrichment, function(res) {
  bp <- res$BP %>%
    mutate(p_value = as.numeric(weightFisher)) %>%
    filter(p_value < 0.05)

  mf <- res$MF %>%
    mutate(p_value = as.numeric(weightFisher)) %>%
    filter(p_value < 0.05)

  cc <- res$CC %>%
    mutate(p_value = as.numeric(weightFisher)) %>%
    filter(p_value < 0.05)

  list(
    category = res$category,
    BP = bp,
    MF = mf,
    CC = cc
  )
})

names(tier3_significant) <- all_cats


############################################################
## 3. Helper to shorten GO term names
############################################################

shorten_terms <- function(x, max = 40) {
  ifelse(nchar(x) > max, paste0(substr(x, 1, max), "..."), x)
}


############################################################
## 4. Build full dataframe for all categories
############################################################

build_full_df <- function(sig_list, categories) {
  df_list <- list()

  for (cat in categories) {
    bp <- sig_list[[cat]]$BP

    # extract metadata
    temp <- sub(".*_(\\d+C)$", "\\1", cat)
    type <- sub("_(\\d+C)$", "", cat)

    # shorten GO terms
    bp$Term <- shorten_terms(bp$Term)

    # compute gene ratio
    bp$GeneRatio <- bp$Significant / bp$Annotated

    # add metadata
    bp$Category <- type
    bp$Temperature <- temp

    df_list[[cat]] <- bp
  }

  do.call(rbind, df_list)
}

full_df <- build_full_df(tier3_significant, all_cats)


############################################################
## 5. Select the TOP 10 GO TERMS overall (lowest p-values)
############################################################

full_df <- full_df[order(full_df$p_value), ]
top_terms <- unique(full_df$Term)[1:10]
dot_df <- full_df[full_df$Term %in% top_terms, ]


############################################################
## 6. Publication-ready dotplot
############################################################

library(ggplot2)

ggplot(dot_df, aes(x = Category, 
                   y = Term,
                   size = GeneRatio,
                   color = -log10(p_value))) +
  geom_point() +
  scale_color_gradient(low = "lightblue", high = "red") +
  scale_size(range = c(2, 8)) +
  facet_wrap(~ Temperature, ncol = 1) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Top 10 GO Terms Across All Categories and Temperatures",
    x = "Category",
    y = "GO Term",
    color = "-log10(p-value)",
    size = "Gene Ratio"
  )
```

```{r}
ggplot(dot_df, aes(x = Temperature, 
                   y = Term,
                   size = GeneRatio,
                   color = -log10(p_value))) +
  geom_point() +
  scale_color_gradient(low = "lightblue", high = "red") +
  scale_size(range = c(2, 8)) +
  facet_wrap(~ Category, ncol = 1) +   # <-- FACET BY CATEGORY
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Top 10 GO Terms Across Temperatures for Each Category",
    x = "Temperature",
    y = "GO Term",
    color = "-log10(p-value)",
    size = "Gene Ratio"
  )
```



```{r}
############################################################
## 1. Polished dotplots for Tier 3 (BP only)
############################################################

# Correct relative path to your summary tables
bp_files <- list.files(
  "results/functional_enrichment/tier_3/summary_tables",
  pattern = "_BP_significant.csv",
  full.names = TRUE
)

# Dotplot function with duplicate-term fix
plot_dotplot <- function(file, title) {
  df <- read.csv(file)

  df %>%
    arrange(weightFisher) %>%
    distinct(Term, .keep_all = TRUE) %>%   # FIX: remove duplicated GO terms
    slice_head(n = 15) %>%
    mutate(Term = factor(Term, levels = rev(Term))) %>%
    ggplot(aes(x = -log10(weightFisher), y = Term, size = Significant)) +
    geom_point(color = "#2C7BB6") +
    labs(
      x = "-log10(p-value)",
      y = "",
      title = title,
      size = "Significant genes"
    ) +
    theme_minimal(base_size = 12)
}

# Print all dotplots in RStudio
for (f in bp_files) {
  title <- gsub("_BP_significant.csv", "", basename(f))
  p <- plot_dotplot(f, paste("Tier 3 —", title))
  print(p)
}
```


```{r}
############################################################
## 2. Barplot of number of enriched GO terms
############################################################

count_terms <- function(file) {
  df <- read.csv(file)
  tibble(
    category = gsub("_BP_significant.csv", "", basename(file)),
    n_terms = nrow(df)
  )
}

term_counts <- map_df(bp_files, count_terms)

ggplot(term_counts, aes(x = category, y = n_terms, fill = n_terms)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#ABD9E9", high = "#2C7BB6") +
  labs(
    title = "Tier 3 — Number of Enriched GO:BP Terms",
    x = "",
    y = "Number of significant GO terms"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
############################################################
## Tier 3 — Readable heatmap of GO terms across temperatures
############################################################

library(tidyverse)
library(pheatmap)
library(purrr)

# Load all significant BP files
bp_files <- list.files(
  "results/functional_enrichment/tier_3/summary_tables",
  pattern = "_BP_significant.csv",
  full.names = TRUE
)

# Read and combine all significant GO terms
all_bp <- map_df(bp_files, function(f) {
  df <- read.csv(f)
  df$category <- gsub("_BP_significant.csv", "", basename(f))
  df
})

# Remove duplicated GO terms within each category
all_bp <- all_bp %>% distinct(Term, category, .keep_all = TRUE)

# Build matrix: rows = GO terms, columns = categories
mat <- all_bp %>%
  dplyr::select(Term, category, weightFisher) %>%   # <-- EXPLICIT dplyr::select FIX
  mutate(value = -log10(weightFisher)) %>%
  dplyr::select(Term, category, value) %>%
  tidyr::pivot_wider(
    names_from = category,
    values_from = value,
    values_fill = 0
  )

# Convert to matrix
mat2 <- as.matrix(mat[,-1])
rownames(mat2) <- mat$Term

# Plot heatmap
pheatmap(
  mat2,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  main = "Tier 3 — GO:BP Enrichment Across Temperatures",
  fontsize_row = 7,
  fontsize_col = 10,
  color = colorRampPalette(c("white", "#ABD9E9", "#2C7BB6"))(100)
)
```

