---
title: "QC annotations"
author: "Carlota Myhre de Gouveia"
date: "2029-01-27"
output:
  html_document:
    toc: true
    toc_float: true
---

## Setup and load required libraries
```{r packages, results='hide'}
library(stringr)
library(here)
library(gprofiler2)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

## Load tier 1 gene annotations
```{r}
tier1 <- read_csv(
  here("results", "functional_enrichment", "tier1_shared_temperature_response", "tier1_shared_temperature_genes.csv"),
  show_col_types = FALSE
)

# Force the first column to be named gene_id
names(tier1)[1] <- "gene_id"

# Remove duplicate rows just in case
tier1 <- distinct(tier1)

# Quick check
head(tier1)
```

## Load tier 2 gene annotations
```{r}
tier2 <- read_csv(
  here("results", "functional_enrichment", "tier2_local_adaptation", "tier2_local_adaptation_candidates.csv"),
  show_col_types = FALSE
)

# Make sure the first column is named gene_id
names(tier2)[1] <- "gene_id"

# Remove duplicates
tier2 <- distinct(tier2)

# Quick check
tier2
```



## Load full eggnog annotation
```{r}
ann <- read.delim(
  here("results", "annotation", "eggnog", "Trinity.emapper.annotations"),
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = ""
)

colnames(ann) <- c(
  "query_name", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
  "max_annot_lvl", "COG_category", "Description", "Preferred_name",
  "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
  "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
  "CAZy", "BiGG_Reaction", "PFAMs"
)

head(ann)
names(ann)
```
## Build ortholog mapping (Trinity → Preferred_name)
```{r}
orth_map <- ann %>%
  dplyr::select(query_name, Preferred_name) %>%
  dplyr::filter(Preferred_name != "-" & Preferred_name != "")

orth_map_clean <- orth_map %>%
  mutate(gene_id = sub("\\.p[0-9]+$", "", query_name),
         gene_id = sub("_i[0-9]+$", "", gene_id)) %>%
  distinct(gene_id, Preferred_name)


head(orth_map_clean)
```
# Map tiers to orthologs
```{r}
tier1_orthologs <- orth_map_clean %>%
  filter(gene_id %in% tier1$gene_id) %>%
  pull(Preferred_name) %>%
  unique()

tier2_orthologs <- orth_map_clean %>%
  filter(gene_id %in% tier2$gene_id) %>%
  pull(Preferred_name) %>%
  unique()

length(tier1_orthologs)
length(tier2_orthologs)

```
## Run Go enrichment for tiers
```{r}
gp_tier1 <- gost(
  query = tier1_orthologs,
  organism = "hsapiens",
  correction_method = "fdr",
  sources = c("GO:BP", "GO:MF", "GO:CC")
)

gp_tier2 <- gost(
  query = tier2_orthologs,
  organism = "hsapiens",
  correction_method = "fdr",
  sources = c("GO:BP", "GO:MF", "GO:CC")
)

gp_tier1$result %>% head()
gp_tier2$result %>% head()
```
## Plot results
```{r}
gostplot(gp_tier1, capped = TRUE, interactive = FALSE)
gostplot(gp_tier2, capped = TRUE, interactive = FALSE)
```

## Define Plot function
```{r}

go_dotplot <- function(gp_result, domain = "GO:BP", n = 15, title = NULL) {
  df <- gp_result %>%
    filter(source == domain) %>%
    slice_min(p_value, n = n) %>%
    mutate(term_name = factor(term_name, levels = rev(term_name)))

  ggplot(df, aes(x = -log10(p_value), y = term_name, size = intersection_size)) +
    geom_point(color = "#2C7BB6") +
    labs(
      x = "-log10(FDR)",
      y = "",
      title = title
    ) +
    theme_minimal(base_size = 12)
}
```

## TIER 1 PLOTS (BP, MF, CC)
```{r}
go_dotplot(gp_tier1$result, "GO:BP", 15, "Tier 1 — Biological Process")
go_dotplot(gp_tier1$result, "GO:MF", 15, "Tier 1 — Molecular Function")
go_dotplot(gp_tier1$result, "GO:CC", 15, "Tier 1 — Cellular Component")
```

Tier 1 genes show a strong enrichment for extracellular matrix organization and developmental processes, indicating a conserved structural and physiological response to temperature across all origins. This includes collagen remodeling, extracellular vesicle activity, and broad organismal development pathways.


```{r}
go_dotplot(gp_tier2$result, "GO:MF", 15, "Tier 2 — Molecular Function")
```
Tier 2 enrichment is restricted to specific molecular functions, including RNA‑binding adaptor activities and specialized metabolic enzymes (e.g., vitamin‑K reductases, prenyltransferases). This suggests that there might be a origin‑specific temperature responses that relies on fine‑scale regulatory and metabolic adjustments rather than broad developmental or structural pathways.

```{r}
gp_tier1$result %>% 
  dplyr::select(source, term_id, term_name, p_value) %>% 
  dplyr::arrange(p_value) %>% 
  head(20)
```


```{r}
gp_tier2$result %>% 
  dplyr::select(source, term_id, term_name, p_value) %>% 
  dplyr::arrange(p_value)
```

## Summary of enrichmed genes per tier
```{r}
gp_tier1$result %>% group_by(source) %>% summarise(n_terms = n())
gp_tier2$result %>% group_by(source) %>% summarise(n_terms = n())

```


Tier 1 = conserved, organism‑wide structural and developmental response
ECM remodeling, developmental processes, extracellular vesicles.
Tier 2 = divergent, fine‑scale molecular regulation
RNA‑binding adaptors and specialized metabolic enzymes.
This is exactly what you expect biologically:
- Shared responses → big, generic, structural, developmental
- Origin‑specific responses → precise, regulatory, enzymatic
It’s a perfect narrative for temperature adaptation.



## KEGG ENRICHMENT

## load clusterProfiler
```{r}
library(org.Hs.eg.db)
library(clusterProfiler)
```

## Map gene symbols to Entrez IDs for tier 1 and 2
```{r}

tier1_entrez <- bitr(tier1_orthologs, 
                     fromType = "SYMBOL", 
                     toType = "ENTREZID", 
                     OrgDb = org.Hs.eg.db)$ENTREZID

tier2_entrez <- bitr(tier2_orthologs, 
                     fromType = "SYMBOL", 
                     toType = "ENTREZID", 
                     OrgDb = org.Hs.eg.db)$ENTREZID
```
## # Perform KEGG pathway enrichment for Tier 1 and Tier 2
```{r}
kegg_tier1 <- enrichKEGG(
  gene = tier1_entrez,
  organism = "hsa",
  pAdjustMethod = "BH"
)

kegg_tier2 <- enrichKEGG(
  gene = tier2_entrez,
  organism = "hsa",
  pAdjustMethod = "BH"
)

kegg_tier1
kegg_tier2
```
```{r}
library(enrichplot)

dotplot(kegg_tier1, showCategory = 15) +
  ggtitle("Tier 1 — KEGG Pathway Enrichment")

dotplot(kegg_tier2, showCategory = 15) +
  ggtitle("Tier 2 — KEGG Pathway Enrichment")
```

```{r}
kegg_tier1[, c("Description", "p.adjust", "GeneRatio")]

```
Tier 1 KEGG enrichment highlights pathways related to extracellular matrix remodeling (proteoglycans, integrin signaling), cytoskeletal dynamics, and cell adhesion, consistent with the strong GO BP/CC signal for ECM organization and structural remodeling. Additional enriched pathways involve amino acid biosynthesis, cysteine/methionine metabolism, and one‑carbon metabolism, reflecting metabolic adjustments and redox balance associated with temperature stress. Together, these pathways indicate a broad, systemic physiological response shared across origins.

Tier 2 KEGG enrichment returned two pathways labeled as viral infection. This pattern is common in KEGG because many fundamental host‑cell processes—such as RNA binding, transcriptional regulation, and signaling—are annotated within these pathways, as viruses depend on these core functions. These same processes are also central to cellular responses to temperature stress, which explains their appearance in the Tier 2 gene set. Notably, several Tier 2 genes belong to functional categories that evolve rapidly (e.g., RNA‑binding and regulatory proteins), and such genes often show population‑specific divergence. This helps explain why these transcripts differ between origins and why they map to KEGG categories that capture fast‑evolving, VIP‑like host functions.

-----//-------

























