---
title: "Functional Enrichment"
author: "Carlota Myhre de Gouveia"
date: "2026-03-18"
output:
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


# Setup and Package Loading
```{r packages}
# Load required packages
suppressPackageStartupMessages({
  library(stringr)
  library(gprofiler2)
  library(tidyverse)
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(methods)
  library(org.Hs.eg.db)
  library(clusterProfiler)
  library(enrichplot)
})
```

# Data Import and Inspection
```{r load_and_inspect_data}
# Helper to load a tier file consistently
load_tier <- function(path) {
  df <- read_csv(path, show_col_types = FALSE)
  names(df)[1] <- "gene_id"
  df <- distinct(df)
  return(df)
}

# Helper to collapse Trinity IDs to gene IDs (for Tier 3 DE outputs)
extract_gene_id <- function(x) {
  x %>%
    str_remove("\\.p\\d+$") %>%   # remove .p1, .p2
    str_remove("_i\\d+$")         # remove isoform _i1, _i2
}

# -------------------------------
# Load Tier 1 and Tier 2 gene sets
# -------------------------------
tier1  <- load_tier("results/functional_enrichment/tier1_shared_temperature_response/tier1_shared_temperature_genes.csv")
tier2  <- load_tier("results/functional_enrichment/tier2_pop_divergence/tier2_pop_divergence_genes.csv")
tier2b <- load_tier("results/functional_enrichment/tier2_pop_divergence/tier2b_local_adaptation/tier2_local_adaptation_genes.csv")

# -------------------------------
# Load Tier 3 DE outputs
# -------------------------------
tier3_mis_12C <- extract_gene_id(read_tsv("results/DE/tier3_DE/misexpressed/misexpressed_12C.tsv", col_names = FALSE)$X1)
tier3_mis_15C <- extract_gene_id(read_tsv("results/DE/tier3_DE/misexpressed/misexpressed_15C.tsv", col_names = FALSE)$X1)
tier3_mis_18C <- extract_gene_id(read_tsv("results/DE/tier3_DE/misexpressed/misexpressed_18C.tsv", col_names = FALSE)$X1)

tier3_parent_S_12C <- extract_gene_id(read_tsv("results/DE/tier3_DE/parent_like/parent_like_S_12C.tsv", col_names = FALSE)$X1)
tier3_parent_S_15C <- extract_gene_id(read_tsv("results/DE/tier3_DE/parent_like/parent_like_S_15C.tsv", col_names = FALSE)$X1)
tier3_parent_S_18C <- extract_gene_id(read_tsv("results/DE/tier3_DE/parent_like/parent_like_S_18C.tsv", col_names = FALSE)$X1)

tier3_parent_W_12C <- extract_gene_id(read_tsv("results/DE/tier3_DE/parent_like/parent_like_W_12C.tsv", col_names = FALSE)$X1)
tier3_parent_W_15C <- extract_gene_id(read_tsv("results/DE/tier3_DE/parent_like/parent_like_W_15C.tsv", col_names = FALSE)$X1)
tier3_parent_W_18C <- extract_gene_id(read_tsv("results/DE/tier3_DE/parent_like/parent_like_W_18C.tsv", col_names = FALSE)$X1)


# -------------------------------
# Load merged Tier 3 sets
# -------------------------------
tier3_mis_all       <- load_tier("results/functional_enrichment/tier3_hybrid_inheritance/tier3_misexpressed_genes.csv")
tier3_parent_S_all  <- load_tier("results/functional_enrichment/tier3_hybrid_inheritance/tier3_parent_like_S_genes.csv")
tier3_parent_W_all  <- load_tier("results/functional_enrichment/tier3_hybrid_inheritance/tier3_parent_like_W_genes.csv")

# -------------------------------
# Load intersections
# -------------------------------
tier1_tier3_mis       <- load_tier("results/functional_enrichment/tier_intersections/tier1_tier3_misexpressed.csv")
tier1_tier3_parent_S  <- load_tier("results/functional_enrichment/tier_intersections/tier1_tier3_parent_like_S.csv")
tier1_tier3_parent_W  <- load_tier("results/functional_enrichment/tier_intersections/tier1_tier3_parent_like_W.csv")

tier2b_tier3_mis      <- load_tier("results/functional_enrichment/tier_intersections/tier2b_tier3_misexpressed.csv")
tier2b_tier3_parent_S <- load_tier("results/functional_enrichment/tier_intersections/tier2b_tier3_parent_like_S.csv")
tier2b_tier3_parent_W <- load_tier("results/functional_enrichment/tier_intersections/tier2b_tier3_parent_like_W.csv")

# -------------------------------
# Define all tiers once here
# -------------------------------
tiers <- list(
  "Tier1"                  = tier1$gene_id,
  "Tier2"                  = tier2$gene_id,
  "Tier2b"                 = tier2b$gene_id,
  "Tier3_mis_12C"          = tier3_mis_12C,
  "Tier3_mis_15C"          = tier3_mis_15C,
  "Tier3_mis_18C"          = tier3_mis_18C,
  "Tier3_parent_S_12C"     = tier3_parent_S_12C,
  "Tier3_parent_S_15C"     = tier3_parent_S_15C,
  "Tier3_parent_S_18C"     = tier3_parent_S_18C,
  "Tier3_parent_W_12C"     = tier3_parent_W_12C,
  "Tier3_parent_W_15C"     = tier3_parent_W_15C,
  "Tier3_parent_W_18C"     = tier3_parent_W_18C,
  "Tier3_mis_all"          = tier3_mis_all$gene_id,
  "Tier3_parent_S_all"     = tier3_parent_S_all$gene_id,
  "Tier3_parent_W_all"     = tier3_parent_W_all$gene_id,
  "Tier1∩Tier3_mis"        = tier1_tier3_mis$gene_id,
  "Tier1∩Tier3_parent_S"   = tier1_tier3_parent_S$gene_id,
  "Tier1∩Tier3_parent_W"   = tier1_tier3_parent_W$gene_id,
  "Tier2b∩Tier3_mis"       = tier2b_tier3_mis$gene_id,
  "Tier2b∩Tier3_parent_S"  = tier2b_tier3_parent_S$gene_id,
  "Tier2b∩Tier3_parent_W"  = tier2b_tier3_parent_W$gene_id
)

```

## Load full eggnog annotation
```{r}
ann <- read.delim(
  "results/annotation/eggnog/Trinity.emapper.annotations",
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = ""
)

colnames(ann) <- c(
  "query_name", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
  "max_annot_lvl", "COG_category", "Description", "Preferred_name",
  "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
  "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
  "CAZy", "BiGG_Reaction", "PFAMs"
)

head(ann)
names(ann)
```

## Build ortholog mapping (Trinity → Preferred_name)
```{r}
# Start from annotation table
orth_map <- ann %>%
  dplyr::select(query_name, seed_ortholog, evalue, score, Preferred_name) %>%
  dplyr::filter(Preferred_name != "-" & Preferred_name != "")

# Clean Trinity IDs down to gene_id and keep evalue/score
orth_map_clean <- orth_map %>%
  mutate(
    gene_id = sub("\\.p[0-9]+$", "", query_name),   # remove .p1, .p2
    gene_id = sub("_i[0-9]+$", "", gene_id)         # remove isoform _i1, _i2
  ) %>%
  filter(Preferred_name != "-" & Preferred_name != "") %>%
  distinct(gene_id, Preferred_name, evalue, score)

# Keep ALL valid orthologs per gene (no collapsing to best)
orth_map_all <- orth_map_clean %>%
  distinct(gene_id, Preferred_name, evalue, score)

# Quick inspection
head(orth_map_all)
```

## Map tiers to orthologs
```{r}
# Helper function: map a tier tibble/vector to orthologs
map_to_orthologs <- function(gene_ids, orth_map) {
  orth_map %>%
    filter(gene_id %in% gene_ids) %>%
    pull(Preferred_name) %>%
    unique()
}

# Helper: count how many genes in a set have at least one ortholog
count_mapped_genes <- function(gene_ids, orth_map) {
  mapped <- orth_map %>%
    filter(gene_id %in% gene_ids) %>%
    distinct(gene_id)
  nrow(mapped)
}

# Apply mapping to all tiers using ALL orthologs
tier_orthologs <- lapply(tiers, map_to_orthologs, orth_map = orth_map_all)

# Orthology summary (all orthologs retained)
ortholog_summary <- tibble(
  tier = names(tiers),
  total_genes = sapply(tiers, length),
  genes_with_orthologs = sapply(tiers, count_mapped_genes, orth_map = orth_map_all),
  distinct_orthologs = sapply(tier_orthologs, length)
) %>%
  mutate(
    coverage_pct = round((genes_with_orthologs / total_genes) * 100, 1)
  )

ortholog_summary
```

# Run GO enrichment for all tiers
```{r}
# Helper function: run g:Profiler enrichment 
run_enrichment <- function(orthologs, tier_name, min_genes = 15) {
  if (length(orthologs) >= min_genes) {
    gp <- gost(
      query = orthologs,
      organism = "hsapiens",
      correction_method = "fdr",
      sources = c("GO:BP", "GO:MF", "GO:CC")
    )
    if (!is.null(gp$result)) {
      gp$result$tier <- tier_name
      return(list(name = tier_name, gp = gp))
    } else {
      message(paste("No enrichment results for", tier_name))
      return(NULL)
    }
  } else {
    message(paste("Skipping", tier_name, "- too few orthologs (", length(orthologs), ")", sep = " "))
    return(NULL)
  }
}

# Apply enrichment to all tiers (now includes merged Tier3 sets)
enrichment_outputs <- lapply(names(tier_orthologs), function(tn) {
  run_enrichment(tier_orthologs[[tn]], tn)
})
names(enrichment_outputs) <- names(tier_orthologs)
```

# Plot enrichment results for all tiers
```{r}
for (tn in names(enrichment_outputs)) {
  obj <- enrichment_outputs[[tn]]
  if (!is.null(obj)) {
    suppressWarnings({
      plot <- gostplot(obj$gp, capped = TRUE, interactive = FALSE) +
        ggtitle(paste("GO Enrichment –", tn))
      print(plot)
    })
  }
}
```


# Dotplots for all tiers 
```{r}
# Define dotplot function
go_dotplot <- function(gp_result, domain = "GO:BP", n = 10, title = NULL) {
  df <- gp_result %>%
    filter(source == domain) %>%
    arrange(p_value) %>%        # sort by p-value
    head(n) %>%                 # take top n terms only
    mutate(term_name = str_wrap(term_name, width = 40)) %>%
    mutate(term_name = factor(term_name, levels = rev(term_name)))

  if (nrow(df) == 0) return(NULL)  # skip if no terms for this domain

  ggplot(df, aes(x = -log10(p_value), y = term_name, size = intersection_size)) +
    geom_point(color = "#2C7BB6") +
    labs(
      x = "-log10(FDR)",
      y = "",
      title = title
    ) +
    theme_minimal(base_size = 12) +
    theme(axis.text.y = element_text(size = 9))  # smaller labels for clarity
}

# Dotplots for all tiers (BP, MF, CC)
for (tn in names(enrichment_outputs)) {
  obj <- enrichment_outputs[[tn]]
  if (!is.null(obj)) {
    df <- obj$gp$result
    if (nrow(df) > 0) {
      if (any(df$source == "GO:BP")) {
        print(go_dotplot(df, domain = "GO:BP", n = 10,
                         title = paste(tn, "— Biological Process")))
      }
      if (any(df$source == "GO:MF")) {
        print(go_dotplot(df, domain = "GO:MF", n = 10,
                         title = paste(tn, "— Molecular Function")))
      }
      if (any(df$source == "GO:CC")) {
        print(go_dotplot(df, domain = "GO:CC", n = 10,
                         title = paste(tn, "— Cellular Component")))
      }
    }
  }
}
```

## KEGG Enrichment
```{r}

# Helper function: map gene symbols to Entrez IDs
map_to_entrez <- function(symbols) {
  bitr(symbols,
       fromType = "SYMBOL",
       toType   = "ENTREZID",
       OrgDb    = org.Hs.eg.db)$ENTREZID
}

# Perform KEGG enrichment for all tiers 
kegg_outputs <- list()

for (tn in names(tier_orthologs)) {
  symbols <- tier_orthologs[[tn]]
  if (!is.null(symbols) && length(symbols) > 0) {
    entrez_ids <- map_to_entrez(symbols)
    if (length(entrez_ids) > 0) {
      kegg_outputs[[tn]] <- enrichKEGG(
        gene          = entrez_ids,
        organism      = "hsa",
        pAdjustMethod = "BH"
      )
    } else {
      message(paste("No Entrez IDs mapped for", tn))
    }
  } else {
    message(paste("Skipping", tn, "- no symbols"))
  }
}

# Dotplots for all tiers (KEGG)
for (tn in names(kegg_outputs)) {
  ek <- kegg_outputs[[tn]]
  if (!is.null(ek) && nrow(as.data.frame(ek)) > 0) {
    print(dotplot(ek, showCategory = 15) +
            ggtitle(paste(tn, "— KEGG Pathway Enrichment")))
  }
}
```
## KEGG enrichment summary 
```{r}
kegg_summary <- tibble(
  tier = names(tier_orthologs),
  total_symbols = sapply(tier_orthologs, length),
  mapped_entrez = sapply(tier_orthologs, function(x) {
    ids <- map_to_entrez(x)
    if (!is.null(ids)) length(ids) else 0
  }),
  enriched_pathways = sapply(names(tier_orthologs), function(tn) {
    ek <- kegg_outputs[[tn]]
    if (!is.null(ek)) {
      nrow(as.data.frame(ek))
    } else {
      0
    }
  })
)

kegg_summary
```

```{r}
# Extract KEGG summary (Description, p.adjust, GeneRatio) for all tiers
kegg_tables <- lapply(names(kegg_outputs), function(tn) {
  ek <- kegg_outputs[[tn]]
  if (!is.null(ek) && nrow(as.data.frame(ek)) > 0) {
    df <- as.data.frame(ek)[, c("Description", "p.adjust", "GeneRatio")]
    df$tier <- tn
    return(df)
  } else {
    return(NULL)
  }
})

# Combine into one tibble
kegg_all <- bind_rows(kegg_tables)

kegg_all
```

------
## Otholog summary
```{r}
ortholog_summary
```

## GO enrichment results
```{r}
go_results <- lapply(enrichment_outputs, function(obj) {
  if (!is.null(obj)) {
    df <- as_tibble(obj$gp$result) %>%
      arrange(p_value) %>%
      head(10) %>%
      dplyr::select(term_name, source, p_value, intersection_size) %>%
      mutate(tier = obj$name)
    return(df)
  } else {
    return(NULL)
  }
})

go_all <- bind_rows(go_results)
knitr::kable(go_all, caption = "Top GO enrichment terms across tiers")
```

## KEGG summary
```{r}
kegg_summary %>%
  arrange(desc(enriched_pathways)) %>%
  knitr::kable(caption = "KEGG enrichment summary per tier")
```

## 
```{r}
ggplot(kegg_summary, aes(x = reorder(tier, enriched_pathways), y = enriched_pathways)) +
  geom_col(fill = "#2C7BB6") +
  coord_flip() +
  labs(x = "Tier", y = "Number of enriched KEGG pathways",
       title = "KEGG enrichment across tiers") +
  theme_minimal()
```

```{r}
as.data.frame(kegg_outputs[["Tier1"]])
as.data.frame(kegg_outputs[["Tier2"]])
s.data.frame(kegg_outputs[["Tier2b"]])
as.data.frame(kegg_outputs[["Tier3_mis_all"]])
as.data.frame(kegg_outputs[["Tier3_mis_12C"]])
as.data.frame(kegg_outputs[["Tier3_mis_15C"]])
as.data.frame(kegg_outputs[["Tier3_mis_18C"]])
as.data.frame(kegg_outputs[["Tier1∩Tier3_mis"]])
as.data.frame(kegg_outputs[["Tier2b∩Tier3_mis"]])
as.data.frame(kegg_outputs[[""]])
```

```{r}
# Example: print only key KEGG columns
tiers <- c("Tier1", "Tier2", "Tier2b", "Tier3_mis_all", "Tier3_mis_12C", 
           "Tier3_mis_15C", "Tier3_mis_18C", "Tier3_parent_S_all", 
           "Tier3_parent_S_12C", "Tier3_parent_S_15C", "Tier3_parent_S_18C", 
           "Tier3_parent_W_all", "Tier3_parent_W_12C", "Tier3_parent_W_15C", 
           "Tier3_parent_W_18C", "Tier1∩Tier3_mis", "Tier2b∩Tier3_mis", 
           "Tier1∩Tier3_parent_S", "Tier1∩Tier3_parent_W", 
           "Tier2b∩Tier3_parent_S", "Tier2b∩Tier3_parent_W")

for (t in tiers) {
  cat("\n###", t, "###\n")
  df <- as.data.frame(kegg_outputs[[t]])
  if (nrow(df) > 0) {
    print(df[, c("ID", "Description", "GeneRatio", "p.adjust")])
  } else {
    cat("No enrichment\n")
  }
}
```

--------


## Results

| Tier / Subset              | Total genes | Ortho coverage % | GO enrichment (themes)                                                                 | KEGG enrichment (pathways)                                                                 |
|-----------------------------|-------------|------------------|-----------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------|
| Tier1                       | 585         | 76.4             | ECM, collagen, extracellular matrix, developmental processes, endogenous stimulus        | Proteoglycans in cancer; Integrin signaling; Amino acid biosynthesis; Cysteine & methionine metabolism; Folate cycle; Leukocyte migration |
| Tier2                       | 1135        | 67.0             | Cytoplasm, cytosol, protein binding, intracellular transport, regulation of processes    | IgSF CAM signaling; MAPK signaling pathway                                                  |
| Tier2b                      | 32          | 56.2             | Specialized enzymatic functions (RNA adaptor, vitamin‑K reductase, farnesyl diphosphate synthase, alcohol binding) | No enrichment                                                                               |
| Tier3_mis_12C               | 215         | 70.7             | ATP/nucleotide binding; carbohydrate derivative binding; purine ribonucleotide binding   | No enrichment                                                                               |
| Tier3_mis_15C               | 189         | 65.6             | Cytoplasm; catalytic activity; small molecule metabolism; cytoskeletal functions         | Sulfur metabolism                                                                           |
| Tier3_mis_18C               | 206         | 76.2             | ATP/nucleotide binding; cytoplasm; heterocyclic compound binding                         | No enrichment                                                                               |
| Tier3_mis_all               | 526         | 74.0             | Cytoplasm; nucleotide binding; catalytic activity; exosome terms                         | Motor proteins                                                                              |
| Tier3_parent_S_12C          | 271         | 74.2             | Protein binding; cytoplasm; membrane; axon; vesicle; organelle membrane                  | No enrichment                                                                               |
| Tier3_parent_S_15C          | 163         | 73.6             | Cytoplasm; cytosol; phosphatidylinositol binding; postsynapse; dendritic tree; synapse   | No enrichment                                                                               |
| Tier3_parent_S_18C          | 323         | 76.8             | Cytoplasm; cytosol; cell junction; supramolecular fibers; cytoskeletal fibers; ECM       | Cytoskeleton in muscle cells; Pathogenic *E. coli* infection; Tight junction; Integrin signaling; Motor proteins; Cardiomyopathies; Salmonella infection; Focal adhesion; PI3K‑Akt signaling; Leukocyte migration |
| Tier3_parent_S_all          | 647         | 78.4             | Synapse; vesicle; neuron projection; membrane; cell junction; cytosol                    | Pathogenic *E. coli* infection; Cytoskeleton; Protein digestion & absorption; Salmonella infection; Tight junction; MAPK signaling; Yersinia infection; Actin cytoskeleton; Cadherin signaling; Focal adhesion; Rap1 signaling; Integrin signaling; HIV‑1 infection; Arrhythmogenic cardiomyopathy; Ras signaling |
| Tier3_parent_W_12C          | 321         | 69.5             | Cytoplasm; protein binding; synapse; transport; developmental process                    | Cytoskeleton in muscle cells                                                                |
| Tier3_parent_W_15C          | 1048        | 74.1             | Cytoplasm; cytosol; protein binding; endomembrane system; vesicle; catalytic activity    | Inositol phosphate metabolism; Phosphatidylinositol signaling; Nucleocytoplasmic transport; mRNA surveillance; Insulin signaling; Focal adhesion; Actin cytoskeleton; Cytoskeleton in muscle cells; Oxytocin signaling; Tight junction; Integrin signaling; Circadian entrainment; Shigellosis; Motor proteins; HIV‑1 infection |
| Tier3_parent_W_18C          | 1881        | 71.3             | Cytoplasm; protein binding; cytosol; endomembrane system; vesicle; extracellular exosome | Protein digestion & absorption; Spinocerebellar ataxia; Insulin signaling; Cytoskeleton in muscle cells; Steroid biosynthesis; Autophagy; Neurodegeneration pathways; mRNA surveillance; ER protein processing; ALS; Cell cycle; Tight junction; Sulfur metabolism; Glutathione metabolism; Glucagon signaling; Focal adhesion; Integrin signaling; Actin cytoskeleton; Amino acid biosynthesis; p53 signaling |
| Tier3_parent_W_all          | 2584        | 78.9             | Cytoplasm; cytosol; protein binding; endomembrane system; vesicle; extracellular exosome | Cytoskeleton; Integrin signaling; Spinocerebellar ataxia; Insulin signaling; mRNA surveillance; Arginine & proline metabolism; Focal adhesion; Tight junction; Steroid biosynthesis; Salmonella infection; Motor proteins; Neurodegeneration; Sulfur metabolism; Cell cycle; Shigellosis; Thyroid hormone signaling; Autophagy; Circadian entrainment; Adrenergic signaling |
| Tier1 ∩ Tier3_mis           | 20          | 90.0             | ECM; collagen; cystathionine/methionine activity; tensile strength structural constituent | Selenocompound metabolism; Folate cycle; Cysteine & methionine metabolism; Amino acid biosynthesis |
| Tier2b ∩ Tier3_mis          | 1           | 100.0            | Single driver gene: CD81‑like tetraspanin (immune signaling, cell communication)         | Virion – Hepatitis viruses; Malaria; B cell receptor signaling; Hepatitis C                  |
| Tier1 ∩ Tier3_parent_S      | 30          | 86.7             | ECM structural constituent; collagen trimer; vesicle lumen; ER lumen                     | Integrin signaling; PI3K‑Akt signaling; Focal adhesion; ECM‑receptor interaction; Protein digestion & absorption |
| Tier1 ∩ Tier3_parent_W      | 134         | 88.1             | Cytoplasm; response to endogenous stimulus; ECM; ER lumen; exosome                       | Integrin signaling; Alanine/aspartate/glutamate metabolism; Arginine biosynthesis            |
| Tier2b ∩ Tier3_parent_S     | 1           | 0.0              | Single unannotated gene                                                                 | No enrichment                                                                               |
| Tier2b ∩ Tier3_parent_W     | 26          | 57.7             | RNA adaptor; vitamin‑K reductase; integrin binding; complement binding                   | No enrichment                                                                               |



------------------


Tier 1 - Temperature Response Core Genes: 

  - Total genes: 585
  - Genes with orthologs: 447
  
  GO enrichment:
  - Extracellular matrix (ECM) components
  - Collagen structural constituents
  - Cytoplasm
  - Cytosol
  - Developmental processes
  - Response to endogenous stimulus

  These enrichments highlight their role in maintaining tissue integrity and coordinating temperature‑dependent physiological adjustments.
  
  KEGG enrichment:
  - Proteoglycans in cancer (structural ECM remodeling)
  - Integrin signaling (cell adhesion and communication)
  - Biosynthesis of amino acids
  - Cysteine and methionine metabolism
  - One carbon pool by folate (methylation and metabolic regulation)
  - Leukocyte transendothelial migration (immune system function)

  These pathways emphasize the integration of ECM remodeling, metabolic regulation, and immune function in temperature response.
  
Tier 1 defines the baseline molecular architecture of temperature response. Enrichment of ECM/collagen GO terms together with KEGG pathways for proteoglycans and integrin signaling reflects the interplay between extracellular matrix components and cell–matrix communication. This co‑enrichment shows that both the structural properties of the ECM and the signaling mechanisms that regulate tissue remodeling are engaged. Enrichment of amino acid biosynthesis, cysteine/methionine metabolism, and the folate cycle further shows that structural remodeling is coupled to metabolic re‑programming, enabling tissues to adjust their biochemical environment under different temperature regimes. Finally, the leukocyte transendothelial migration pathway links these processes to immune surveillance, indicating that Tier 1 captures a coordinated adjustment across structural, metabolic, and immune systems.




Tier 2 - South and West Population Divergence

  - Total genes: 1135
  - Genes with orthologs: 761

  GO enrichment:
  - Cytoplasm
  - Cytosol
  - Nucleoplasm
  - Intracellular anatomical structure
  - Intracellular signal transduction
  - Intracellular transport
  - Positive regulation of biological process
  - Positive regulation of cellular process
  - Regulation of biological quality
  
  These enrichments highlight widespread intracellular organization and regulatory processes, pointing to divergence in how cells manage signaling, transport, and quality control.
  
  KEGG enrichment:
  - IgSF CAM signaling (cell adhesion and communication)
  - MAPK signaling pathway (signal transduction and stress response)
  
  These pathways illustrate the intracellular signaling and regulatory processes highlighted by the GO enrichment, showing that divergence between South and West populations involves variation in cell adhesion and MAPK‑mediated signaling.
  
  


Tier 2b - Local Thermal Adaptation

  - Total genes: 32
  - Genes with orthologs: 18
  
  GO enrichment:
  - Protein‑RNA adaptor activity
  - Vitamin‑K‑epoxide reductase (warfarin‑insensitive) activity
  - Protein‑RNA sequence‑specific adaptor activity
  - Dimethylallyltranstransferase activity
  - (2E,6E)‑farnesyl diphosphate synthase activity
  - Alcohol binding
  - Vitamin‑K‑epoxide reductase (warfarin‑sensitive) activity
  
   Locally adapted temperature‑response genes with diverse enzymatic functions, not forming a single KEGG pathway. GO enrichment points to RNA adaptor activities, vitamin K reductases, and enzymes in isoprenoid biosynthesis. These functions suggest that divergence between South and West populations involves differences in RNA regulation, redox balance, and membrane composition, all of which can influence how organisms perform across local thermal regimes. Together, these enrichments indicate that local thermal adaptation is mediated through fine‑scale biochemical adjustments rather than broad structural or signaling processes.



Tier 3 - Hybrid Ineheritance

  Hybrid Missexpression
  
    12°C
    
      - Total genes: 215
      - Genes with orthologs: 152
      
      GO enrichment:
      - ATP/nucleotide binding
      - Carbohydrate derivative binding
      - Purine ribonucleotide binding

      At 12°C, misexpressed genes were enriched for nucleotide binding and carbohydrate derivative binding. These terms reflect            disruption of broad housekeeping functions. KEGG did not return significant pathways, consistent with literature showing that         hybrid misexpression often scatters across diverse metabolic and structural processes rather than concentrating in a single           pathway.
      
      
    15°C
    
      - Total genes: 189
      - Genes with orthologs: 124
      
      GO enrichment:
      - Cytoplasm
      - Catalytic activity
      - Small molecule metabolism
      - Cytoskeletal functions

      KEGG enrichment:
      - Sulfur metabolism

      Misexpression at 15°C highlights maladaptive regulation of core metabolic and cytoskeletal functions, with sulfur metabolism          suggesting condition‑specific vulnerability in redox balance.

    18°C
    
      - Total genes: 206
      - Genes with orthologs: 157
      
      GO enrichment:
      - ATP/nucleotide binding
      - Cytoplasm
      - Heterocyclic compound binding

      Misexpression at 18°C highlights maladaptive regulation of nucleotide metabolism and cytoplasmic functions, reinforcing the           pattern of disrupted housekeeping processes across hybrid treatments.
      
    Merged Temperatures Enrichment
    
      - Total genes: 526
      - Genes with orthologs: 389
      
      GO enrichment:
      - Cytoplasm
      - Nucleotide binding
      - Catalytic activity
      - Exosome terms
      
      These enrichments point to widespread disruption of intracellular metabolism and communication, including exosome‑mediated            processes.
      
      KEGG enrichment:
      - Motor proteins
      
      This pathway suggests disturbed intracellular transport and cytoskeletal dynamics, consistent with broad dysregulation of             structural and transport functions in hybrids.
      
      The combined misexpressed set reflects a pervasive maladaptive breakdown of core cellular machinery, with GO enrichment pointing       to disrupted housekeeping functions such as nucleotide metabolism, catalytic activity, and exosome‑mediated communication, and        KEGG enrichment highlighting motor proteins involved in intracellular transport. Together, these results indicate that hybrid         misexpression scatters across fundamental processes that maintain cellular integrity.

    
  Tier 1 ∩ Tier 3 Misexpression 
  
  - Total genes: 20
  - Genes with orthologs: 18
  
  GO enrichment:
  - ECM structural constituents
  - Collagen trimer
  - Cystathionine/methionine activity
  - Tensile strength structural constituent
  
  These enrichments highlight that hybrid misexpression directly affects extracellular matrix integrity, with collagen and tensile      strength terms indicating disruption of the structural scaffolding central to temperature‑dependent tissue organization, while        cystathionine/methionine activity connects these structural changes to underlying biochemical processes.
  
  KEGG enrichment:
  - Selenocompound metabolism
  - Folate cycle
  - Cysteine & methionine metabolism
  - Amino acid biosynthesis
  
  These pathways emphasize that hybrid misexpression disrupts metabolic regulation, with selenocompound metabolism directly tied to     antioxidant defense, cysteine and methionine metabolism feeding into glutathione synthesis, and the folate cycle linking one‑carbon   metabolism to homocysteine turnover. Together, these enrichments connect structural disruption to impaired redox balance and amino acid biosynthesis, underscoring the biochemical vulnerability of hybrids in maintaining temperature‑dependent tissue organization.

  The overlap between Tier 1 core temperature‑response genes and the misexpressed hybrid set shows that hybrids misregulate genes       central to extracellular matrix organization and metabolic support. Structural ECM and collagen terms indicate that tissue            scaffolding genes are disrupted, while enrichment for selenocompound metabolism, cysteine/methionine turnover, and the folate cycle   highlights compromised antioxidant defense and methylation balance. Together, these results suggest that hybrid misexpression         undermines both the physical scaffolding of tissues and the biochemical pathways that sustain remodeling, pointing to a maladaptive   breakdown of structural and metabolic integration.

  
  Tier 2b ∩ Tier 3 Misexpression  
      
  - Total genes: 1
  - Genes with orthologs: 1
  
  GO:
  - CD81‑like tetraspanin
  
  The single overlap gene is a CD81‑like tetraspanin, a membrane protein with conserved roles in immune receptor signaling and          pathogen entry. In the context of Tier 2b, which captures genes underlying thermal local adaptation, misexpression of this            immune‑related membrane protein in hybrids suggests a maladaptive link between immune responsiveness and temperature dependence.      Because fish are ectotherms, their immune function is tightly coupled to ambient temperature, and misexpression of CD81‑like          tetraspanins may therefore compromise hybrids’ temperature‑dependent immune signaling.


  South Parental Inheritance
  
    12°C
    
      - Total genes: 271
      - Genes with orthologs: 201
      
      GO enrichment:
      - Protein binding
      - Cytoplasm
      - Membrane
      - Axon
      - Vesicle
      - Organelle membrane
      
      At 12 °C, South‑like inheritance involves 271 genes (201 orthologs). Enrichment is dominated by protein binding, cytoplasmic and membrane processes, axonal structures, and vesicle functions. This profile suggests that at cooler temperatures, South‑like regulation emphasizes basic cellular organization and neuronal communication, reflecting a stable physiological baseline

      
    15°C
    
      - Total genes: 163
      - Genes with orthologs: 120
      
      GO enrichment:
      - Cytoplasm
      - Cytosol
      - Phosphatidylinositol binding
      - Postsynapse
      - Dendritic tree
      - Synapse
      
      At 15 °C, the South‑like set narrows to 163 genes (120 orthologs). Enrichment shifts toward synaptic and dendritic functions, postsynapse, and phosphatidylinositol binding. This indicates a focus on signal transduction and neuronal connectivity, pointing to fine‑tuned regulation of communication processes as temperatures rise slightly.

    
    18°C
    
      - Total genes: 323
      - Genes with orthologs: 248
      
      GO enrichment:
      - Cytoplasm
      - Cytosol
      - Cell junction
      - Supramolecular fibers
      - Cytoskeletal fibers
      - Extracellular matrix (ECM)

      KEGG enrichment:
      - Cytoskeleton in muscle cells
      - Pathogenic E. coli infection
      - Tight junction
      - Integrin signaling
      - Motor proteins
      - Cardiomyopathies
      - Salmonella infection
      - Focal adhesion
      - PI3K‑Akt signaling
      - Leukocyte migration
      
      At 18 °C, South‑like inheritance expands to 323 genes (248 orthologs). Enrichment includes cell junctions, supramolecular fibers, cytoskeletal structures, and extracellular matrix (ECM). KEGG pathways highlight tight junctions, integrin signaling, focal adhesion, PI3K‑Akt signaling, and leukocyte migration. Taken together, this indicates that at warmer temperatures, South‑like inheritance emphasizes structural integrity, adhesion, and cytoskeletal organization, maintaining tissue cohesion and cellular architecture rather than mobilizing broad stress pathways.


      
      Merged Temperatures
      
      - Total genes: 647
      - Genes with orthologs: 508
      
      GO enrichment:
      - Synapse
      - Vesicle
      - Neuron projection
      - Membrane
      - Cell junction
      - Cytosol
      
      KEGG enrichment:
      - Pathogenic E. coli infection
      - Cytoskeleton
      - Protein digestion & absorption
      - Salmonella infection
      - Tight junction
      - MAPK signaling
      - Yersinia infection
      - Actin cytoskeleton
      - Cadherin signaling
      - Focal adhesion
      - Rap1 signaling
      - Integrin signaling
      - HIV‑1 infection
      - Arrhythmogenic cardiomyopathy
      - Ras signaling
      
      Across all regimes, South‑like inheritance totals 647 genes (508 orthologs). Enrichment consistently highlights synapse, vesicle, neuron projection, membrane, and cell junction functions, with KEGG pathways including cytoskeleton, focal adhesion, integrin signaling, MAPK signaling, and Ras signaling. This merged profile reinforces that South‑like inheritance is concentrated in structural and adhesion‑related regulation, with repeated emphasis on ECM and cytoskeletal pathways across temperatures.


    Tier 1 ∩ Tier 3_parent_S 
      
      - Total genes: 30
      - Genes with orthologs: 26
  
      GO enrichment:
      - ECM structural constituent
      - Collagen trimer
      - Vesicle lumen
      - ER lumen
      
      KEGG enrichment:
      - Integrin signaling
      - PI3K‑Akt signaling
      - Focal adhesion
      - ECM‑receptor interaction
      - Protein digestion & absorption
      
      Out of the 585 temperature‑responsive genes consistently differentially expressed between temperature regimes, only 30 show           South‑like expression in hybrids. GO enrichment indicates that these genes are concentrated in extracellular matrix and collagen      structural functions, together with vesicle and ER lumen processes. KEGG enrichment highlights adhesion and signaling pathways        (integrin, PI3K‑Akt, focal adhesion, ECM‑receptor interaction) and protein digestion/absorption. This shows that South‑like           expression in hybrids is restricted to a small subset of the temperature‑responsive genes, specifically those involved in ECM         organization and adhesion signaling.
   
  
    Tier 2b ∩ Tier 3_parent_S
    
      - Total genes: 1
      - Genes with orthologs: 0
      
    Western Parental Inheritance
    
      12°C
        - Total genes: 321
        - Genes with orthologs: 223
        
        GO enrichment:
        - Cytoplasm
        - Protein binding
        - Synapse
        - Transport
        - Developmental process
        
        KEGG enrichment:
        - Cytoskeleton in muscle cells
        
        At 12 °C, West‑like inheritance involves 321 genes (223 orthologs). Enrichment includes cytoplasm, protein binding, synapse, transport, and developmental processes, with KEGG highlighting cytoskeleton in muscle cells. This profile reflects baseline cellular organization and developmental regulation, without strong evidence of stress‑specific pathways
  
  
      15°C
      
        - Total genes: 1048
        - Genes with orthologs: 776
    
        GO enrichment:
        - Cytoplasm
        - Cytosol
        - Protein binding
        - Endomembrane system
        - Vesicle
        - Catalytic activity
        
        KEGG enrichment:
        - Inositol phosphate metabolism
        - Phosphatidylinositol signaling
        - Nucleocytoplasmic transport
        - mRNA surveillance
        - Insulin signaling
        - Focal adhesion
        - Actin cytoskeleton
        - Cytoskeleton in muscle cells
        - Oxytocin signaling
        - Tight junction
        - Integrin signaling
        - Circadian entrainment
        - Shigellosis
        - Motor proteins
        - HIV‑1 infection
        
        At 15 °C, the West‑like set expands substantially to 1048 genes (776 orthologs). Enrichment covers cytoplasm, cytosol, protein binding, endomembrane system, vesicle, and catalytic activity, with KEGG pathways including mRNA surveillance, insulin signaling, focal adhesion, actin cytoskeleton, circadian entrainment, and pathogen response pathways. This indicates that at intermediate temperatures, West‑like inheritance mobilizes broader regulatory functions, including homeostasis, signaling, and defense pathways, marking the onset of stress‑associated regulation.
  
      
      18°C
      
        - Total genes: 1881
        - Genes with orthologs: 1341
        
        GO enrichment:
          - Cytoplasm
          - Protein binding
          - Cytosol
          - Endomembrane system
          - Vesicle
          - Extracellular exosome
        
        KEGG enrichment:
        - Protein digestion & absorption
        - Spinocerebellar ataxia
        - Insulin signaling
        - Cytoskeleton in muscle cells
        - Steroid biosynthesis
        - Autophagy
        - Neurodegeneration pathways
        - mRNA surveillance
        - ER protein processing
        - ALS
        - Cell cycle
        - Tight junction
        - Sulfur metabolism
        - Glutathione metabolism
        - Glucagon signaling
        - Focal adhesion
        - Integrin signaling
        - Actin cytoskeleton
        - Amino acid biosynthesis
        - p53 signaling
        
        At 18 °C, West‑like inheritance reaches 1881 genes (1341 orthologs). Enrichment includes cytoplasm, cytosol, endomembrane system, vesicle, and extracellular exosome, with KEGG pathways spanning autophagy, neurodegeneration, ER protein processing, p53 signaling, amino acid metabolism, and immune/redox pathways. These terms are hallmark stress‑responsive functions, covering protein homeostasis, metabolic regulation, and immune signaling. This demonstrates a clear temperature‑dependent increase in stress‑associated regulation at warmer conditions.
  
        
      Merged Temperatures
        
        - Total genes: 2584
        - Genes with orthologs: 2039
        
        GO enrichment:
        - Cytoplasm
        - Cytosol
        - Protein binding
        - Endomembrane system
        - Vesicle
        - Extracellular exosome
  
        KEGG enrichment:
        - Cytoskeleton
        - Integrin signaling
        - Spinocerebellar ataxia
        - Insulin signaling
        - mRNA surveillance
        - Arginine & proline metabolism
        - Focal adhesion
        - Tight junction
        - Steroid biosynthesis
        - Salmonella infection
        - Motor proteins
        - Neurodegeneration
        - Sulfur metabolism
        - Cell cycle
        - Shigellosis
        - Thyroid hormone signaling
        - Autophagy
        - Circadian entrainment
        - Adrenergic signaling
        
        Across all regimes, West‑like inheritance totals 2584 genes (2039 orthologs). Enrichment consistently highlights cytoplasm, cytosol, protein binding, endomembrane system, vesicle, and extracellular exosome, with KEGG pathways including integrin signaling, cytoskeleton, mRNA surveillance, neurodegeneration, autophagy, circadian entrainment, and adrenergic signaling. This merged profile confirms that West‑like inheritance is broad and stress‑responsive, mobilizing diverse pathways across cellular organization, metabolism, and immune regulation, with the strongest expansion at 18 °C.
  
  
    Tier 1 ∩ Tier 3_parent_W
    
      - Total genes: 134
      - Genes with orthologs: 118
  
      GO enrichment:
      - Cytoplasm
      - Response to endogenous stimulus
      - ECM
      - ER lumen
      - Exosome
      
      KEGG enrichment:
      - Integrin signaling
      - Alanine/aspartate/glutamate metabolism
      - Arginine biosynthesis
      
      Out of the 585 temperature‑responsive genes consistently differentially expressed between regimes, 134 show West‑like expression      in hybrids. Of these, 118 have identifiable orthologs. GO enrichment indicates concentration in cytoplasmic and ECM functions,        while KEGG highlights integrin signaling and amino acid metabolism. This shows that hybrids inherit West‑like expression for a        larger fraction of the temperature‑responsive genes compared to South, specifically those involved in cytoplasmic organization,       ECM regulation, and metabolic pathways.
  
    Tier 2b ∩ Tier 3_parent_W
    
      - Total genes: 26
      - Genes with orthologs: 15
      
      GO enrichment:
      - RNA adaptor
      - Vitamin‑K reductase
      - Integrin binding
      - Complement binding
  
      The overlap between Tier 2b and West‑like expression highlights hybrid inheritance of locally adapted genes in the direction of       the Western population. Enrichments include RNA adaptor activity and vitamin‑K reductase, pointing to biochemical adjustments in      RNA regulation and redox balance, alongside integrin binding (structural adhesion) and complement binding (immune‑related).           Because fish immune function is strongly coupled to ambient temperature, the complement binding term indicates that hybrids           inherit Western‑like regulation of immune‑associated genes within the locally adapted set. Together, these enrichments suggest        that hybrids carry Western population inheritance of biochemical fine‑tuning and adhesion/immune functions that contribute to         local thermal adaptation.   

    


    


  
  
  
------------



- West inheritance expands with temperature because the Western population’s regulatory architecture is naturally broader and more responsive across regimes. Hybrids adopt that pattern.

- Misexpression staying stable (~200 genes) shows that the South fluctuation is not just noise — if sample size were the main driver, misexpression would also vary, but it doesn’t.

Additive vs. Condition framing
- Additive model: partitions variance across origin and temperature. It’s conservative, shrinks small contrasts, and reports only genes that are significantly different in a global average sense. That’s why South vs. Hybrid looks small (11 up, 19 down).
- Condition model: explicitly compares hybrids to each parental origin at each temperature. This is the model you used for the inheritance analysis, and it’s much better suited to asking “Do hybrids look South‑like or West‑like at this regime?”

- Without condition contrasts, the additive results alone can give the misleading impression that hybrids are “closer to South” (because of the small DE set). But the condition contrasts show that hybrids are not simply intermediate — they adopt West‑like regulation in a temperature‑dependent way.

Demographic reconstructions suggest a colonization route from the British Isles into western Norway and subsequently into the Skagerrak. Western populations, with higher gene flow, represent the ancestral regulatory architecture, whereas southern Skagerrak populations, more strongly differentiated, appear to have evolved a muted transcriptional response to cope with broader thermal variation. Hybrids increasingly adopt the dominant West‑like program under warming, inheriting its stress‑responsive regulatory functions, consistent with the idea that ancestral states exert stronger influence in hybrid expression than derived adaptations. 

 Comparative takeaway
- South inheritance: restricted, structural, ECM/adhesion focus; stable across temperatures, even at 18 °C.
- West inheritance: expands with temperature into stress‑responsive pathways, especially at 18 °C.
- Hybrids: inherit South’s ECM subset but adopt West’s stress‑responsive program more broadly, showing dominance of the ancestral West architecture under warming.




  KEGG enrichment:


- Tier2b alone: Locally adapted temperature‑response genes with diverse enzymatic functions, not forming a single KEGG pathway.
- Tier2b ∩ misexpression:  This intersection yielded a single, well‑annotated gene: a CD81‑like tetraspanin. CD81 is a membrane protein involved in immune signaling and cell communication. Temperature appears to affect the regulation of this immune‑related gene, and in hybrids its expression is disrupted. Because immune pathways in fish are strongly influenced by ambient temperature, this case highlights a biologically plausible mechanism by which thermal adaptation of immune function may be compromised in hybrids.
- Tier2b ∩ parent‑like W: GO enrichment includes immune‑related binding (integrin, complement), reinforcing that Tier2b genes are tied to immune regulation.


Tier2b: Because fish are ectotherms, immune function is strongly temperature‑dependent. Tier2b represents locally adapted temperature‑response genes; the fact that misexpression pulls out immune pathways means hybrids undermine temperature‑dependent immune adaptations between South and West populations.



------------






















