---
title: "Sample Clustering: PCA"
author: "Carlota Myhre de Gouveia"
date: "2026-10-04"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(
  root.dir   = here::here())
```

## Introduction

We applied Principal Component Analysis (PCA) to transcriptomic profiles of *Symphodus melops* (corkwing wrasse) populations from Norway to explore major sources of variation among samples. In addition to population origin (South, West, Hybrid) and temperature treatment (12°C, 15°C, 18°C), we tested correlations between PCA axes and morphometric traits (length, weight, condition factor) to evaluate whether developmental stage or body condition contributed to transcriptomic variation.

## Setup and Libraries
```{r packages, results='hide'}

suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(ggrepel)
  library(readxl)
	library(patchwork)
})

```

## Data Import and Preprocessing
```{r load_data}
# Load gene abundance matrix
abundance_matrix <- read.table(
  "data/DE/abundance_matrix.txt",
  header = TRUE, row.names = 1, check.names = FALSE
)

# Clean column names
colnames(abundance_matrix) <- sub(".*Melops", "", colnames(abundance_matrix))

# Load metadata and recode origin
sample_info <- read_excel("data/sample_conditions_rnaseq.xlsx") %>%
  mutate(
    sample = sub("L", "", RNA_Later_ID),   # matches abundance matrix columns
    origin = recode(Origin,
      "W" = "West",
      "H" = "Hybrid",
      "S" = "South"
    ),
    origin = factor(origin, levels = c("Hybrid", "South", "West")),
    temperature = factor(`Temperature (celsius)`, levels = c(12,15,18)),
    condition = factor(RNAseqgroup,
                       levels = c("W12","W15","W18",
                                  "H12","H15","H18",
                                  "S12","S15","S18")),
    length = `Total length (mm)`,
    weight = `Weight (g)`,
    condition_factor = 100 * (weight / (length/10)^3)  # convert mm→cm
  ) %>%
  select(sample, origin, temperature, condition,
         length, weight, condition_factor) %>%
  arrange(match(sample, colnames(abundance_matrix)))

# Show preview of the abundance matrix
knitr::kable(abundance_matrix[1:10, 1:10],
             caption = "Gene abundance matrix (first 10 genes × 10 samples)")

# Show metadata
knitr::kable(sample_info, caption = "Sample metadata with morphometrics")
```

## Prepare abundance matrix
```{r prep_abundance_matrix}

# Ensure non-negative integer count
abundance_matrix <- round(pmax(abundance_matrix, 0))  

# Keep transcripts with total count ≥ 10 across all samples
keep_genes <- rowSums(abundance_matrix) >= 10
abundance_matrix <- abundance_matrix[keep_genes, ]

dds <- DESeqDataSetFromMatrix(
  countData = abundance_matrix,
  colData = sample_info,
  design = ~ origin + temperature
)
```

## Variance-stabilizing transformation (VST)
```{r normalize}
# Apply variance-stabilizing transformation (VST) to normalize count data
vsd <- vst(dds, blind = TRUE)
```

## PCA using prcomp
```{r run_PCA}
# Perform PCA on transposed VST matrix (samples as rows, transcripts as columns)
pca <- prcomp(t(assay(vsd)))

# Convert PCA results to data frame
pcaData <- as.data.frame(pca$x)

# Add sample names and metadata to PCA data
pcaData$Sample <- colnames(vsd)
pcaData <- cbind(pcaData, as.data.frame(colData(vsd)))

# Calculate percentage of variance explained by each PC
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

```


## Principal Component Analysis

### Variance explained
```{r scree_plot}
var_explained <- summary(pca)$importance[2,] * 100
barplot(var_explained,
        main = "Variance explained by principal components",
        xlab = "Principal Component",
        ylab = "Percentage of variance",
        col = "steelblue")
```

### PCA plot function

```{r PCA_plot_func}
# Color palette

origin_colors <- c(
  "West"   = "#8da0cb",  # lavender
  "Hybrid" = "#66c2a5",  # turquoise
  "South"  = "#fc8d62"   # salmon
)

temperature_colors <- c(
  "12" = "#4C78A8",  # muted blue
  "15" = "#B279A2",  # muted purple
  "18" = "#E45756"   # muted red
)

# PCA plot function
plot_pca <- function(
    pcaData,
    percentVar,
    combo,
    color_by = "temperature",
    shape_by = "origin",
    ellipse = FALSE,
    save = FALSE,
    save_dir = "results/sample_clustering/PCA"
) {

  x_pc <- combo[1]
  y_pc <- combo[2]

  color_palette <- if (color_by == "temperature") {
    temperature_colors
  } else {
    origin_colors
  }

  p <- ggplot(
    pcaData,
    aes_string(
      x = paste0("PC", x_pc),
      y = paste0("PC", y_pc),
      color = color_by,
      shape = shape_by
    )
  ) +
    geom_point(size = 3) +
    xlab(paste0("PC", x_pc, " (", percentVar[x_pc], "% variance)")) +
    ylab(paste0("PC", y_pc, " (", percentVar[y_pc], "% variance)")) +
    scale_color_manual(values = color_palette) +
    scale_shape_manual(values = c(
      "Hybrid" = 16,
      "South"  = 17,
      "West"   = 18,
      "12"     = 16,
      "15"     = 17,
      "18"     = 18
    )) +
    theme_minimal(base_size = 14)

  if (ellipse) {
    p <- p +
      stat_ellipse(
        aes_string(
          group = color_by,
          color = color_by
        ),
        type = "t",
        level = 0.95,
        linewidth = 0.8,
        linetype = 2
      )
  }

  print(p)

  if (save) {
    ggsave(
      filename = file.path(
        save_dir,
        paste0(
          "PCA_",
          color_by,
          "_PC", x_pc,
          "_vs_PC", y_pc,
          ifelse(ellipse, "_ellipse", ""),
          ".png"
        )
      ),
      plot = p,
      width = 8,
      height = 6,
      dpi = 300
    )
  }
}

```

### Generate PCA plots

#### Temperature coloring

```{r PCA_temperature_panel, fig.width=18, fig.height=6}

pc_combos <- list(c(1,2), c(1,3), c(2,3))

temp_plots <- lapply(pc_combos, function(combo) {

  x_pc <- combo[1]
  y_pc <- combo[2]

  ggplot(
    pcaData,
    aes_string(
      x = paste0("PC", x_pc),
      y = paste0("PC", y_pc),
      color = "temperature",
      shape = "origin"
    )
  ) +
    geom_point(size = 3) +
    stat_ellipse(
      aes(group = temperature, color = temperature),
      type = "t",
      level = 0.95,
      linewidth = 0.8,
      linetype = 2
    ) +
    scale_color_manual(values = temperature_colors) +
    scale_shape_manual(values = c(
      "Hybrid" = 16,
      "South"  = 17,
      "West"   = 18
    )) +
    xlab(paste0("PC", x_pc, " (", percentVar[x_pc], "% variance)")) +
    ylab(paste0("PC", y_pc, " (", percentVar[y_pc], "% variance)")) +
    theme_minimal(base_size = 14)
})

temp_panel <- wrap_plots(
  temp_plots,
  ncol = 3,
  guides = "collect"
) &
  theme(
    legend.position = "right"
  )

temp_panel

```
PCA revealed strong separation among temperature treatments. Samples exposed to 12°C formed a distinct cluster, while 15°C and 18°C samples were more closely positioned in multivariate expression space and exhibited partial overlap. This pattern suggests that the largest transcriptomic shift occurred between 12°C and the warmer temperature treatments, whereas the transition from 15°C to 18°C involved comparatively smaller changes in global gene expression.

#### Origin coloring

```{r PCA_origin_panel, fig.width=18, fig.height=6}

origin_plots <- lapply(pc_combos, function(combo) {

  x_pc <- combo[1]
  y_pc <- combo[2]

  ggplot(
    pcaData,
    aes_string(
      x = paste0("PC", x_pc),
      y = paste0("PC", y_pc),
      color = "origin",
      shape = "temperature"
    )
  ) +
    geom_point(size = 3) +
    stat_ellipse(
      aes(group = origin, color = origin),
      type = "t",
      level = 0.95,
      linewidth = 0.8,
      linetype = 2
    ) +
    scale_color_manual(values = origin_colors) +
    scale_shape_manual(values = c(
      "12" = 16,
      "15" = 17,
      "18" = 18
    )) +
    xlab(paste0("PC", x_pc, " (", percentVar[x_pc], "% variance)")) +
    ylab(paste0("PC", y_pc, " (", percentVar[y_pc], "% variance)")) +
    theme_minimal(base_size = 14)
})

origin_panel <- wrap_plots(
  origin_plots,
  ncol = 3,
  guides = "collect"
) &
  theme(
    legend.position = "right"
  )

origin_panel

```
Population origin contributed a weaker component of transcriptomic variation than temperature. In the PC1-PC2 projection, Hybrid samples clustered within the range of variation observed for West individuals, whereas South samples tended to occupy a more distinct region of the ordination space. Despite this pattern, substantial overlap remained among all origin groups.

## Correlation Analyses

### Build dataset for correlation tests and plots
```{r build_corData}

## Build corData with PCA scores + morphometrics
pc_scores <- as.data.frame(pca$x)
pc_scores$Sample <- rownames(pc_scores)

corData <- sample_info %>%
  select(sample, origin, temperature, length, weight, condition_factor) %>%
  left_join(pc_scores, by = c("sample" = "Sample"))

# Preview merged data
knitr::kable(head(corData), caption = "PCA scores with morphometric variables")
```


### Spearman Correlation Analysis: PC1 vs Morphometrics
```{r Spearman_tests_PC1}
# Run Spearman correlations between PC1 scores and each trait
cor_results <- list(
  PC1_vs_Length = cor.test(corData$PC1, corData$length, method = "spearman"),
  PC1_vs_Weight = cor.test(corData$PC1, corData$weight, method = "spearman"),
  PC1_vs_CF     = cor.test(corData$PC1, corData$condition_factor, method = "spearman")
)

# Collect results into a tidy summary table
cor_table <- data.frame(
  Trait = c("Length", "Weight", "Condition Factor"),
  Rho   = c(cor_results$PC1_vs_Length$estimate,
            cor_results$PC1_vs_Weight$estimate,
            cor_results$PC1_vs_CF$estimate),
  Pval  = c(cor_results$PC1_vs_Length$p.value,
            cor_results$PC1_vs_Weight$p.value,
            cor_results$PC1_vs_CF$p.value)
)

# Display results
knitr::kable(cor_table, digits = 3,
             caption = "Spearman correlations between PC1 and morphometric traits")
```
Spearman correlation tests confirmed that PC1 was significantly associated with both length (ρ = 0.42, p = 0.007) and weight (ρ = 0.50, p = 0.001), but not with condition factor (ρ ≈ 0, p = 0.78). 

### Spearman Correlation Analysis: PC2 vs Morphometrics
```{r Spearman_tests_PC2}
## Spearman Correlation Analysis: PC2 vs Morphometrics
cor_results_PC2 <- list(
  PC2_vs_Length = cor.test(corData$PC2, corData$length, method = "spearman"),
  PC2_vs_Weight = cor.test(corData$PC2, corData$weight, method = "spearman"),
  PC2_vs_CF     = cor.test(corData$PC2, corData$condition_factor, method = "spearman")
)

# Collect results into a tidy summary table
cor_table_PC2 <- data.frame(
  Trait = c("Length", "Weight", "Condition Factor"),
  Rho   = c(cor_results_PC2$PC2_vs_Length$estimate,
            cor_results_PC2$PC2_vs_Weight$estimate,
            cor_results_PC2$PC2_vs_CF$estimate),
  Pval  = c(cor_results_PC2$PC2_vs_Length$p.value,
            cor_results_PC2$PC2_vs_Weight$p.value,
            cor_results_PC2$PC2_vs_CF$p.value)
)

# Display results
knitr::kable(cor_table_PC2, digits = 3,
             caption = "Spearman correlations between PC2 and morphometric traits")
```
Spearman correlation tests showed that PC2 was not significantly associated with length (ρ = –0.11, p = 0.51), weight (ρ = –0.18, p = 0.28), or condition factor (ρ = –0.08, p = 0.64).

### Scatterplots of PC1 vs Morphometric Traits

```{r plot_PC1_vs_morphometrics, fig.width=18, fig.height=6}

# Plotting function
plot_pc_trait <- function(data, pc, trait, rho, pval){

  ggplot(
    data,
    aes_string(
      x = pc,
      y = trait,
      color = "temperature",
      shape = "origin"
    )
  ) +
    geom_point(size = 3) +
    geom_smooth(
      method = "lm",
      se = FALSE,
      color = "black",
      aes(group = 1)
    ) +
    scale_color_manual(values = temperature_colors) +
    theme_minimal(base_size = 14) +
    labs(
      x = paste0(pc, " score"),
      y = trait,
      subtitle = paste0(
        "Spearman rho = ",
        round(rho, 2),
        ", p = ",
        signif(pval, 3)
      )
    )
}

# PC1 vs length
p_length <- plot_pc_trait(
  corData,
  "PC1",
  "length",
  cor_results$PC1_vs_Length$estimate,
  cor_results$PC1_vs_Length$p.value
)

# PC1 vs weight
p_weight <- plot_pc_trait(
  corData,
  "PC1",
  "weight",
  cor_results$PC1_vs_Weight$estimate,
  cor_results$PC1_vs_Weight$p.value
)

# PC1 vs condition factor
p_cf <- plot_pc_trait(
  corData,
  "PC1",
  "condition_factor",
  cor_results$PC1_vs_CF$estimate,
  cor_results$PC1_vs_CF$p.value
)

library(patchwork)

supp_fig <- (
  p_length +
    theme(legend.position = "none")
) | (
  p_weight +
    theme(legend.position = "none")
) | (
  p_cf +
    theme(legend.position = "right")
)

supp_fig

```
Scatterplots of PC1 against condition factor, length, and weight were overlaid with regression lines to visualize potential associations. The regression line for condition factor was flat, indicating that relative body condition (K) was not associated with the transcriptomic variation captured by PC1. In contrast, PC1 showed positive relationships with both length and weight, consistent with the significant Spearman correlations and suggesting that body size contributes to the observed variation. Together, these results suggest that developmental stage, as reflected by differences in body size, is more closely associated with the variation captured by PC1 than variation in nutritional status or energy reserves. Because length and weight are highly collinear, we included length as a covariate in downstream differential expression models to account for developmental stage effects. Length was chosen over weight because it provides a more stable and biologically interpretable measure of growth stage, whereas weight can fluctuate with short‑term physiological factors.

## Summary 

PCA identified temperature as the dominant source of transcriptomic variation. The first two principal components explained 17.3% and 7.0% of the variance, respectively. Samples exposed to 12°C formed a distinct cluster, whereas the 15°C and 18°C treatments occupied more similar regions of multivariate expression space and showed partial overlap. Population origin contributed a weaker component of variation. In the PC1-PC2 projection, Hybrid samples clustered within the range of variation observed for West individuals, whereas South samples tended to occupy a more distinct region of the ordination space. Despite this pattern, substantial overlap remained among all origin groups.

PC1 was significantly associated with both length and weight, suggesting that variation related to body size contributes to transcriptomic variation among samples. In contrast, condition factor showed no association with the principal components. Because length and weight were highly correlated, length was retained as the preferred proxy for developmental stage and used as a covariate in downstream differential expression analyses.

Together, these results indicate that temperature is the primary factor structuring transcriptomic profiles in *Symphodus melops*. Population origin contributed weaker transcriptomic structure, while variation in body size was associated with PC1, supporting the inclusion of length as a covariate in downstream differential expression analyses.

## Session Info
```{r session_info} 
sessionInfo()
```

```{r save_figures, include=FALSE}

# -----------------------------
# PCA panels
# -----------------------------

ggsave(
  "results/sample_clustering/PCA/PCA_origin_panel.png",
  origin_panel,
  width = 18,
  height = 6,
  dpi = 300
)

ggsave(
  "results/sample_clustering/PCA/PCA_temperature_panel.png",
  temp_panel,
  width = 18,
  height = 6,
  dpi = 300
)

# -----------------------------
# Individual PCA plots
# -----------------------------

for (combo in pc_combos) {

  plot_pca(
    pcaData,
    percentVar,
    combo,
    color_by = "origin",
    shape_by = "temperature",
    ellipse = TRUE,
    save = TRUE
  )

  plot_pca(
    pcaData,
    percentVar,
    combo,
    color_by = "temperature",
    shape_by = "origin",
    ellipse = TRUE,
    save = TRUE
  )

}

# -----------------------------
# Morphometric panel
# -----------------------------

ggsave(
  "results/sample_clustering/PCA/Supplementary_Figure_S4_PC1_morphometrics.png",
  supp_fig,
  width = 18,
  height = 6,
  dpi = 300
)

```
