---
title: "Model Comparison and Gene Extraction for Tier 2b"
author: "Carlota Myhre de Gouveia"
date: "2026-09-04"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(
  root.dir   = here::here())
```

# Model comparison via likelihood ratio tests (LRTs)

## Setup and load required libraries
```{r packages, results='hide'}
# Load libraries 
suppressPackageStartupMessages({
  library(DESeq2)
  library(tidyverse)
  library(dplyr)
  library(UpSetR)
  library(ggplot2)
  library(variancePartition)
  library(BiocParallel)
  library(limma)
  library(readxl)
	library(knitr)
	library(pheatmap)
})
```


## Data Import and Preprocessing
```{r load_data}
# Load transcript abundance matrix
abundance_matrix <- read.table("data/DE/abundance_matrix.txt",
                               header = TRUE, row.names = 1, check.names = FALSE)

colnames(abundance_matrix) <- sub(".*Melops", "", colnames(abundance_matrix))

sample_info <- read_excel("data/sample_conditions_rnaseq.xlsx") %>%
  mutate(
    sample = sub("L", "", RNA_Later_ID),   # matches abundance matrix columns
    origin = factor(Origin, levels = c("W","H","S")),
    temperature = factor(`Temperature (celsius)`, levels = c(12,15,18)),
    condition = factor(RNAseqgroup,
                       levels = c("W12","W15","W18",
                                  "H12","H15","H18",
                                  "S12","S15","S18")),
    length = scale(`Total length (mm)`)
  ) %>%
  select(sample, origin, temperature, condition, length)

# Reorder sample_info rows to match abundance_matrix columns
sample_info <- sample_info %>%
  arrange(match(sample, colnames(abundance_matrix)))


# Show preview of the abundance matrix
knitr::kable(abundance_matrix[1:10, 1:10], caption = "Gene abundance matrix (first 10 genes × 10 samples)")

# Show metadata
knitr::kable(sample_info, caption = "Sample metadata")

```

## Prepare abundance matrix
```{r prep_abundance_matrix}

# Ensure non-negative integer count
abundance_matrix <- round(pmax(abundance_matrix, 0))  

# Keep transcripts with total count ≥ 10 across all samples
keep_genes <- rowSums(abundance_matrix) >= 10
abundance_matrix <- abundance_matrix[keep_genes, ]
```


## Create DESeq2 datasets for each model
```{r define_model_datasets}

# Condition model
dds_condition <- DESeqDataSetFromMatrix(
  countData = abundance_matrix,
  colData   = sample_info,
  design    = ~ condition + length
)

# Additive model
dds_additive <- DESeqDataSetFromMatrix(
  countData = abundance_matrix,
  colData   = sample_info,
  design    = ~ temperature + origin + length
)

# Interaction model
dds_interaction <- DESeqDataSetFromMatrix(
  countData = abundance_matrix,
  colData   = sample_info,
  design    = ~ temperature * origin + length
)
```

## Compare simple additive vs additive with lenght
```{r ltr_add_vs_add_len}
## Run DESeq2 for length-adjusted additive model
dds_additive <- DESeq(dds_additive)

## Compare additive with length vs simple additive (testing length effect)
dds_lrt_length_vs_add <- DESeq(
  dds_additive,
  test    = "LRT",
  reduced = ~ temperature + origin
)

res_length_vs_add <- results(dds_lrt_length_vs_add, alpha = 0.05)
summary(res_length_vs_add)
```

The likelihood ratio test comparing the additive model with and without length identified 6,900 genes where body size explained significant variation. Because body size serves as a proxy for developmental stage, we included length as a covariate in all subsequent models. This ensures that observed effects in condition, additive, and interaction analyses reflect biological factors of interest rather than developmental differences


## Run DESeq2 for full model
```{r run_DESeq2}
dds_interaction <- DESeq(dds_interaction)
```

## Compare interaction vs additive model
```{r ltr_interaction_vs_additive}
dds_lrt_inter_vs_add <- DESeq(
  dds_interaction,
  test    = "LRT",
  reduced = ~ temperature + origin + length
)
res_inter_vs_add <- results(dds_lrt_inter_vs_add, alpha = 0.05)
summary(res_inter_vs_add)

```
## Extract genes with significant interaction effects for Tier 2b. These are candidates for origin-specific temperature responses.
```{r extract_interaction_genes_for_tier2b}

# Filter transcripts with adjusted p-value < 0.05 from the LRT
interaction_transcripts <- rownames(res_inter_vs_add)[which(res_inter_vs_add$padj < 0.05)]

# Save to file for use in additive model DE script (Tier 2 overlap)
write.csv(
  data.frame(transcript_id = interaction_transcripts),
  "results/DE/interaction_effects/significant_interaction_genes.csv",
  row.names = FALSE
)

# Print number of transcripts
cat("Number of genes with significant interaction effects:", length(interaction_transcripts), "\n")
```

## Compare condition vs null model
```{r ltr_condition_vs_null}
dds_condition <- DESeq(dds_condition)
dds_lrt_cond_vs_null <- DESeq(dds_condition, test = "LRT", reduced = ~ 1)
res_cond_vs_null <- results(dds_lrt_cond_vs_null, alpha = 0.05)
summary(res_cond_vs_null)
```

## Compare additive vs null model
```{r ltr_additive_vs_null}
dds_additive <- DESeq(dds_additive)
dds_lrt_add_vs_null <- DESeq(dds_additive, test = "LRT", reduced = ~ 1)
res_add_vs_null <- results(dds_lrt_add_vs_null, alpha = 0.05)
summary(res_add_vs_null)
```

## Compare interaction vs null model
```{r ltr_interaction_vs_null}
dds_lrt_inter_vs_null <- DESeq(dds_interaction, test = "LRT", reduced = ~ 1)
res_inter_vs_null <- results(dds_lrt_inter_vs_null, alpha = 0.05)
summary(res_inter_vs_null)
```

## Visualize results
```{r model_overlap}

# Create list of significant genes from each comparison
listInput <- list(
  Inter_vs_Add = rownames(res_inter_vs_add)[which(res_inter_vs_add$padj < 0.05)],
  Cond_vs_Null = rownames(res_cond_vs_null)[which(res_cond_vs_null$padj < 0.05)],
  Add_vs_Null  = rownames(res_add_vs_null)[which(res_add_vs_null$padj < 0.05)],
  Inter_vs_Null = rownames(res_inter_vs_null)[which(res_inter_vs_null$padj < 0.05)]
)

# UpSet plot
upset(fromList(listInput), order.by = "freq", main.bar.color = "steelblue")

# Intersection table
all_genes <- unique(unlist(listInput))
presence_df <- tibble(
  gene = all_genes,
  Inter_vs_Add = all_genes %in% listInput$Inter_vs_Add,
  Cond_vs_Null = all_genes %in% listInput$Cond_vs_Null,
  Add_vs_Null  = all_genes %in% listInput$Add_vs_Null,
  Inter_vs_Null = all_genes %in% listInput$Inter_vs_Null
)

summary_table <- presence_df %>%
  group_by(Inter_vs_Add, Cond_vs_Null, Add_vs_Null, Inter_vs_Null) %>%
  summarise(GeneCount = n(), .groups = "drop") %>%
  arrange(desc(GeneCount))

print(summary_table)

# Totals for each model
totals_table <- tibble(
  Model = c("Interaction vs Additive", "Condition vs Null", "Additive vs Null", "Interaction vs Null"),
  SignificantGenes = c(
    length(listInput$Inter_vs_Add),
    length(listInput$Cond_vs_Null),
    length(listInput$Add_vs_Null),
    length(listInput$Inter_vs_Null)
  )
)

print(totals_table)

```

The overlap analysis shows that the additive vs null comparison yielded the largest set of significant genes (26,402), including 3,810 genes unique to the additive model. In contrast, the condition vs null and interaction vs null comparisons each identified 24,736 genes, with nearly identical sets because both models flexibly partition variance across origin–temperature combinations. Within this overlap, 2,053 genes were significant in condition and interaction models but not in the additive framework, reflecting transcripts that are only detected when variance is modeled at the full condition level. The interaction vs additive comparison identified only 977 genes (0.6%), highlighting a small subset where the interaction term improves fit. Together, these results indicate that the additive model not only captures the broadest biological signal but also contributes a larger number of unique genes compared to the condition and interaction frameworks, while the condition/interaction models identify a modest subset of genes tied to fine‑grained group effects.


# Variance partition

## Setup and shared data
```{r setup_variance_partition}

# Register parallel backend 
register(SnowParam(4))  

# Use variance-stabilized counts 
vsd <- vst(dds_interaction, blind = FALSE)
expr_matrix <- assay(vsd)
metadata <- as.data.frame(colData(vsd))

# Sanity check: ensure sample names align
stopifnot(setequal(colnames(expr_matrix), rownames(metadata)))

# Filter out low-variance genes
expr_filtered <- expr_matrix[rowVars(expr_matrix) > 0.1, ]

# Ensure factors
metadata$origin <- factor(metadata$origin)
metadata$temperature <- factor(metadata$temperature)

```

## Check number of filtered genes
```{r check_filtering}
# Number of genes before filtering
n_before <- nrow(expr_matrix)

# Number of genes after filtering
n_after <- nrow(expr_filtered)

# Number of genes removed
n_removed <- n_before - n_after

cat("Genes before filtering:", n_before, "\n")
cat("Genes after filtering:", n_after, "\n")
cat("Genes removed:", n_removed, "\n")

```

## Variance partition and plotting
```{r run_and_plot_variance_partition}

# Define formulas
form_condition   <- ~ (1|condition) + length
form_additive    <- ~ (1|origin) + (1|temperature) + length
form_interaction <- ~ (1|origin) + (1|temperature) + (1|origin:temperature) + length

# Fit models
fit_condition   <- fitVarPartModel(expr_filtered, form_condition, metadata)
fit_additive    <- fitVarPartModel(expr_filtered, form_additive, metadata)
fit_interaction <- fitVarPartModel(expr_filtered, form_interaction, metadata)

# Extract variance explained
varPart_condition   <- extractVarPart(fit_condition)
varPart_additive    <- extractVarPart(fit_additive)
varPart_interaction <- extractVarPart(fit_interaction)


# Plot 1: Condition model
plotVarPart(varPart_condition) +
  theme_minimal(base_size = 14) +
  ggtitle("Variance Partitioning - Condition Model")

# Plot 2: Additive model
plotVarPart(varPart_additive) +
  theme_minimal(base_size = 14) +
  ggtitle("Variance Partitioning - Additive Model")

# Plot 3: Interaction model
plotVarPart(varPart_interaction) +
  theme_minimal(base_size = 14) +
  ggtitle("Variance Partitioning - Interaction Model")

# Convert to data frames and add model labels
vp_add <- as.data.frame(varPart_additive) %>% mutate(Model = "Additive")
vp_int <- as.data.frame(varPart_interaction) %>% mutate(Model = "Interaction")
vp_cond <- as.data.frame(varPart_condition) %>% mutate(Model = "Condition")

# Combine and reshape
vp_combined <- bind_rows(vp_add, vp_int, vp_cond) %>%
  pivot_longer(cols = -Model, names_to = "Factor", values_to = "Variance") %>%
  filter(is.finite(Variance))

# Plot 4a: Boxplot comparison
ggplot(vp_combined, aes(x = Factor, y = Variance, fill = Model)) +
  geom_boxplot(outlier.size = 0.5) +
  theme_minimal(base_size = 14) +
  labs(title = "Variance Explained by Factor: All Models")

# Plot 4b: Violin plot comparison 
ggplot(vp_combined, aes(x = Factor, y = Variance, fill = Model)) +
  geom_violin(trim = FALSE, scale = "width") +
  theme_minimal(base_size = 14) +
  labs(title = "Variance Distribution by Factor: All Models")


```
Boxplots and violin plots show that temperature consistently explains the largest share of variance, while origin contributes little.

## Lenght correlations
```{r len_correlations}
cor(as.numeric(metadata$length), as.numeric(metadata$origin))
cor(as.numeric(metadata$length), as.numeric(metadata$temperature))
```
The correlation between length and origin was weak (r ≈ 0.14), suggesting that body length is largely independent of population of origin. In contrast, length showed a modest negative correlation with temperature (r ≈ –0.26), indicating that individuals from higher temperature treatments tended to be slightly shorter. 
Together, these results imply that length is not confounded with origin, but partially overlaps with temperature effects. Since variance partitioning showed that length explains more variance in gene expression than origin, it represents an important axis of biological variation. Including length as a fixed effect in downstream models helps control for growth differences while still allowing clear interpretation of origin and temperature effects.


## Average variance explained by each factor across models
```{r variance table}
# Compute averages separately
avg_condition   <- colMeans(varPart_condition)
avg_additive    <- colMeans(varPart_additive)
avg_interaction <- colMeans(varPart_interaction)

# Convert each to tidy format
df_condition   <- data.frame(Factor = names(avg_condition),   Condition   = avg_condition)
df_additive    <- data.frame(Factor = names(avg_additive),    Additive    = avg_additive)
df_interaction <- data.frame(Factor = names(avg_interaction), Interaction = avg_interaction)

# Join them by Factor
avg_var <- full_join(df_condition, df_additive, by="Factor") %>%
           full_join(df_interaction, by="Factor")

# Display table
kable(avg_var, digits = 3,
      caption = "Average variance explained by each factor across models")
```
Model choice justification: Variance partitioning showed that temperature (~22–24%) and length (~7–8%) were the dominant sources of variance, while origin contributed little (~3%). The origin × temperature interaction explained ~5–6% variance overall, and likelihood ratio tests identified only ~1% of genes with significant interaction effects. Because the interaction term contributes modestly at the global level, we adopted the additive model for variance partitioning to maintain parsimony and interpretability. Interaction‑responsive genes identified by LTR were analyzed separately to highlight population‑specific responses to temperature.


# Sample Clustering

## MDS Plot
```{r mds_plot, fig.width=8, fig.height=6}
# Compute Euclidean distance matrix
dist_matrix <- dist(t(assay(vsd)))  # transpose so samples are rows

# Perform classical MDS
mds_coords <- cmdscale(dist_matrix, k = 2)
mds_df <- as.data.frame(mds_coords)
colnames(mds_df) <- c("MDS1", "MDS2")
mds_df$Sample <- colnames(vsd)
mds_df <- cbind(mds_df, as.data.frame(colData(vsd)))

# Plot MDS with color by Temperature and shape by Origin
ggplot(mds_df, aes(x = MDS1, y = MDS2, 
                   color = factor(temperature), shape = factor(origin))) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = temperature), type = "t", level = 0.95, linetype = 2) +
  labs(title = "MDS Plot of Trinity-derived Counts",
       color = "Temperature", shape = "Origin") +
  theme_minimal(base_size = 14)

# Save to file
ggsave(
  filename = "results/sample_clustering/MDS/MDS_plot.png",
  width = 8, height = 6, dpi = 300
)

```
Samples cluster primarily by temperature, with weaker separation by origin, supporting the additive model.

## Sample-to-sample distance heatmap
```{r sample_distance_heatmap}
## Sample-to-sample distance heatmap
sampleDists <- dist(t(expr_matrix))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(expr_matrix)
colnames(sampleDistMatrix) <- colnames(expr_matrix)

# Use only categorical annotations
ann_col <- metadata %>%
  select(condition, origin, temperature) %>%
  as.data.frame()
rownames(ann_col) <- metadata$sample

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         annotation_col = ann_col,
         main = "Sample-to-sample distances")
```



The sample‑to‑sample distance heatmap revealed clear grouping of individuals at 12°C across origins, while clustering among the 15°C and 18°C conditions was less distinct. This indicates that temperature is the dominant factor structuring transcriptomic profiles, whereas population origin contributed little, and the nine experimental groups did not form fully distinct clusters.

# Summary

This script compares condition, additive, and interaction models using variance partitioning and LRTs to evaluate differential expression patterns in *Symphodus melops*. 
The likelihood ratio test comparing the additive model with and without length identified 6,900 genes where body size explained significant variation. Because body size serves as a proxy for developmental stage, we included length as a covariate in all subsequent models. This ensures that observed effects reflect biological factors of interest rather than developmental differences.
The condition model explained the highest proportion of variance, reflecting its flexibility in treating each origin-temperature combination as a distinct group. However, this structure does not share information across conditions and is prone to overfitting, especially with limited replication.
In contrast, the additive model explained substantial variance, particularly for temperature, and identified the largest number of significantly differentially expressed genes. This is supported by the MDS plot, which shows that samples primarily cluster by temperature, with less separation by origin — reinforcing the additive model’s ability to capture the dominant biological signal.
LRT comparing the interaction model to the additive model revealed that only 977 genes (~0.6%) showed significantly improved fit when including the interaction term. These genes were considered candidates for origin-dependent temperature responses and were extracted for downstream analysis. Based on these results, the additive model was selected for global differential expression analysis.


# Session Info
```{r session_info} 
sessionInfo()
```







