---
title: "Tier 3 Differential Expression Analysis (Hybrid Misregulation)"
author: "Carlota Myhre de Gouveia"
date: "2026-02-15"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


# Setup and Package Loading
```{r packages}
# Load required packages
suppressPackageStartupMessages({
  library(DESeq2)      # Differential expression analysis
  library(ashr)        # Shrinkage for log2 fold changes
  library(tidyverse)   # Data manipulation
  library(here)        # Project-root relative paths
  library(dplyr)
  library(rlang)
})
```



# Data Import and Inspection
```{r load_and_inspect_data}
# ================================
# Data Import and Cleaning
# ================================

# -----------------------------
# 1. Load raw files exactly as-is
# -----------------------------
counts_raw <- read.delim(
  counts_file,
  header = TRUE,
  sep = "\t",
  check.names = FALSE
)

tpm_raw <- read.delim(
  tpm_file,
  header = TRUE,
  sep = "\t",
  check.names = FALSE
)

# -----------------------------
# 2. Fix the missing first column name
# -----------------------------
colnames(counts_raw)[1] <- "isoform"
colnames(tpm_raw)[1]    <- "isoform"

# -----------------------------
# 3. Remove the metadata column
# -----------------------------
counts_raw <- counts_raw[, colnames(counts_raw) != "abundance_estimation"]
tpm_raw    <- tpm_raw[, colnames(tpm_raw)    != "abundance_estimation"]

# -----------------------------
# 4. Set isoform IDs as rownames
# -----------------------------
rownames(counts_raw) <- counts_raw$isoform
rownames(tpm_raw)    <- tpm_raw$isoform

counts_raw <- counts_raw[, -1]   # remove isoform column
tpm_raw    <- tpm_raw[, -1]

# -----------------------------
# 5. Convert to numeric safely
# -----------------------------
abundance_matrix <- as.data.frame(lapply(counts_raw, as.numeric))
tpm_matrix       <- as.data.frame(lapply(tpm_raw,    as.numeric))

rownames(abundance_matrix) <- rownames(counts_raw)
rownames(tpm_matrix)       <- rownames(tpm_raw)

# -----------------------------
# 6. Remove rows with any NA
# -----------------------------
keep_complete <- complete.cases(abundance_matrix) & complete.cases(tpm_matrix)

abundance_matrix <- abundance_matrix[keep_complete, ]
tpm_matrix       <- tpm_matrix[keep_complete, ]

cat("Rows removed due to NA:", sum(!keep_complete), "\n")
cat("Final number of isoforms:", nrow(abundance_matrix), "\n")

# ================================
# Inspect Cleaned Count Matrix
# ================================

cat("Dimensions:", dim(abundance_matrix), "\n")

# Check column types
cat("\nColumn classes:\n")
print(sapply(abundance_matrix, class))

# Summary of first few columns
cat("\nSummary of first 5 columns:\n")
print(summary(abundance_matrix[, 1:5]))

# Check for any negative values (should be none)
cat("\nAny negative values?", any(abundance_matrix < 0), "\n")

# Check for non-integers (DESeq2 will round anyway, but good to know)
cat("\nAny non-integers?", any(abundance_matrix %% 1 != 0), "\n")

# Look at first few rows
cat("\nHead of abundance_matrix:\n")
print(head(abundance_matrix[, 1:6]))

# Check row sums distribution
cat("\nRow sums summary:\n")
print(summary(rowSums(abundance_matrix)))

# Check if any sample has suspiciously low total counts
cat("\nLibrary sizes (column sums):\n")
print(colSums(abundance_matrix))

```
```{r}
sample_info
```


## Data Cleaning and Filtering
```{r filtering}

# Align column names with sample metadata
colnames(abundance_matrix) <- sample_info$sample
colnames(tpm_matrix)       <- sample_info$sample

# Convert to matrices
abund_num <- as.matrix(abundance_matrix)
tpm_num   <- as.matrix(tpm_matrix)

# Filtering rules
keep_counts <- rowSums(abund_num) >= 10
keep_tpm    <- rowSums(tpm_num >= 1) >= 2

# Combined filter
keep <- keep_counts & keep_tpm

# Apply filter
abundance_matrix <- abundance_matrix[keep, ]
tpm_matrix       <- tpm_matrix[keep, ]

# Report
initial_n <- nrow(abund_num)
final_n   <- nrow(abundance_matrix)

cat("Initial isoforms:", initial_n, "\n")
cat("After filtering:", final_n, "\n")
cat("Retained:", round(final_n / initial_n * 100, 2), "%\n")
```
```{r}
# Inspect the filtered count matrix
cat("Class of abundance_matrix:\n")
print(class(abundance_matrix))

cat("\nStorage mode:\n")
print(storage.mode(abundance_matrix))

cat("\nAny NA values?\n")
print(any(is.na(abundance_matrix)))

cat("\nAny non-integers?\n")
print(any(abundance_matrix %% 1 != 0))

cat("\nSummary of values:\n")
print(summary(as.vector(as.matrix(abundance_matrix))))
```

# DESeq2 Model Setup
```{r deseq_setup}
# Convert to integer matrix (RSEM expected counts → DESeq2)
abund_num <- round(as.matrix(abundance_matrix))
storage.mode(abund_num) <- "integer"

# Build DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = abund_num,
  colData   = sample_info,
  design    = ~ condition
)

# Run DESeq2
dds <- DESeq(dds)

cat("DESeq2 model fitted for Tier 3 analysis\n")
```
# Differential Expression for Tier 3 Contrasts
```{r DE_contrasts}

output_dir <- here("results", "DE", "tier3_DE")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

temps <- c(12, 15, 18)

for (t in temps) {
  
  cond_H <- paste0("H", t)
  cond_S <- paste0("S", t)
  cond_W <- paste0("W", t)
  
  # -------------------------
  # H vs S
  # -------------------------
  message("\nRunning contrast: ", cond_H, " vs ", cond_S)
  
  res_HS <- results(dds, contrast = c("condition", cond_H, cond_S))
  res_HS_shrunk <- lfcShrink(dds, contrast = c("condition", cond_H, cond_S),
                             res = res_HS, type = "ashr")
  
  iso <- rownames(res_HS_shrunk)
  df_HS <- as.data.frame(res_HS_shrunk)
  df_HS$isoform <- iso
  df_HS <- df_HS[, c("isoform", setdiff(colnames(df_HS), "isoform"))]
  
  write.table(df_HS,
              file.path(output_dir, paste0(cond_H, "_vs_", cond_S, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  
  # -------------------------
  # H vs W
  # -------------------------
  message("Running contrast: ", cond_H, " vs ", cond_W)
  
  res_HW <- results(dds, contrast = c("condition", cond_H, cond_W))
  res_HW_shrunk <- lfcShrink(dds, contrast = c("condition", cond_H, cond_W),
                             res = res_HW, type = "ashr")
  
  iso <- rownames(res_HW_shrunk)
  df_HW <- as.data.frame(res_HW_shrunk)
  df_HW$isoform <- iso
  df_HW <- df_HW[, c("isoform", setdiff(colnames(df_HW), "isoform"))]
  
  write.table(df_HW,
              file.path(output_dir, paste0(cond_H, "_vs_", cond_W, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  
  # -------------------------
  # S vs W (parental contrast)
  # -------------------------
  message("Running parental contrast: ", cond_S, " vs ", cond_W)
  
  res_SW <- results(dds, contrast = c("condition", cond_S, cond_W))
  res_SW_shrunk <- lfcShrink(dds, contrast = c("condition", cond_S, cond_W),
                             res = res_SW, type = "ashr")
  
  iso <- rownames(res_SW_shrunk)
  df_SW <- as.data.frame(res_SW_shrunk)
  df_SW$isoform <- iso
  df_SW <- df_SW[, c("isoform", setdiff(colnames(df_SW), "isoform"))]
  
  write.table(df_SW,
              file.path(output_dir, paste0(cond_S, "_vs_", cond_W, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
}
```

```{r}
res_HS <- read.delim(here("results","DE","tier3_DE","H12_vs_S12.tsv"))
res_HW <- read.delim(here("results","DE","tier3_DE","H12_vs_W12.tsv"))
res_SW <- read.delim(here("results","DE","tier3_DE","S12_vs_W12.tsv"))

DE_test <- merge(res_HS, res_HW, by="isoform")
DE_test <- merge(DE_test, res_SW[,c("isoform","padj")], by="isoform")

dim(DE_test)
head(DE_test)
```


```{r}
file_to_check <- here("results", "DE", "tier3_DE", "H12_vs_S12.tsv")
head(read.delim(file_to_check))
colnames(read.delim(file_to_check))
```


```{r}
# Path to the file
file_to_check <- here("results", "DE", "tier3_DE", "H15_vs_S15.tsv")

# Load it
res_test <- read.delim(file_to_check)

# Basic structure
cat("Dimensions:\n")
print(dim(res_test))

cat("\nColumn names:\n")
print(colnames(res_test))

# Peek at the first rows
cat("\nHead:\n")
print(head(res_test))

# Summary of key columns
cat("\nSummary of log2FoldChange:\n")
print(summary(res_test$log2FoldChange))

cat("\nSummary of padj:\n")
print(summary(res_test$padj))

# Count DE genes
cat("\nNumber of DE isoforms (padj < 0.05):\n")
print(sum(res_test$padj < 0.05, na.rm = TRUE))

# Check for NA in critical columns
cat("\nAny NA in log2FoldChange?\n")
print(any(is.na(res_test$log2FoldChange)))

cat("\nAny NA in padj?\n")
print(any(is.na(res_test$padj)))
```

# Build sample-to-condition mapping
```{r sample_mapping}

# sample_info must contain columns: condition, sample
# This creates a list where each element is a vector of TPM column names
sample_map <- split(sample_info$sample, sample_info$condition)

# Example:
# sample_map$H12 gives all TPM columns for Hybrid at 12°C
# sample_map$S15 gives all TPM columns for South at 15°C
# sample_map$W18 gives all TPM columns for West at 18°C

sample_map
```

Expression patterns
│
├── Normal inheritance
│     (H is NOT DE from at least one pure line)
│
│     ├── Conserved
│     ├── Additive
│     └── Parent-like
│           ├── Parent-like S
│           └── Parent-like W
│
└── Misexpression
      (H is DE from BOTH pure lines)
      |
      ├── Within‑range misregulated
      │     (H between S and W)
      │
      └── Outside‑range misregulated (Transgressive)
            ├── Transgressive up
            └── Transgressive down
            

# Expression Classification for Tier 3 (per temperature)
```{r express_classification}

classify_expression <- function(temp) {
  
  # Load DE results
  res_HS <- read.delim(here("results", "DE", "tier3_DE",
                            paste0("H", temp, "_vs_S", temp, ".tsv")))
  res_HW <- read.delim(here("results", "DE", "tier3_DE",
                            paste0("H", temp, "_vs_W", temp, ".tsv")))
  res_SW <- read.delim(here("results", "DE", "tier3_DE",
                            paste0("S", temp, "_vs_W", temp, ".tsv")))
  
  # Merge DE tables
  DE <- merge(res_HS, res_HW, by = "isoform", suffixes = c("_HS", "_HW"))
  
  res_SW_small <- res_SW[, c("isoform", "padj")]
  colnames(res_SW_small)[2] <- "padj_SW"
  
  DE <- merge(DE, res_SW_small, by = "isoform")
  
  # Extract padj
  padj_HS <- DE$padj_HS
  padj_HW <- DE$padj_HW
  padj_SW <- DE$padj_SW
  
  # Get TPM columns for this temperature
  H_cols <- sample_map[[paste0("H", temp)]]
  S_cols <- sample_map[[paste0("S", temp)]]
  W_cols <- sample_map[[paste0("W", temp)]]
  
  # Compute mean TPM per group
  meanH <- rowMeans(tpm_matrix[DE$isoform, H_cols, drop = FALSE])
  meanS <- rowMeans(tpm_matrix[DE$isoform, S_cols, drop = FALSE])
  meanW <- rowMeans(tpm_matrix[DE$isoform, W_cols, drop = FALSE])
  
  # DE flags
  DE_HS <- padj_HS < 0.05
  DE_HW <- padj_HW < 0.05
  DE_SW <- padj_SW < 0.05
  
  notDE_HS <- !DE_HS
  notDE_HW <- !DE_HW
  
  # Hybrid position
  H_between <- (meanH > pmin(meanS, meanW)) & (meanH < pmax(meanS, meanW))
  H_above   <- meanH > pmax(meanS, meanW)
  H_below   <- meanH < pmin(meanS, meanW)
  
  # Initialize category
  category <- rep("other", nrow(DE))
  
  # Normal inheritance
  conserved <- notDE_HS & notDE_HW & !DE_SW
  additive  <- notDE_HS & notDE_HW & H_between
  
  parent_like_S <- DE_SW & notDE_HS & DE_HW
  parent_like_W <- DE_SW & notDE_HW & DE_HS
  
  category[conserved]       <- "conserved"
  category[additive]        <- "additive"
  category[parent_like_S]   <- "parent_like_S"
  category[parent_like_W]   <- "parent_like_W"
  
  # Misexpression
  misregulated <- DE_HS & DE_HW
  
  within_mis <- misregulated & H_between
  trans_up   <- misregulated & H_above
  trans_down <- misregulated & H_below
  
  category[within_mis] <- "within_range_misregulated"
  category[trans_up]   <- "transgressive_up"
  category[trans_down] <- "transgressive_down"
  
  category[misregulated & !(within_mis | trans_up | trans_down)] <- "misregulated_other"
  
  DE$category    <- category
  DE$temperature <- temp
  
  # Add TPM means so we can analyze "other"
  DE$meanH <- meanH
  DE$meanS <- meanS
  DE$meanW <- meanW


  
  return(DE)
}
```

# Run classification for all temperatures
```{r temp_class}
DE_12 <- classify_expression(12)
DE_15 <- classify_expression(15)
DE_18 <- classify_expression(18)

DE_all <- bind_rows(DE_12, DE_15, DE_18)
table(DE_all$temperature, DE_all$category)

```
# Decompose "other" category into meaningful subgroups
```{r other_decomposition}

decompose_other <- function(DE) {
  
  # Keep only the "other" rows
  DEo <- DE[DE$category == "other", ]
  
  # Extract padj
  padj_HS <- DEo$padj_HS
  padj_HW <- DEo$padj_HW
  padj_SW <- DEo$padj_SW
  
  # DE flags
  DE_HS <- padj_HS < 0.05
  DE_HW <- padj_HW < 0.05
  DE_SW <- padj_SW < 0.05
  
  # Hybrid position
  meanH <- DEo$meanH
  meanS <- DEo$meanS
  meanW <- DEo$meanW
  
  H_between <- (meanH > pmin(meanS, meanW)) & (meanH < pmax(meanS, meanW))
  
  # Subcategories inside "other"
  only_HS <- DE_HS & !DE_HW
  only_HW <- DE_HW & !DE_HS
  only_SW <- DE_SW & !DE_HS & !DE_HW
  
  low_expr <- is.na(meanH) | is.na(meanS) | is.na(meanW)
  
  ambiguous_between <- H_between & !(DE_HS | DE_HW | DE_SW)
  
  discordant <- (DE_HS | DE_HW) & H_between
  
  # Return counts
  data.frame(
    only_HS = sum(only_HS, na.rm = TRUE),
    only_HW = sum(only_HW, na.rm = TRUE),
    only_SW = sum(only_SW, na.rm = TRUE),
    low_expr = sum(low_expr, na.rm = TRUE),
    ambiguous_between = sum(ambiguous_between, na.rm = TRUE),
    discordant = sum(discordant, na.rm = TRUE)
  )
}

# Apply to each temperature
other_12 <- decompose_other(DE_12)
other_15 <- decompose_other(DE_15)
other_18 <- decompose_other(DE_18)

other_12
other_15
other_18
```
Most isoforms in the ‘other’ category showed asymmetric divergence, where the hybrid differed from one parent but not the other, while the parents themselves did not differ. Because parent‑like inheritance requires parental divergence, these isoforms do not fit into the parent‑like category. They also do not meet criteria for additive or misexpressed patterns, and therefore remain unclassified.

# Summary of expression categories across temperatures
```{r summarize_categories}
summary_table <- DE_all %>%
  group_by(temperature, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = category, values_from = n, values_fill = 0)

summary_table

```

# Visualization: Stacked barplot of categories per temperature
```{r plot_categories}

library(ggplot2)

ggplot(DE_all, aes(x = factor(temperature), fill = category)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Temperature (°C)",
    y = "Proportion of isoforms",
    fill = "Category",
    title = "Distribution of expression inheritance categories across temperatures"
  ) +
  theme_minimal()
```

# Visualization: Focus on rare categories
```{r plot_rare_categories}

rare_cats <- c("parent_like_S", "parent_like_W",
               "transgressive_up", "transgressive_down",
               "within_range_misregulated")

DE_rare <- DE_all %>% 
  filter(category %in% rare_cats)

ggplot(DE_rare, aes(x = factor(temperature), fill = category)) +
  geom_bar(position = "stack") +
  labs(
    x = "Temperature (°C)",
    y = "Number of isoforms",
    fill = "Category",
    title = "Rare inheritance categories across temperatures"
  ) +
  theme_minimal()
```

Most misregulated isoforms in hybrids show transgressive expression, meaning hybrid expression lies outside the parental range. Transgressive misexpression is the most extreme form of regulatory incompatibility and is often associated with reduced hybrid fitness.

# Export ALL misexpressed isoforms per temperature (combined)
```{r export_misexpressed_combined}

# Misexpression categories
mis_cats <- c("transgressive_up",
              "transgressive_down",
              "within_range_misregulated")

# Output directory
output_dir <- "results/DE/tier3_DE/misexpressed"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Loop through temperatures
for (temp in c(12, 15, 18)) {
  
  DE_temp <- DE_all %>% filter(temperature == temp)
  
  mis_temp <- DE_temp %>% filter(category %in% mis_cats)
  
  file <- file.path(output_dir, paste0("misexpressed_", temp, "C.tsv"))
  
  write.table(
    mis_temp,
    file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}
```

