---
title: "Tier 3 Differential Expression Analysis (Hybrid Inheritance)"
author: "Carlota Myhre de Gouveia"
date: "2026-04-24"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(
  root.dir   = here::here())
```


## Package Loading
```{r packages}

suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(ashr)
  library(readxl)
  library(rlang)
  library(ggVennDiagram)
  library(DT)
	library(here)
})

```


## Data Import and Preprocessing
```{r load_and_inspect_data}

## Define isoform-level abundance files
counts_file <- "results/assembly/trinity/abundance_estimation/compiled_abundance/abundance_matrix_isoforms.isoform.counts.matrix"
tpm_file    <- "results/assembly/trinity/abundance_estimation/compiled_abundance/abundance_matrix_isoforms.isoform.TPM.not_cross_norm"

## Load raw matrices
counts_raw <- read.delim(counts_file, header = TRUE, sep = "\t", check.names = FALSE)
tpm_raw    <- read.delim(tpm_file,    header = TRUE, sep = "\t", check.names = FALSE)

# Fix first column name
colnames(counts_raw)[1] <- "isoform"
colnames(tpm_raw)[1]    <- "isoform"

# Remove metadata column if present
counts_raw <- counts_raw[, !colnames(counts_raw) %in% "abundance_estimation"]
tpm_raw    <- tpm_raw[, !colnames(tpm_raw) %in% "abundance_estimation"]

# Set isoform IDs as rownames
rownames(counts_raw) <- counts_raw$isoform
rownames(tpm_raw)    <- tpm_raw$isoform

# Drop the isoform column
counts_raw <- counts_raw[, -1]
tpm_raw    <- tpm_raw[, -1]

# Convert to numeric safely
abundance_matrix <- as.data.frame(lapply(counts_raw, as.numeric))
tpm_matrix       <- as.data.frame(lapply(tpm_raw,    as.numeric))

rownames(abundance_matrix) <- rownames(counts_raw)
rownames(tpm_matrix)       <- rownames(tpm_raw)

# Remove rows with any NA
keep_complete <- complete.cases(abundance_matrix) & complete.cases(tpm_matrix)
abundance_matrix <- abundance_matrix[keep_complete, ]
tpm_matrix       <- tpm_matrix[keep_complete, ]

# Fix sample names to match Excel metadata
colnames(abundance_matrix) <- sub(".*Melops", "", colnames(abundance_matrix))
colnames(tpm_matrix)       <- sub(".*Melops", "", colnames(tpm_matrix))


# Load sample metadata from Excel
sample_info <- readxl::read_excel("data/sample_conditions_rnaseq.xlsx") %>%
  mutate(
    sample = sub("L", "", RNA_Later_ID),   # create 'sample' column to match abundance matrix
    origin = factor(Origin, levels = c("W","H","S")),
    temperature = factor(`Temperature (celsius)`, levels = c(12,15,18)),
    condition = factor(RNAseqgroup,
                       levels = c("W12","W15","W18",
                                  "H12","H15","H18",
                                  "S12","S15","S18")),
    length = scale(`Total length (mm)`)
  ) %>%
  select(sample, origin, temperature, condition, length) %>%
  arrange(match(sample, colnames(abundance_matrix)))

# Preview matrices and metadata
knitr::kable(abundance_matrix[1:10, 1:10],
             caption = "Isoform count matrix (first 10 isoforms × 10 samples)")
knitr::kable(tpm_matrix[1:10, 1:10],
             caption = "Isoform TPM matrix (first 10 isoforms × 10 samples)")
knitr::kable(sample_info, caption = "Sample metadata")
```

## Data Cleaning and Filtering
```{r filtering}

# Align column names with sample metadata
colnames(abundance_matrix) <- sample_info$sample
colnames(tpm_matrix)       <- sample_info$sample

# Convert to matrices
abund_num <- as.matrix(abundance_matrix)
tpm_num   <- as.matrix(tpm_matrix)

# Filtering rules
keep_counts <- rowSums(abund_num) >= 10
keep_tpm    <- rowSums(tpm_num >= 1) >= 2

# Combined filter
keep <- keep_counts & keep_tpm

# Apply filter
abundance_matrix <- abundance_matrix[keep, ]
tpm_matrix       <- tpm_matrix[keep, ]

# Report
initial_n <- nrow(abund_num)
final_n   <- nrow(abundance_matrix)

cat("Initial isoforms:", initial_n, "\n")
cat("After filtering:", final_n, "\n")
cat("Retained:", round(final_n / initial_n * 100, 2), "%\n")
```

## DESeq2 Model Setup
```{r deseq_setup}
# Ensure condition is a factor with correct ordering
sample_info$condition <- factor(
  sample_info$condition,
  levels = c(
    "H12","S12","W12",
    "H15","S15","W15",
    "H18","S18","W18"
  )
)

# Convert to integer matrix (RSEM expected counts → DESeq2)
abund_num <- round(as.matrix(abundance_matrix))
storage.mode(abund_num) <- "integer"

# Build DESeq2 dataset with length covariate
dds <- DESeqDataSetFromMatrix(
  countData = abund_num,
  colData   = sample_info,
  design    = ~ length + condition
)

# Run DESeq2
dds <- DESeq(dds)

cat("DESeq2 model fitted for Tier 3 analysis\n")
```
## Differential Expression for Tier 3 Contrasts
```{r DE_contrasts}

# Define output directory 
output_dir <- "results/DE/tier3_DE"

temps <- c(12, 15, 18)

for (t in temps) {
  
  cond_H <- paste0("H", t)
  cond_S <- paste0("S", t)
  cond_W <- paste0("W", t)
  
  # -------------------------
  # H vs S
  # -------------------------
  message("\nRunning contrast: ", cond_H, " vs ", cond_S)
  
  res_HS <- results(dds, contrast = c("condition", cond_H, cond_S))
  res_HS_shrunk <- lfcShrink(dds, contrast = c("condition", cond_H, cond_S),
                             res = res_HS, type = "ashr")
  
  iso <- rownames(res_HS_shrunk)
  df_HS <- as.data.frame(res_HS_shrunk)
  df_HS$isoform <- iso
  df_HS <- df_HS[, c("isoform", setdiff(colnames(df_HS), "isoform"))]
  
  write.table(df_HS,
              file.path(output_dir, paste0(cond_H, "_vs_", cond_S, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  
  # -------------------------
  # H vs W
  # -------------------------
  message("Running contrast: ", cond_H, " vs ", cond_W)
  
  res_HW <- results(dds, contrast = c("condition", cond_H, cond_W))
  res_HW_shrunk <- lfcShrink(dds, contrast = c("condition", cond_H, cond_W),
                             res = res_HW, type = "ashr")
  
  iso <- rownames(res_HW_shrunk)
  df_HW <- as.data.frame(res_HW_shrunk)
  df_HW$isoform <- iso
  df_HW <- df_HW[, c("isoform", setdiff(colnames(df_HW), "isoform"))]
  
  write.table(df_HW,
              file.path(output_dir, paste0(cond_H, "_vs_", cond_W, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  
  # -------------------------
  # S vs W (parental contrast)
  # -------------------------
  message("Running parental contrast: ", cond_S, " vs ", cond_W)
  
  res_SW <- results(dds, contrast = c("condition", cond_S, cond_W))
  res_SW_shrunk <- lfcShrink(dds, contrast = c("condition", cond_S, cond_W),
                             res = res_SW, type = "ashr")
  
  iso <- rownames(res_SW_shrunk)
  df_SW <- as.data.frame(res_SW_shrunk)
  df_SW$isoform <- iso
  df_SW <- df_SW[, c("isoform", setdiff(colnames(df_SW), "isoform"))]
  
  write.table(df_SW,
              file.path(output_dir, paste0(cond_S, "_vs_", cond_W, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
}
```


## Build sample-to-condition mapping
```{r sample_mapping}

# sample_info must contain columns: condition, sample
# This creates a list where each element is a vector of TPM column names
sample_map <- split(sample_info$sample, sample_info$condition)

```

## Classification scheme for hybrid expression patterns
```text
Expression patterns
│
├── Normal inheritance
│     (H is NOT DE from at least one pure line)
│
│     ├── Conserved
│     ├── Additive
│     └── Parent-like
│           ├── Parent-like S
│           └── Parent-like W
│
└── Misexpression
      (H is DE from BOTH pure lines)
      |
      ├── Within‑range misregulated
      │     (H between S and W)
      │
      └── Outside‑range misregulated (Transgressive)
            ├── Transgressive up
            └── Transgressive down
```

## Expression Classification for Tier 3 (per temperature)
```{r express_classification}

classify_expression <- function(temp) {
  
  # Load DE results for this temperature
  res_HS <- read.delim(file.path("results", "DE", "tier3_DE",
                                 paste0("H", temp, "_vs_S", temp, ".tsv")))
  res_HW <- read.delim(file.path("results", "DE", "tier3_DE",
                                 paste0("H", temp, "_vs_W", temp, ".tsv")))
  res_SW <- read.delim(file.path("results", "DE", "tier3_DE",
                                 paste0("S", temp, "_vs_W", temp, ".tsv")))
  
  # Merge DE tables
  DE <- merge(res_HS, res_HW, by = "isoform", suffixes = c("_HS", "_HW"))
  
  res_SW_small <- res_SW[, c("isoform", "padj")]
  colnames(res_SW_small)[2] <- "padj_SW"
  
  DE <- merge(DE, res_SW_small, by = "isoform")
  
  # Extract padj
  padj_HS <- DE$padj_HS
  padj_HW <- DE$padj_HW
  padj_SW <- DE$padj_SW
  
  # Get TPM columns for this temperature
  H_cols <- sample_map[[paste0("H", temp)]]
  S_cols <- sample_map[[paste0("S", temp)]]
  W_cols <- sample_map[[paste0("W", temp)]]
  
  # Compute mean TPM per group
  meanH <- rowMeans(tpm_matrix[DE$isoform, H_cols, drop = FALSE])
  meanS <- rowMeans(tpm_matrix[DE$isoform, S_cols, drop = FALSE])
  meanW <- rowMeans(tpm_matrix[DE$isoform, W_cols, drop = FALSE])
  
  # DE flags
  DE_HS <- padj_HS < 0.05
  DE_HW <- padj_HW < 0.05
  DE_SW <- padj_SW < 0.05
  
  notDE_HS <- !DE_HS
  notDE_HW <- !DE_HW
  
  # Hybrid position
  H_between <- (meanH > pmin(meanS, meanW)) & (meanH < pmax(meanS, meanW))
  H_above   <- meanH > pmax(meanS, meanW)
  H_below   <- meanH < pmin(meanS, meanW)
  
  # Initialize category
  category <- rep("other", nrow(DE))
  
  # Normal inheritance
  conserved <- notDE_HS & notDE_HW & !DE_SW
  additive  <- notDE_HS & notDE_HW & H_between
  
  parent_like_S <- DE_SW & notDE_HS & DE_HW
  parent_like_W <- DE_SW & notDE_HW & DE_HS
  
  category[conserved]       <- "conserved"
  category[additive]        <- "additive"
  category[parent_like_S]   <- "parent_like_S"
  category[parent_like_W]   <- "parent_like_W"
  
  # Misexpression
  misregulated <- DE_HS & DE_HW
  
  within_mis <- misregulated & H_between
  trans_up   <- misregulated & H_above
  trans_down <- misregulated & H_below
  
  category[within_mis] <- "within_range_misregulated"
  category[trans_up]   <- "transgressive_up"
  category[trans_down] <- "transgressive_down"
  
  category[misregulated & !(within_mis | trans_up | trans_down)] <- "misregulated_other"
  
  DE$category    <- category
  DE$temperature <- temp
  
  # Add TPM means so we can analyze "other"
  DE$meanH <- meanH
  DE$meanS <- meanS
  DE$meanW <- meanW
  
  return(DE)
}
```

## Run classification for all temperatures
```{r temp_class}
DE_12 <- classify_expression(12)
DE_15 <- classify_expression(15)
DE_18 <- classify_expression(18)

DE_all <- bind_rows(DE_12, DE_15, DE_18)
table(DE_all$temperature, DE_all$category)

```
## Decompose "other" category into meaningful subgroups
```{r other_decomposition}


decompose_other <- function(DE) {
  
  # Keep only the "other" rows
  DEo <- DE[DE$category == "other", ]
  
  # Extract padj
  padj_HS <- DEo$padj_HS
  padj_HW <- DEo$padj_HW
  padj_SW <- DEo$padj_SW
  
  # DE flags
  DE_HS <- padj_HS < 0.05
  DE_HW <- padj_HW < 0.05
  DE_SW <- padj_SW < 0.05
  
  # Hybrid position
  meanH <- DEo$meanH
  meanS <- DEo$meanS
  meanW <- DEo$meanW
  
  H_between <- (meanH > pmin(meanS, meanW)) & (meanH < pmax(meanS, meanW))
  
  # Subcategories inside "other"
  only_HS <- DE_HS & !DE_HW
  only_HW <- DE_HW & !DE_HS
  only_SW <- DE_SW & !DE_HS & !DE_HW
  discordant <- (DE_HS | DE_HW) & H_between
  
  # Counts
  counts <- c(
    only_HS = sum(only_HS, na.rm = TRUE),
    only_HW = sum(only_HW, na.rm = TRUE),
    only_SW = sum(only_SW, na.rm = TRUE),
    discordant = sum(discordant, na.rm = TRUE)
  )
  
  # Residual = total "other" minus sum of subcategories
  residual <- nrow(DEo) - sum(counts)
  
  data.frame(t(c(counts, residual = residual)))
}

# Apply to each temperature
other_12 <- decompose_other(DE_12)
other_15 <- decompose_other(DE_15)
other_18 <- decompose_other(DE_18)

knitr::kable(other_12, caption = "Decomposition of 'other' isoforms at 12°C")
knitr::kable(other_15, caption = "Decomposition of 'other' isoforms at 15°C")
knitr::kable(other_18, caption = "Decomposition of 'other' isoforms at 18°C")
```

We decomposed the heterogeneous ‘other’ category into interpretable subgroups. Isoforms classified as only_HS or only_HW showed asymmetric divergence, differing from one parent but not the other, but were not considered parent‑like because parental divergence was absent. The only_SW group captured parental divergence without hybrid divergence. Discordant cases represented hybrids positioned between parents yet still significantly diverging from one or both. The majority of isoforms did not meet these definitions and are reported as a residual group, reflecting missing values or heterogeneous patterns that cannot be assigned to a specific inheritance category under our criteria.

## Summary of expression categories across temperatures
```{r summarize_categories}
summary_table <- DE_all %>%
  group_by(temperature, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = category, values_from = n, values_fill = 0)

summary_table

```


## Visualization: Stacked barplot of categories per temperature
```{r plot_categories}

library(ggplot2)

ggplot(DE_all, aes(x = factor(temperature), fill = category)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Temperature (°C)",
    y = "Proportion of isoforms",
    fill = "Category",
    title = "Distribution of expression inheritance categories across temperatures"
  ) +
  theme_minimal()
```


The majority of isoforms fall into the conserved and additive categories, showing that hybrid expression is largely consistent with parental regulation or intermediate between parental levels without being differentially expressed from either parent. A substantial fraction also falls into the other category, which reflects asymmetric divergence where the hybrid differs from one parent but not the other, while the parents themselves do not differ. Because parent‑like inheritance requires parental divergence, these isoforms cannot be assigned to parent‑like categories and remain unclassified.

At 12°C, balanced replication (n=5 per origin) meant that DESeq2 produced valid padj values for nearly all isoforms. As a result, our classification function could assign most isoforms cleanly, leading to a large increase in the “conserved” category (86,896) and a much smaller “other” category (3,290). At 15°C and 18°C, however, the lower replication of South origin fish (n=2) limited statistical power in the parental contrast, producing many NA padj values and inflating the “other” category (37,799 and 43,112 isoforms, respectively).

## Visualization: Focus on rare categories
```{r plot_rare_categories}

rare_cats <- c("parent_like_S", "parent_like_W",
               "transgressive_up", "transgressive_down",
               "within_range_misregulated")

DE_rare <- DE_all %>% 
  filter(category %in% rare_cats)

ggplot(DE_rare, aes(x = factor(temperature), fill = category)) +
  geom_bar(position = "stack") +
  labs(
    x = "Temperature (°C)",
    y = "Number of isoforms",
    fill = "Category",
    title = "Rare inheritance categories across temperatures"
  ) +
  theme_minimal()
```

Most misregulated isoforms in hybrids show transgressive expression, meaning hybrid expression lies outside the parental range. Transgressive misexpression is the most extreme form of regulatory incompatibility and is often associated with reduced hybrid fitness.

## Export ALL misexpressed isoforms per temperature (combined)
```{r export_misexpressed_combined}

# Ensure output directory exists
dir.create("results/DE/tier3_DE/misexpressed", recursive = TRUE, showWarnings = FALSE)

mis_cats <- c("transgressive_up", "transgressive_down", "within_range_misregulated")

for (temp in c(12, 15, 18)) {

  iso <- DE_all %>%
    filter(temperature == temp, category %in% mis_cats) %>%
    pull(isoform) %>%
    unique()

  write.table(
    iso,
    paste0("results/DE/tier3_DE/misexpressed/misexpressed_", temp, "C.tsv"),
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
}
```


## Export parent like isoforms per temperature
```{r export_parent_like}
# Ensure output directory exists
dir.create("results/DE/tier3_DE/parent_like", recursive = TRUE, showWarnings = FALSE)

for (cat in c("parent_like_S", "parent_like_W")) {
  for (temp in c(12, 15, 18)) {

    iso <- DE_all %>%
      filter(temperature == temp, category == cat) %>%
      pull(isoform) %>% 
      unique()

    write.table(
      iso,
      paste0("results/DE/tier3_DE/parent_like/", cat, "_", temp, "C.tsv"),
      quote = FALSE, row.names = FALSE, col.names = FALSE
    )
  }
}
```

## Summarize DE counts per contrast and temperature
```{r contrast_summary}

contrast_summary <- data.frame()

for (temp in c(12, 15, 18)) {
  
  # Load contrasts using relative paths
  HS <- read.delim(file.path("results","DE","tier3_DE", paste0("H",temp,"_vs_S",temp,".tsv")))
  HW <- read.delim(file.path("results","DE","tier3_DE", paste0("H",temp,"_vs_W",temp,".tsv")))
  SW <- read.delim(file.path("results","DE","tier3_DE", paste0("S",temp,"_vs_W",temp,".tsv")))
  
  # Count DE genes (padj < 0.05)
  n_HS <- sum(HS$padj < 0.05, na.rm = TRUE)
  n_HW <- sum(HW$padj < 0.05, na.rm = TRUE)
  n_SW <- sum(SW$padj < 0.05, na.rm = TRUE)
  
  # Add row
  contrast_summary <- rbind(
    contrast_summary,
    data.frame(
      temperature = temp,
      DE_HS = n_HS,
      DE_HW = n_HW,
      DE_SW = n_SW
    )
  )
}

knitr::kable(contrast_summary, caption = "Counts of DE isoforms per contrast and temperature")
```

## Misexpressed isoforms overlap
```{r missexpress_overlap}
# Read misexpressed isoforms per temperature
mis_list <- list(
  `12°C` = read.delim("results/DE/tier3_DE/misexpressed/misexpressed_12C.tsv", header = FALSE)$V1,
  `15°C` = read.delim("results/DE/tier3_DE/misexpressed/misexpressed_15C.tsv", header = FALSE)$V1,
  `18°C` = read.delim("results/DE/tier3_DE/misexpressed/misexpressed_18C.tsv", header = FALSE)$V1
)

# Plot
ggVennDiagram(mis_list, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  ggtitle("Overlap of Misexpressed Isoforms Across Temperatures")
```
## Parent‑like S overlap
```{r S_like_overlap}
pls_list <- list(
  `12°C` = read.delim("results/DE/tier3_DE/parent_like/parent_like_S_12C.tsv", header = FALSE)$V1,
  `15°C` = read.delim("results/DE/tier3_DE/parent_like/parent_like_S_15C.tsv", header = FALSE)$V1,
  `18°C` = read.delim("results/DE/tier3_DE/parent_like/parent_like_S_18C.tsv", header = FALSE)$V1
)

ggVennDiagram(pls_list, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#009E73") +
  ggtitle("Overlap of Parent-like S Isoforms Across Temperatures")
```

## Parent‑like W overlap
```{r W_like_overlap}
plw_list <- list(
  `12°C` = read.delim("results/DE/tier3_DE/parent_like/parent_like_W_12C.tsv", header = FALSE)$V1,
  `15°C` = read.delim("results/DE/tier3_DE/parent_like/parent_like_W_15C.tsv", header = FALSE)$V1,
  `18°C` = read.delim("results/DE/tier3_DE/parent_like/parent_like_W_18C.tsv", header = FALSE)$V1
)

ggVennDiagram(plw_list, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#D55E00") +
  ggtitle("Overlap of Parent-like W Isoforms Across Temperatures")
```

## Define Tier 3 DE file paths
```{r define_files}
## Define Tier 3 DE file paths
tier3_files <- list(
  misexpressed_12C   = "results/DE/tier3_DE/misexpressed/misexpressed_12C.tsv",
  misexpressed_15C   = "results/DE/tier3_DE/misexpressed/misexpressed_15C.tsv",
  misexpressed_18C   = "results/DE/tier3_DE/misexpressed/misexpressed_18C.tsv",
  parent_like_S_12C  = "results/DE/tier3_DE/parent_like/parent_like_S_12C.tsv",
  parent_like_S_15C  = "results/DE/tier3_DE/parent_like/parent_like_S_15C.tsv",
  parent_like_S_18C  = "results/DE/tier3_DE/parent_like/parent_like_S_18C.tsv",
  parent_like_W_12C  = "results/DE/tier3_DE/parent_like/parent_like_W_12C.tsv",
  parent_like_W_15C  = "results/DE/tier3_DE/parent_like/parent_like_W_15C.tsv",
  parent_like_W_18C  = "results/DE/tier3_DE/parent_like/parent_like_W_18C.tsv"
)
```

## Load EggNOG transcript‑level annotation
```{r load_eggnog_anns}
eggnog <- read.delim(
  "results/annotation/eggnog/Trinity.emapper.annotations",
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = ""
)

colnames(eggnog) <- c(
  "query_name", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
  "max_annot_lvl", "COG_category", "Description", "Preferred_name",
  "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
  "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
  "CAZy", "BiGG_Reaction", "PFAMs"
)

DT::datatable(
  head(eggnog, 10),
  options = list(pageLength = 5, scrollX = TRUE),
  caption = "Preview of EggNOG transcript-level annotations"
)
```
## Build gene-level GO annotation table from EggNOG output
```{r build_gene_go_mapping}
## Build gene-level GO annotation table
eggnog_go <- eggnog %>%
  dplyr::filter(GOs != "" & !is.na(GOs)) %>%
  dplyr::select(trinity_id = query_name, GOs) %>%
  dplyr::mutate(
    trinity_id = sub("\\.p[0-9]+$", "", trinity_id),  # remove protein suffix
    gene_id    = sub("_i[0-9]+$", "", trinity_id)     # collapse isoforms → gene
  )

gene2go <- eggnog_go %>%
  tidyr::separate_rows(GOs, sep = ",|;") %>%
  dplyr::mutate(GOs = stringr::str_trim(GOs)) %>%
  dplyr::filter(GOs != "") %>%
  dplyr::distinct(gene_id, GOs)

head(gene2go)

## Universe of annotated genes
universe_genes <- unique(gene2go$gene_id)

```
## Build annotation coverage table and extract gene lists for Venn diagrams
```{r ann_cov}
gene_lists <- list()

annotation_summary <- lapply(names(tier3_files), function(lbl) {

  # Load transcript IDs for this Tier 3 category
  df <- read.delim(tier3_files[[lbl]], header = FALSE)
  colnames(df) <- "transcript_id"
  transcripts <- df$transcript_id

  # Collapse isoforms to gene IDs (e.g., TRINITY_DN123_c0_g1_i2 → TRINITY_DN123_c0_g1)
  genes <- unique(sub("_i[0-9]+$", "", transcripts))

  # Store gene IDs for later gene-level Venn diagrams
  gene_lists[[lbl]] <<- genes

  # Count how many transcripts and genes have EggNOG GO annotation
  annotated_transcripts <- sum(transcripts %in% eggnog_go$trinity_id)
  annotated_genes <- sum(genes %in% universe_genes)

  # Return summary row
  data.frame(
    category = lbl,
    total_transcripts = length(transcripts),
    annotated_transcripts = annotated_transcripts,
    percent_transcripts_annot = round(annotated_transcripts / length(transcripts) * 100, 2),
    total_genes = length(genes),
    annotated_genes = annotated_genes,
    percent_genes_annot = round(annotated_genes / length(genes) * 100, 2)
  )
})

annotation_summary <- dplyr::bind_rows(annotation_summary)


datatable(
  annotation_summary,
  extensions = "Buttons",
  options = list(
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel"),
    pageLength = 10,
    scrollX = TRUE
  ),
  caption = "Annotation Coverage for Tier 3 Differential Expression Categories"
)

```

## Gene-level Venn diagram for misexpressed genes
```{r misexpresses_venn_genes}
mis_genes <- list(
  `12°C` = gene_lists$misexpressed_12C,
  `15°C` = gene_lists$misexpressed_15C,
  `18°C` = gene_lists$misexpressed_18C
)

ggVennDiagram(mis_genes, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  ggtitle("Overlap of Misexpressed Genes Across Temperatures")
```


The number of misexpressed genes remains relatively stable across temperatures, suggesting that hybrid misregulation is not strongly temperature‑dependent but represents a consistent background of regulatory incompatibility. 

## Gene-level Venn diagram for parent-like S genes
```{r S_like_venn_genes}
pls_genes <- list(
  `12°C` = gene_lists$parent_like_S_12C,
  `15°C` = gene_lists$parent_like_S_15C,
  `18°C` = gene_lists$parent_like_S_18C
)

ggVennDiagram(pls_genes, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#009E73") +
  ggtitle("Overlap of Parent-like S Genes Across Temperatures")
```


South‑like inheritance shows a dip at 15°C but is otherwise similar across temperature treatments.

## Gene-level Venn diagram for parent-like W genes
```{r W_like_venn_genes}
plw_genes <- list(
  `12°C` = gene_lists$parent_like_W_12C,
  `15°C` = gene_lists$parent_like_W_15C,
  `18°C` = gene_lists$parent_like_W_18C
)

ggVennDiagram(plw_genes, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#D55E00") +
  ggtitle("Overlap of Parent-like W Genes Across Temperatures")
```


West‑like inheritance increases markedly with temperature, indicating a stronger tendency for hybrids to inherit a western population expression under warmer conditions.

The limited overlaps across temperature treatments show that distinct gene sets are involved at each temperature, indicating treatment‑specific regulation rather than a fixed set of misexpressed or parent‑like genes.


## Intersections with Tiers 1 and 2b

### Load Tier 1 and Tier 2b gene sets
```{r load_genes_tier_1_2}
# Tier 1: shared temperature-responsive genes
tier1_genes <- read.csv(
  "results/functional_enrichment/tier1_shared_temperature_response/tier1_shared_temperature_genes.csv",
  header = TRUE
)$gene_id

# Tier 2b: thermal local adaptation candidates
tier2b_genes <- read.csv(
  "results/functional_enrichment/tier2_pop_divergence/tier2b_local_adaptation/tier2b_local_adaptation_candidates.csv",
  header = TRUE
)$gene_id
```
### Build Tier 3 gene sets (combined across temperatures)
```{r genes_tier_3}

# Misexpressed genes 
misexpressed <- unique(c(
  gene_lists$misexpressed_12C,
  gene_lists$misexpressed_15C,
  gene_lists$misexpressed_18C
))

# Parent-like S genes
parent_like_S <- unique(c(
  gene_lists$parent_like_S_12C,
  gene_lists$parent_like_S_15C,
  gene_lists$parent_like_S_18C
))

# Parent-like W genes
parent_like_W <- unique(c(
  gene_lists$parent_like_W_12C,
  gene_lists$parent_like_W_15C,
  gene_lists$parent_like_W_18C
))
```

### Summary table of tier 3 gene sets (temperature combined)
```{r sum_table_tier3_genes_combined}
## Summary table of Tier 3 gene sets
tier3_summary <- data.frame(
  Category = c("Misexpressed", "Parent-like S", "Parent-like W"),
  Count    = c(length(misexpressed),
               length(parent_like_S),
               length(parent_like_W))
)

DT::datatable(
  tier3_summary,
  options = list(pageLength = 5, scrollX = TRUE),
  caption = "Sizes of Tier 3 Gene Sets (combined across temperatures)"
)
```

```{r save_tier3_genes, echo=FALSE, results='hide'}
# Save Tier 3 gene sets
write_csv(data.frame(gene_id = misexpressed),
          "results/functional_enrichment/tier3_hybrid_inheritance/tier3_misexpressed_genes.csv")

write_csv(data.frame(gene_id = parent_like_S),
          "results/functional_enrichment/tier3_hybrid_inheritance/tier3_parent_like_S_genes.csv")

write_csv(data.frame(gene_id = parent_like_W),
          "results/functional_enrichment/tier3_hybrid_inheritance/tier3_parent_like_W_genes.csv")
```

### Intersections with Tier 1 and Tier 2b
```{r}
# Tier 1 intersections
tier1_misexpressed  <- intersect(tier1_genes, misexpressed)
tier1_parent_like_S <- intersect(tier1_genes, parent_like_S)
tier1_parent_like_W <- intersect(tier1_genes, parent_like_W)

# Tier 2b intersections 
tier2b_misexpressed  <- intersect(tier2b_genes, misexpressed)
tier2b_parent_like_S <- intersect(tier2b_genes, parent_like_S)
tier2b_parent_like_W <- intersect(tier2b_genes, parent_like_W)
```

### Summary table of Tier 1 & Tier 2b intersections
```{r}
intersection_summary <- data.frame(
  Category = c("Misexpressed", "Parent-like S", "Parent-like W"),
  
  Tier1_Count = c(
    length(tier1_misexpressed),
    length(tier1_parent_like_S),
    length(tier1_parent_like_W)
  ),
  
  Tier2b_Count = c(
    length(tier2b_misexpressed),
    length(tier2b_parent_like_S),
    length(tier2b_parent_like_W)
  )
)

DT::datatable(
  intersection_summary,
  options = list(pageLength = 5, scrollX = TRUE),
  caption = "Summary of Tier 1 and Tier 2b Intersections with Tier 3 Categories"
)
```
```{r save_intersections, echo=FALSE, results='hide'}
write_csv(tibble(gene_id = tier1_misexpressed),  "results/functional_enrichment/tier_intersections/tier1_tier3_misexpressed.csv")
write_csv(tibble(gene_id = tier1_parent_like_S), "results/functional_enrichment/tier_intersections/tier1_tier3_parent_like_S.csv")
write_csv(tibble(gene_id = tier1_parent_like_W), "results/functional_enrichment/tier_intersections/tier1_tier3_parent_like_W.csv")

write_csv(tibble(gene_id = tier2b_misexpressed),  "results/functional_enrichment/tier_intersections/tier2b_tier3_misexpressed.csv")
write_csv(tibble(gene_id = tier2b_parent_like_S), "results/functional_enrichment/tier_intersections/tier2b_tier3_parent_like_S.csv")
write_csv(tibble(gene_id = tier2b_parent_like_W), "results/functional_enrichment/tier_intersections/tier2b_tier3_parent_like_W.csv")
```


## Long-format table of all intersected genes
```{r}
intersection_genes <- list(
  Tier1_Misexpressed   = tier1_misexpressed,
  Tier1_ParentLike_S   = tier1_parent_like_S,
  Tier1_ParentLike_W   = tier1_parent_like_W,
  Tier2b_Misexpressed  = tier2b_misexpressed,
  Tier2b_ParentLike_S  = tier2b_parent_like_S,
  Tier2b_ParentLike_W  = tier2b_parent_like_W
)

intersection_genes_df <- stack(intersection_genes)
colnames(intersection_genes_df) <- c("Gene_ID", "Intersection")

DT::datatable(
  intersection_genes_df,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Gene IDs in Each Tier 1 and Tier 2b Intersection"
)
```


## Summary of annotation coverage for intersected gene sets.
```{r}
# Add annotation status
intersection_genes_df$Has_Annotation <- intersection_genes_df$Gene_ID %in% universe_genes

# Summary of annotation coverage
annotation_status_summary <- intersection_genes_df %>%
  dplyr::group_by(Intersection) %>%
  dplyr::summarise(
    Total_Genes = n(),
    Annotated = sum(Has_Annotation),
    Percent_Annotated = round(Annotated / Total_Genes * 100, 2)
  )

DT::datatable(
  annotation_status_summary,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Annotation Coverage for Intersected Gene Sets"
)
```

## Session Info
```{r session_info} 
sessionInfo()
```


