---
title: "Comparison and Integration of Annotation Sources"
author: "Carlota Gouveia"
date: "2026-04-24"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(
  root.dir   = here::here())
```

# Introduction

This report summarises, compares, and integrates the different annotation sources available for the *Symphodus melops* transcriptome, including EggNOG, InterProScan, Trinotate, and genome-based annotations.

## Load necessary packages
```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(ggplot2)
  library(UpSetR)
  library(readr)
  library(stringr)
})
```

## Load and Inspect Annotation Sources
```{r}

## EggNOG annotation
# Read file as raw text
egg_raw <- readLines("results/annotation/eggnog/Trinity.emapper.annotations")
# Convert literal "\t" into real tabs
egg_raw <- gsub("\\\\t", "\t", egg_raw)
# Identify the header line (the one starting with "#query")
header_line <- egg_raw[grepl("^#query", egg_raw)]
# Extract column names from the header
col_names <- strsplit(sub("^#", "", header_line), "\t")[[1]]
# Extract all data lines (those NOT starting with "#")
data_lines <- egg_raw[!grepl("^#", egg_raw)]
# Read the data into a table using the extracted header
eggnog <- read.delim(
  textConnection(data_lines),
  header = FALSE,
  sep = "\t",
  quote = "",
  stringsAsFactors = FALSE
)
# Assign the correct column names
colnames(eggnog) <- col_names


## InterProScan output
ips <- read_tsv(
  "results/annotation/ips/IPS_merged.tsv",
  col_names = FALSE
)

## Trinotate full report
trinotate_report <- read_tsv(
  "results/annotation/trinotate/trinotate_final/trinotate_annotation_report.xls"
)

## SignalP predictions
signalp <- read_tsv(
  "results/annotation/trinotate/signalp/signalp_merged.output.gff3",
  col_names = FALSE,
  comment = "##",
  show_col_types = FALSE
)
## Assign the correct column names
colnames(signalp) <- c(
  "seqid", "source", "type", "start", "end",
  "score", "strand", "phase", "attributes"
)


# DeepTMHMM GFF3
tmhmm <- read_tsv(
  "results/annotation/trinotate/tmhmm/tmhmm_clean.gff3",
  col_names = FALSE,
  comment = "##",
  show_col_types = FALSE
)
# Assign column names
colnames(tmhmm) <- c(
  "seqid", "source", "type", "start", "end",
  "score", "strand", "phase", "attributes"
)

## Genome-based annotation
genome_based <- read_tsv(
  "results/annotation/genome_based/trinities_filteringby_cov_summary.tsv"
)

# ---- Quick row-count inspection ----

cat("EggNOG rows:", nrow(eggnog), "\n")
cat("IPS rows:", nrow(ips), "\n")
cat("Trinotate rows:", nrow(trinotate_report), "\n")
cat("SignalP rows:", nrow(signalp), "\n")
cat("TMHMM GFF3 rows:", nrow(tmhmm), "\n")
cat("Genome-based rows:", nrow(genome_based), "\n")

# ---- Compact structure preview ----

cat("\nEggNOG (head):\n")
head(eggnog)

cat("\nIPS (head):\n")
head(ips)

cat("\nTrinotate report (head):\n")
head(trinotate_report)

cat("\nSignalP (head):\n")
head(signalp)

cat("\nTMHMM GFF3 (head):\n")
head(tmhmm)

cat("\nGenome-based annotation (head):\n")
head(genome_based)
```

## Standardize Trinity IDs across annotation sources
```{r extract_gene_ids}

# Helper: extract gene-level ID from any Trinity-style ID
extract_gene_id <- function(x) {
  x %>%
    str_remove("\\.p\\d+$") %>%   # remove .p1, .p2, etc.
    str_remove("_i\\d+$") %>%     # remove isoform _i1, _i2, etc.
    str_replace("^#", "")         # remove leading # if present
}

# ---- EggNOG ----
# first column is protein ID like TRINITY_GG_1000_c0_g1_i1.p1
eggnog <- eggnog %>%
  rename(protein_id = query) %>%
  mutate(
    transcript_id = str_remove(protein_id, "\\.p\\d+$"),
    gene_id       = extract_gene_id(transcript_id)
  )

# ---- IPS ----
ips <- ips %>%
  rename(protein_id = X1) %>%
  mutate(
    transcript_id = str_remove(protein_id, "\\.p\\d+$"),
    gene_id       = extract_gene_id(transcript_id)
  )

# ---- Trinotate ----
trinotate_report <- trinotate_report %>%
  rename(gene_id = `#gene_id`) %>%
  mutate(
    gene_id       = extract_gene_id(gene_id),
    transcript_id = transcript_id
  )

# ---- SignalP ----
signalp <- signalp %>%
  mutate(
    # Extract the transcript ID from the seqid field
    transcript_id = sub("\\.p[0-9]+.*$", "", seqid),

    # Derive gene_id (not used for merging)
    gene_id = sub("_i.*", "", transcript_id)
  )

# ---- TMHMM ----
tmhmm <- tmhmm %>%
  mutate(
    transcript_id = sub("\\.p[0-9]+$", "", seqid)
  )



# ---- Genome-based ----
genome_based <- genome_based %>%
  rename(transcript_id = Trinity_ID) %>%
  mutate(
    gene_id = extract_gene_id(transcript_id)
  )

```

## Integrate SignalP-6.0 into Trinotate report
```{r}
# Identify transcripts with a signal peptide (YES)
signalp_hits <- signalp %>%
  filter(type == "signal_peptide") %>%
  distinct(transcript_id) %>%      # <-- ONLY transcript_id
  mutate(SignalP = "YES")

# Remove old Trinotate SignalP column
trinotate_report <- trinotate_report %>% select(-SignalP)

# Build full transcript list from Trinotate
all_transcripts <- trinotate_report %>%
  select(transcript_id) %>%        # <-- ONLY transcript_id
  distinct()

# Create complete YES/NO table (join ONLY on transcript_id)
signalp_full <- all_transcripts %>%
  left_join(signalp_hits, by = "transcript_id") %>%   # <-- ONLY transcript_id
  mutate(SignalP = ifelse(is.na(SignalP), "NO", SignalP))

# Merge into Trinotate report
trinotate_report <- trinotate_report %>%
  left_join(signalp_full, by = "transcript_id")       # <-- ONLY transcript_id
```

## Integrate DeepTmhmm into Trinotate report
```{r}
# ---- Identify transcripts with ≥1 TM helix ----
tmhmm_hits <- tmhmm %>%
  filter(type == "TMhelix") %>%
  distinct(transcript_id) %>%
  mutate(TMHMM = "YES")

# ---- Build full transcript list ----
all_transcripts <- trinotate_report %>%
  select(transcript_id) %>%
  distinct()

tmhmm_full <- all_transcripts %>%
  left_join(tmhmm_hits, by = "transcript_id") %>%
  mutate(TMHMM = ifelse(is.na(TMHMM), "NO", TMHMM))

# ---- Merge TMHMM into Trinotate report ----
trinotate_report <- trinotate_report %>%
  { if ("TMHMM" %in% colnames(.)) select(., -TMHMM) else . } %>% 
  left_join(tmhmm_full, by = "transcript_id")
```

```{r}
colnames(eggnog)
```

## EggNog extraction function
```{r}
extract_eggnog_annotations <- function(gene_list, eggnog) {
  
  clean_field <- function(x) {
    x <- unique(na.omit(x))
    x <- x[x != "-" & x != ""]
    if (length(x) == 0) return(NA)
    x <- gsub(",", ";", x)   # normalize EggNOG comma lists
    paste(x, collapse = ";")
  }
  
  eggnog %>%
    filter(gene_id %in% gene_list) %>%
    mutate(GO_list = strsplit(GOs, ",")) %>%
    unnest(GO_list) %>%
    filter(!GO_list %in% c("GO:0003674","GO:0005575","GO:0008150")) %>% # drop roots
    group_by(gene_id) %>%
    summarise(
      egg_ortholog = clean_field(seed_ortholog),
      egg_desc     = clean_field(Description),
      egg_COG      = clean_field(COG_category),
      egg_KO       = clean_field(KEGG_ko),
      egg_KEGGmap  = clean_field(KEGG_Pathway),
      egg_PFAM     = clean_field(PFAMs),
      egg_GO       = paste(unique(GO_list), collapse=";"),  # deduplicate 
      .groups = "drop"
    )
}
```

## IPS extraction function
```{r}
extract_ips_annotations <- function(gene_list, ips) {
  
  clean_field <- function(x) {
    x <- gsub("[|,]", ";", x)      # normalize | and , to ;
    x <- gsub("^;|;$", "", x)      # trim leading/trailing ;
    x <- gsub(";;+", ";", x)       # collapse multiple ;;
    x[x == ""] <- NA               # empty -> NA
    x
  }
  
  ips %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      ips_IPR  = clean_field(
        paste0(na.omit(unique(X12[X12 != "-"])), collapse = ";")
      ),
      ips_PFAM = clean_field(
        paste0(na.omit(unique(X5[X5 != "-"])), collapse = ";")
      ),
      ips_GO   = clean_field(
        paste0(na.omit(unique(X14[X14 != "-"])), collapse = ";")
      ),
      .groups = "drop"
    )
}
```

## Trinotate Extraction Function
```{r}
extract_trinotate_annotations <- function(gene_list, trinotate_report) {
  
  clean_field <- function(x) {
    x <- gsub("^;|;$", "", x)
    x <- gsub(";;+", ";", x)
    x[x == ""] <- NA
    x
  }
  
  trinotate_report %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      trinotate_BLASTX = clean_field(
        paste0(
          na.omit(unique(sprot_Top_BLASTX_hit[sprot_Top_BLASTX_hit != "."])),
          collapse = ";"
        )
      ),
      
      trinotate_BLASTP = clean_field(
        paste0(
          na.omit(unique(sprot_Top_BLASTP_hit[sprot_Top_BLASTP_hit != "."])),
          collapse = ";"
        )
      ),
      
      trinotate_PFAM = clean_field(
        paste0(
          na.omit(unique(Pfam[Pfam != "."])),
          collapse = ";"
        )
      ),
      
      trinotate_GO = clean_field(
        paste0(
          na.omit(unique(c(
            gene_ontology_BLASTX[gene_ontology_BLASTX != "."],
            gene_ontology_BLASTP[gene_ontology_BLASTP != "."],
            gene_ontology_Pfam[gene_ontology_Pfam != "."]
          ))),
          collapse = ";"
        )
      ),
      
      trinotate_SignalP = ifelse(any(SignalP == "YES"), "YES", "NO"),
      trinotate_TMHMM   = ifelse(any(TMHMM == "YES"), "YES", "NO"),
      
      .groups = "drop"
    )
}
```


## Genome based annotation extraction function
```{r}
extract_genome_annotations <- function(gene_list, genome_based) {
  
  genome_based %>%
    filter(gene_id %in% gene_list) %>%
    # keep only the transcript with highest coverage
    group_by(gene_id) %>%
    slice_max(`Coverage_%`, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(
      gene_id,
      genome_best_transcript = transcript_id,   
      genome_gene_id    = Gene_ID,
      genome_coverage   = `Coverage_%`,
      genome_GO         = ifelse(GO_Terms == "" | GO_Terms == "NA", NA, GO_Terms),
      genome_UniProt    = ifelse(UniProt_ID == "" | UniProt_ID == "NA", NA, UniProt_ID),
      genome_Dbxref     = ifelse(Dbxref == "" | Dbxref == "NA", NA, Dbxref)
    )
}
```

## Function to integrate all annotatiosn sources in one single table for a set of genes
```{r}
integrate_annotations <- function(gene_list,
                                  eggnog,
                                  ips,
                                  trinotate_report,
                                  genome_based) {
  
  # Run each extraction function
  eggnog_ann   <- extract_eggnog_annotations(gene_list, eggnog)
  ips_ann      <- extract_ips_annotations(gene_list, ips)
  trinotate_ann <- extract_trinotate_annotations(gene_list, trinotate_report)
  genome_ann   <- extract_genome_annotations(gene_list, genome_based)
  
  # Sequentially join everything by gene_id
  integrated <- gene_list %>%
    tibble(gene_id = .) %>%
    left_join(eggnog_ann,   by = "gene_id") %>%
    left_join(ips_ann,      by = "gene_id") %>%
    left_join(trinotate_ann,by = "gene_id") %>%
    left_join(genome_ann,   by = "gene_id")
  
  return(integrated)
}
```

## Annotation classification function
```{r}
classify_annotations <- function(integrated_table) {
  
  coverage_threshold_strong <- 70
  coverage_threshold_moderate <- 40
  
  integrated_table %>%
    mutate(
      # Strong evidence flags
      has_eggnog   = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
      has_ips      = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
      has_blast    = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                              "trinotate_PFAM","trinotate_GO"),
                            ~ !is.na(.) & . != ""),
      genome_func  = if_any(c("genome_GO","genome_UniProt","genome_Dbxref"),
                            ~ !is.na(.) & . != ""),
      
      # Genome coverage categories
      genome_strong   = genome_func & genome_coverage >= coverage_threshold_strong,
      genome_moderate = genome_func & genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong,
      genome_weak     = genome_func & genome_coverage < coverage_threshold_moderate,
      genome_mapped_only = genome_coverage >= coverage_threshold_strong & !genome_func,
      
      # Weak evidence flags
      has_signalp  = trinotate_SignalP == "YES",
      has_tmhmm    = trinotate_TMHMM == "YES",
      genome_low   = genome_coverage < coverage_threshold_strong & !genome_func,
      
      # Count strong sources (EggNOG, IPS, BLAST, genome strong)
      strong_sources = (has_eggnog + has_ips + has_blast + genome_strong),
      
      # Classification
      annotation_class = case_when(
        strong_sources >= 2 ~ "Well-annotated",
        strong_sources == 1 | genome_moderate ~ "Partially annotated",
        genome_mapped_only ~ "Mapped only",
        genome_weak | has_signalp | has_tmhmm | genome_low ~ "Weakly annotated",
        TRUE ~ "Unannotated"
      )
    )
}
```

##  Create master table
```{r}
# Use Trinotate IDs as master list
all_gene_ids <- unique(trinotate_report$gene_id)

# Integrate once
integrated_table <- integrate_annotations(
  all_gene_ids,
  eggnog,
  ips,
  trinotate_report,
  genome_based
)

```

--------

```{r}
# Show all column names in your integrated table
colnames(integrated_table)

# Or, if you want a quick overview of the structure:
str(integrated_table)

# If you’re using dplyr/tidyverse, glimpse is handy:
glimpse(integrated_table)
```
```{r}
# Look at a random sample of BLASTX hits
integrated_table %>%
  select(gene_id, trinotate_BLASTX) %>%
  filter(!is.na(trinotate_BLASTX)) %>%
  sample_n(10)

# Same for BLASTP
integrated_table %>%
  select(gene_id, trinotate_BLASTP) %>%
  filter(!is.na(trinotate_BLASTP)) %>%
  sample_n(10)

# Look at a random sample of genome_Dbxref entries
integrated_table %>%
  select(gene_id, genome_Dbxref) %>%
  filter(!is.na(genome_Dbxref)) %>%
  sample_n(10)


```

```{r}
library(dplyr)
library(stringr)

# Function to check BLAST hit quality across multiple hits in a string
check_blast_hit <- function(hit_string, min_identity = 30, max_evalue = 1e-5) {
  if (is.na(hit_string)) return(FALSE)
  
  # Split into individual hits (semicolon-delimited)
  hits <- unlist(strsplit(hit_string, ";"))
  
  for (h in hits) {
    perc_identity <- str_extract(h, "\\d+\\.?\\d*%ID")
    evalue <- str_extract(h, "E:[^\\^]+")
    
    perc_identity <- as.numeric(str_remove(perc_identity, "%ID"))
    evalue <- as.numeric(str_remove(evalue, "E:"))
    
    good_identity <- !is.na(perc_identity) && perc_identity >= min_identity
    good_evalue   <- !is.na(evalue) && evalue <= max_evalue
    not_hypo      <- !grepl("hypothetical|uncharacterized", h, ignore.case = TRUE)
    
    if (good_identity && good_evalue && not_hypo) {
      return(TRUE)
    }
  }
  return(FALSE)
}

# Apply to BLASTX and BLASTP fields
blast_test <- integrated_table %>%
  mutate(
    blastx_strong = sapply(trinotate_BLASTX, check_blast_hit),
    blastp_strong = sapply(trinotate_BLASTP, check_blast_hit)
  ) %>%
  select(gene_id, trinotate_BLASTX, blastx_strong, trinotate_BLASTP, blastp_strong)

# Inspect first 20 results
head(blast_test, 20)
```
## Possible Classification function
```{r}
# Function to check BLAST hit quality across multiple hits in a string
check_blast_hit <- function(hit_string, min_identity = 30, max_evalue = 1e-5) {
  if (is.na(hit_string)) return(FALSE)
  
  # Split into individual hits (semicolon-delimited)
  hits <- unlist(strsplit(hit_string, ";"))
  
  for (h in hits) {
    perc_identity <- stringr::str_extract(h, "\\d+\\.?\\d*%ID")
    evalue <- stringr::str_extract(h, "E:[^\\^]+")
    
    perc_identity <- as.numeric(stringr::str_remove(perc_identity, "%ID"))
    evalue <- as.numeric(stringr::str_remove(evalue, "E:"))
    
    good_identity <- !is.na(perc_identity) && perc_identity >= min_identity
    good_evalue   <- !is.na(evalue) && evalue <= max_evalue
    not_hypo      <- !grepl("hypothetical|uncharacterized", h, ignore.case = TRUE)
    
    if (good_identity && good_evalue && not_hypo) {
      return(TRUE)
    }
  }
  return(FALSE)
}

# Main classification function
classify_annotations <- function(integrated_table) {
  
  coverage_threshold_strong <- 70
  coverage_threshold_moderate <- 40
  
  integrated_table %>%
    dplyr::mutate(
      # Transcriptome strong evidence
      has_eggnog = if_any(c("egg_ortholog","egg_desc","egg_COG","egg_KO","egg_KEGGmap","egg_GO"),
                          ~ !is.na(.) & . != ""),
      has_ips_go = !is.na(ips_GO) & ips_GO != "",
      has_trinotate_go = !is.na(trinotate_GO) & trinotate_GO != "",
      blastx_strong = sapply(trinotate_BLASTX, check_blast_hit),
      blastp_strong = sapply(trinotate_BLASTP, check_blast_hit),
      has_blast = blastx_strong | blastp_strong,
      
      n_transcriptome_strong = (has_eggnog + has_ips_go + has_trinotate_go + has_blast),
      
      # Genome evidence
      has_genome_go      = !is.na(genome_GO) & genome_GO != "",
      has_genome_uniprot = !is.na(genome_UniProt) & genome_UniProt != "",
      has_genome_dbxref  = !is.na(genome_Dbxref) & genome_Dbxref != "",
      
      genome_strong      = (has_genome_go | has_genome_uniprot) & genome_coverage >= coverage_threshold_strong,
      genome_conditional = (has_genome_go | has_genome_uniprot) & genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong,
      genome_weak        = (has_genome_go | has_genome_uniprot) & genome_coverage < coverage_threshold_moderate,
      
      # Weak evidence (coalesced to avoid NA)
      has_signalp = trinotate_SignalP == "YES",
      has_tmhmm   = trinotate_TMHMM == "YES",
      has_pfam_ipr = if_any(c("egg_PFAM","ips_IPR","ips_PFAM","trinotate_PFAM"), ~ !is.na(.) & . != ""),
      
      weak_sources = dplyr::coalesce(
        (has_signalp | has_tmhmm | has_pfam_ipr | has_genome_dbxref | genome_weak),
        FALSE
      ),
      
      # Total strong sources depending on coverage
      strong_sources = dplyr::case_when(
        genome_coverage >= coverage_threshold_strong ~ n_transcriptome_strong + genome_strong,
        genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong ~ n_transcriptome_strong + genome_conditional,
        genome_coverage < coverage_threshold_moderate ~ n_transcriptome_strong,
        genome_coverage == 0 ~ n_transcriptome_strong
      ),
      
      # Classification based on full case matrix
      annotation_class = dplyr::case_when(
        # High coverage
        genome_coverage >= coverage_threshold_strong & strong_sources >= 2 ~ "Well-annotated",
        genome_coverage >= coverage_threshold_strong & strong_sources == 1 ~ "Partially annotated",
        genome_coverage >= coverage_threshold_strong & strong_sources == 0 & weak_sources ~ "Weakly annotated",
        genome_coverage >= coverage_threshold_strong & strong_sources == 0 & !weak_sources ~ "Mapped only",
        
        # Moderate coverage
        genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong & n_transcriptome_strong >= 2 ~ "Well-annotated",
        genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong & (n_transcriptome_strong == 1 | genome_conditional) ~ "Partially annotated",
        genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong & n_transcriptome_strong == 0 & !genome_conditional & weak_sources ~ "Weakly annotated",
        genome_coverage >= coverage_threshold_moderate & genome_coverage < coverage_threshold_strong & n_transcriptome_strong == 0 & !genome_conditional & !weak_sources ~ "Mapped only",
        
        # Low coverage
        genome_coverage > 0 & genome_coverage < coverage_threshold_moderate & n_transcriptome_strong >= 2 ~ "Well-annotated",
        genome_coverage > 0 & genome_coverage < coverage_threshold_moderate & n_transcriptome_strong == 1 ~ "Partially annotated",
        genome_coverage > 0 & genome_coverage < coverage_threshold_moderate & n_transcriptome_strong == 0 & weak_sources ~ "Weakly annotated",
        genome_coverage > 0 & genome_coverage < coverage_threshold_moderate & n_transcriptome_strong == 0 & !weak_sources ~ "Mapped only",
        
        # No coverage
        genome_coverage == 0 & n_transcriptome_strong >= 2 ~ "Well-annotated",
        genome_coverage == 0 & n_transcriptome_strong == 1 ~ "Partially annotated",
        genome_coverage == 0 & n_transcriptome_strong == 0 & weak_sources ~ "Weakly annotated",
        genome_coverage == 0 & n_transcriptome_strong == 0 & !weak_sources ~ "Unannotated",
        
        # Catch-all safety net
        TRUE ~ "Unannotated"
      )
    )
}
```

```{r}
colnames(integrated_table)
nrow(integrated_table)
```


## Classify and summarise all annotations
```{r}
# Classify
classified_genes <- classify_annotations(integrated_table)

# Summaries
classified_genes %>%
  count(annotation_class) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

## Save integrated and classified table
write_csv(classified_genes,
          file = "results/annotation/annotations.csv")

```

## Genome-based Annotation Summary
```{r}
# --- Load GFF ---
gff <- read_tsv("resources/genome_based_annotations/fSymMel2.gff.gz",
                comment = "##", col_names = FALSE)

colnames(gff) <- c("seqid","source","type","start","end",
                   "score","strand","phase","attributes")

# --- Genome totals ---
genes <- gff %>% filter(type == "gene") %>%
  mutate(gene_id = str_extract(attributes, "ID=[^;]+"),
         gene_id = str_remove(gene_id, "^ID="))

genome_total <- n_distinct(genes$gene_id)

# --- mRNA features ---
mrna <- gff %>% filter(type == "mRNA") %>%
  mutate(gene_id = str_extract(attributes, "Parent=[^;]+"),
         gene_id = str_remove(gene_id, "^Parent=")) %>%
  mutate(
    has_GO       = str_detect(attributes, "Ontology_term="),
    has_UniProt  = str_detect(attributes, "uniprot_id="),
    has_Dbxref   = str_detect(attributes, "InterPro|Pfam|Dbxref="),
    product_hypo = str_detect(attributes, "product=hypothetical protein"),

    # Adjusted definitions
    curated_function   = has_UniProt,                          # curated = UniProt
    annotated_function = has_GO | has_Dbxref | (!product_hypo),# GO + domain/db evidence
    unannotated        = product_hypo & !has_UniProt & !has_GO & !has_Dbxref
  )

# Collapse mRNA to unique gene-level flags
mrna_flags <- mrna %>%
  group_by(gene_id) %>%
  summarise(
    curated_function   = any(curated_function),
    annotated_function = any(annotated_function),
    unannotated        = any(unannotated),
    .groups = "drop"
  )

# --- Genome-level summary ---
genome_curated_total <- sum(mrna_flags$curated_function)
genome_annotated_total <- sum(mrna_flags$annotated_function)
genome_unannotated_total <- sum(mrna_flags$unannotated)

cat("Genome annotation (GFF) total genes:", genome_total, "\n")
cat("Curated (UniProt):", genome_curated_total,
    "(", round(genome_curated_total/genome_total*100), "%)\n")
cat("Annotated (GO/InterPro/Pfam/etc.):", genome_annotated_total,
    "(", round(genome_annotated_total/genome_total*100), "%)\n")
cat("Unannotated:", genome_unannotated_total,
    "(", round(genome_unannotated_total/genome_total*100), "%)\n\n")

# --- Transcriptome–genome intersection ---
genome_based_flagged <- genome_based %>%
  filter(!is.na(Gene_ID)) %>%
  left_join(mrna_flags, by = c("Gene_ID" = "gene_id"))

genome_based_unique <- genome_based_flagged %>%
  group_by(Gene_ID) %>%
  summarise(
    Coverage = max(`Coverage_%`),
    curated_function   = any(curated_function, na.rm = TRUE),
    annotated_function = any(annotated_function, na.rm = TRUE),
    unannotated        = any(unannotated, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    cov_class = case_when(
      Coverage < 40 ~ "<40%",
      Coverage >= 40 & Coverage < 70 ~ "40-70%",
      Coverage >= 70 ~ ">=70%"
    )
  )

coverage_summary <- genome_based_unique %>%
  group_by(cov_class) %>%
  summarise(
    total_genes       = n(),
    curated           = sum(curated_function),
    annotated         = sum(annotated_function),
    unannotated       = sum(unannotated),
    .groups = "drop"
  )

coverage_totals <- coverage_summary %>%
  summarise(
    cov_class = "Total",
    total_genes = sum(total_genes),
    curated     = sum(curated),
    annotated   = sum(annotated),
    unannotated = sum(unannotated)
  )

coverage_summary <- bind_rows(coverage_summary, coverage_totals) %>%
  mutate(cov_class = factor(cov_class,
                            levels = c("<40%", "40-70%", ">=70%", "Total"))) %>%
  arrange(cov_class)

print(coverage_summary)

# --- Intersection reporting ---
intersection_total <- coverage_summary$total_genes[coverage_summary$cov_class=="Total"]
mod_high <- coverage_summary %>% filter(cov_class %in% c("40-70%", ">=70%"))
mod_high_total <- sum(mod_high$total_genes)
mod_high_curated <- sum(mod_high$curated)
mod_high_annotated <- sum(mod_high$annotated)
mod_high_unannotated <- sum(mod_high$unannotated)

cat("\nTranscriptome–genome intersection total genes:", intersection_total, "\n")
cat("Low coverage (<40%):", coverage_summary$total_genes[coverage_summary$cov_class=="<40%"],
    "genes (", round(coverage_summary$total_genes[coverage_summary$cov_class=="<40%"]/intersection_total*100), "%)\n")
cat("Moderate coverage (40–70%):", coverage_summary$total_genes[coverage_summary$cov_class=="40-70%"],
    "genes (", round(coverage_summary$total_genes[coverage_summary$cov_class=="40-70%"]/intersection_total*100), "%)\n")
cat("High coverage (≥70%):", coverage_summary$total_genes[coverage_summary$cov_class==">=70%"],
    "genes (", round(coverage_summary$total_genes[coverage_summary$cov_class==">=70%"]/intersection_total*100), "%)\n")
cat("Curated (UniProt):", coverage_summary$curated[coverage_summary$cov_class=="Total"],
    "(", round(coverage_summary$curated[coverage_summary$cov_class=="Total"]/intersection_total*100), "%)\n")
cat("Annotated (GO/InterPro/Pfam/etc.):",
    coverage_summary$annotated[coverage_summary$cov_class=="Total"],
    "(", round(coverage_summary$annotated[coverage_summary$cov_class=="Total"]/intersection_total*100), "%)\n")
cat("Unannotated:",
    coverage_summary$unannotated[coverage_summary$cov_class=="Total"],
    "(", round(coverage_summary$unannotated[coverage_summary$cov_class=="Total"]/intersection_total*100), "%)\n")
cat("Moderate to high coverage (≥40%):", mod_high_total, "genes (",
    round(mod_high_total/intersection_total*100), "%), of which",
    mod_high_curated, "are curated (",
    round(mod_high_curated/mod_high_total*100), "%),",
    mod_high_annotated, "are annotated (",
    round(mod_high_annotated/mod_high_total*100), "%), and",
    mod_high_unannotated, "remain unannotated (",
    round(mod_high_unannotated/mod_high_total*100, 1), "%)\n")
```
Of the 19,172 curated genes annotated in the genome, 18,641 (97%) were retained in the transcriptome–genome intersection, while 531 (3%) curated genes were absent. These missing curated genes likely represent loci not expressed in the sampled transcriptome or annotation artifacts from genome gene models. Thus, while the intersection enriches for annotations, it also excludes a subset of genome‑annotated genes.

-----------

## Genome‑level and Trinity gene‑level flags
```{r}
# Genome-level flags (unique genome_gene_id)
genome_flags <- integrated_table %>%
  filter(!is.na(genome_gene_id)) %>%
  group_by(genome_gene_id) %>%
  summarise(
    has_genome = any(
      genome_coverage >= 40 &
      ( (!is.na(genome_GO) & genome_GO != "") |
        (!is.na(genome_UniProt) & genome_UniProt != "") )
    ),
    .groups = "drop"
  )

# Trinity gene-level flags
gene_flags <- integrated_table %>%
  mutate(
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != "")
  ) %>%
  select(gene_id, genome_gene_id, has_eggnog, has_ips, has_trinotate) %>%
  left_join(genome_flags, by = "genome_gene_id") %>%
  mutate(has_genome = replace_na(has_genome, FALSE))
```

## UpSet plots and summary counts
```{r}
# Trinity gene-level UpSet
gene_matrix <- gene_flags %>%
  mutate(across(-gene_id, ~ as.integer(.))) %>%
  as.data.frame()
rownames(gene_matrix) <- gene_matrix$gene_id
gene_matrix <- gene_matrix[,-1]

upset(gene_matrix,
      sets = c("has_eggnog","has_ips","has_trinotate","has_genome"),
      mainbar.y.label = "Trinity gene-level intersections",
      sets.x.label = "Trinity gene counts")

# Genome-level UpSet
genome_matrix <- gene_flags %>%
  filter(!is.na(genome_gene_id)) %>%
  group_by(genome_gene_id) %>%
  summarise(
    has_eggnog    = any(has_eggnog),
    has_ips       = any(has_ips),
    has_trinotate = any(has_trinotate),
    has_genome    = any(has_genome),
    .groups = "drop"
  ) %>%
  mutate(across(-genome_gene_id, ~ as.integer(.))) %>%
  as.data.frame()
rownames(genome_matrix) <- genome_matrix$genome_gene_id
genome_matrix <- genome_matrix[,-1]

upset(genome_matrix,
      sets = c("has_eggnog","has_ips","has_trinotate","has_genome"),
      mainbar.y.label = "Genome-level intersections",
      sets.x.label = "Unique genome loci")

# Summary counts
source_counts_gene <- gene_flags %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate)
  )
source_counts_genome <- genome_flags %>%
  summarise(genome_total = sum(has_genome))

summary_counts <- bind_cols(source_counts_gene, source_counts_genome)
print(summary_counts)
```

## Coverage‑stratified summary table
```{r}
## Coverage‑stratified summary table
coverage_data <- integrated_table %>%
  mutate(
    coverage_category = case_when(
      is.na(genome_gene_id) ~ "Free Trinity genes (0%)",
      genome_coverage < 40 ~ "Low (<40%)",
      genome_coverage >= 40 & genome_coverage < 70 ~ "Moderate (40–70%)",
      genome_coverage >= 70 ~ "High (≥70%)",
      TRUE ~ "Unclassified"
    ),
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != ""),
    # genome functional annotation flag (GO or UniProt present)
    has_genome = (!is.na(genome_GO) & genome_GO != "") |
                 (!is.na(genome_UniProt) & genome_UniProt != ""),
    # functional annotation flag (any source, including genome functional)
    has_functional = has_trinotate | has_eggnog | has_ips | has_genome
  )

final_summary <- coverage_data %>%
  group_by(coverage_category) %>%
  summarise(
    count            = n(),
    has_genome       = sum(has_genome),
    has_trinotate    = sum(has_trinotate),
    has_eggnog       = sum(has_eggnog),
    has_ips          = sum(has_ips),
    functional_total = sum(has_functional),
    .groups = "drop"
  ) %>%
  mutate(
    coverage_category = factor(
      coverage_category,
      levels = c("High (≥70%)", "Moderate (40–70%)", "Low (<40%)", "Free Trinity genes (0%)")
    )
  ) %>%
  arrange(coverage_category)

print(final_summary)
```
We stratified transcriptome genes, assembled with Trinity, according to the maximum coverage achieved by any of their transcripts when aligned to genome‑annotated mRNAs. Most genes were highly covered (≥70%), with 182,826 in this class; 177,192 of them (≈97%) carried at least one functional annotation. Genes with moderate coverage (40–70%) numbered 4,844, of which 4,415 (≈91%) were functionally annotated. The low‑coverage class (<40%) contained 6,888 genes, with 5,601 (≈81%) functionally annotated. The smaller proportion of annotated genes in this group suggests that many of the unannotated ones may represent assembly artifacts. Finally, 63,152 free transcriptome genes had no match to genome‑annotated mRNAs, yet 2,767 (≈4%) still carried functional annotation. These unmatched but annotated genes represent novel content uncovered by the transcriptome assembly, expanding the gene catalog beyond what is captured in the current genome annotation.

-----------


```{r}
# BED file: transcriptome alignments intersected with genome GFF
bed_data <- read_tsv("data/intersected_gff_with_transcriptomebam.bed.gz",
                     col_names = FALSE)

# Inspect first few rows
head(bed_data)

```



```{r}
library(dplyr)
library(tidyr)
library(UpSetR)

# Step 1: Genome-level flag (unique genome_gene_id)
genome_flags <- integrated_table %>%
  filter(!is.na(genome_gene_id)) %>%
  group_by(genome_gene_id) %>%
  summarise(
    has_genome = any(
      genome_coverage >= 70 &
      ( (!is.na(genome_GO) & genome_GO != "") |
        (!is.na(genome_UniProt) & genome_UniProt != "") )
    ),
    .groups = "drop"
  )

# Step 2: Transcript-level flags
overlap_data <- integrated_table %>%
  mutate(
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != "")
  ) %>%
  select(gene_id, genome_gene_id, has_eggnog, has_ips, has_trinotate)

# Step 3: Join genome flag back to transcript table
overlap_data <- overlap_data %>%
  left_join(genome_flags, by = "genome_gene_id") %>%
  mutate(has_genome = replace_na(has_genome, FALSE))

# -------------------------------
# Transcript-level UpSet plot
# -------------------------------
overlap_matrix_transcript <- overlap_data %>%
  mutate(across(-gene_id, ~ as.integer(.))) %>%
  as.data.frame()

rownames(overlap_matrix_transcript) <- overlap_matrix_transcript$gene_id
overlap_matrix_transcript <- overlap_matrix_transcript[,-1]

upset(overlap_matrix_transcript,
      sets = c("has_eggnog","has_ips","has_trinotate","has_genome"),
      mainbar.y.label = "Transcriptome-level intersections",
      sets.x.label = "Transcriptome genes")

# -------------------------------
# Genome-level UpSet plot
# -------------------------------
overlap_data_genome <- overlap_data %>%
  filter(!is.na(genome_gene_id)) %>%
  group_by(genome_gene_id) %>%
  summarise(
    has_eggnog    = any(has_eggnog),
    has_ips       = any(has_ips),
    has_trinotate = any(has_trinotate),
    has_genome    = any(has_genome),
    .groups = "drop"
  )

overlap_matrix_genome <- overlap_data_genome %>%
  mutate(across(-genome_gene_id, ~ as.integer(.))) %>%
  as.data.frame()

rownames(overlap_matrix_genome) <- overlap_matrix_genome$genome_gene_id
overlap_matrix_genome <- overlap_matrix_genome[,-1]

upset(overlap_matrix_genome,
      sets = c("has_eggnog","has_ips","has_trinotate","has_genome"),
      mainbar.y.label = "Genome-level intersections",
      sets.x.label = "Unique genome loci")

# -------------------------------
# Summary counts
# -------------------------------
# Transcript-level totals
source_counts_transcript <- overlap_data %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate)
  )

# Genome-level totals (unique loci)
source_counts_genome <- genome_flags %>%
  summarise(genome_total = sum(has_genome))

# Combined summary
summary_counts <- bind_cols(source_counts_transcript, source_counts_genome)
print(summary_counts)
```

```{r}
# Inspect genome loci that failed the has_genome filter
failed_genome <- genome_flags %>%
  filter(!has_genome) %>%
  left_join(
    integrated_table %>%
      select(genome_gene_id, genome_coverage, genome_GO, genome_UniProt),
    by = "genome_gene_id"
  )

# How many loci failed?
nrow(failed_genome)

# Look at a few examples
head(failed_genome, 20)

# How many Trinity genes do NOT have a genome_gene_id?
unmapped_trinity <- integrated_table %>%
  filter(is.na(genome_gene_id)) %>%
  nrow()
unmapped_trinity

unmapped_genes <- integrated_table %>%
  filter(is.na(genome_gene_id)) %>%
  mutate(
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != "")
  )

# How many unmapped Trinity genes still have annotation?
summary_unmapped <- unmapped_genes %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate)
  )

nrow(unmapped_genes)        # total unmapped Trinity genes
summary_unmapped            # annotation counts among them

# Inspect unmapped Trinity genes with annotation
unmapped_with_annotation <- unmapped_genes %>%
  filter(has_eggnog | has_ips | has_trinotate)

nrow(unmapped_with_annotation)   # should be ~3000
head(unmapped_with_annotation, 20)
```



```{r}
# Step 1: Define overlap flags (bundled scheme, consistent with classification)
overlap_data <- integrated_table %>%
  mutate(
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != ""),
    # Genome functional fields: only GO or UniProt, with 100% coverage
    has_genome = (genome_coverage == 100) &
                 ( (!is.na(genome_GO) & genome_GO != "") |
                   (!is.na(genome_UniProt) & genome_UniProt != "") )
  ) %>%
  select(gene_id, has_eggnog, has_ips, has_trinotate, has_genome)

# Step 2: Convert logicals to numeric (0/1) for UpSetR
overlap_matrix <- overlap_data %>%
  mutate(across(-gene_id, ~ as.integer(.))) %>%
  as.data.frame()

rownames(overlap_matrix) <- overlap_matrix$gene_id
overlap_matrix <- overlap_matrix[,-1]

# Step 3: Global UpSet plot
upset(overlap_matrix, sets = c("has_eggnog","has_ips","has_trinotate","has_genome"))

# Step 4: Extract overlap counts (including unannotated explicitly)
overlap_counts <- overlap_matrix %>%
  mutate(combo = paste0("E", has_eggnog,
                        "_I", has_ips,
                        "_T", has_trinotate,
                        "_G", has_genome)) %>%
  count(combo) %>%
  arrange(desc(n))

# Explicit unannotated count
unannotated <- sum(overlap_matrix$has_eggnog == 0 &
                   overlap_matrix$has_ips == 0 &
                   overlap_matrix$has_trinotate == 0 &
                   overlap_matrix$has_genome == 0)

overlap_counts <- bind_rows(
  overlap_counts,
  data.frame(combo = "E0_I0_T0_G0", n = unannotated)
)

print(overlap_counts)

# Step 5: Raw totals per source
source_counts <- overlap_data %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate),
    genome_total    = sum(has_genome)
  )

print(source_counts)

```

```{r}

```


The overlap analysis reveals several distinct categories of gene clusters. The largest group, comprising over 139,000 genes, are supported by genome annotation with functional evidence (GO or UniProt) but not recovered by EggNOG, InterProScan, or Trinotate. This highlights the breadth of genome‑based annotation, which often identifies many functionally tagged genes that homology and domain pipelines do not capture. The second largest group, around 73,000 genes, remain completely unannotated. Such a fraction is typical in de novo transcriptome assemblies and may represent non‑coding RNAs, assembly artifacts, or genuinely novel genes absent from current databases.
At the other end of the spectrum, approximately 16,000 genes are supported by all annotation sources. This intersection defines the most robust evidence set and provides the highest confidence for downstream biological interpretation. A further 13,000 genes are annotated by all functional pipelines but lack genome support, underscoring the complementary role of homology and domain searches in recovering genes missed by structural annotation. Smaller subsets, ranging from a few hundred to several thousand, reflect unique contributions of individual pipelines, such as Trinotate‑only annotations or EggNOG/InterProScan overlaps.

```{r}
# Step 1: Define overlap flags (EggNOG, IPS, Trinotate at transcript-level)
overlap_data <- integrated_table %>%
  mutate(
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != ""),
    has_genome = (genome_coverage > 70) &
                 ( (!is.na(genome_GO) & genome_GO != "") |
                   (!is.na(genome_UniProt) & genome_UniProt != "") )
  ) %>%
  select(gene_id, genome_gene_id, has_eggnog, has_ips, has_trinotate, has_genome)

# Step 2: Collapsed genome totals (unique genome_gene_id)
genome_total <- overlap_data %>%
  filter(has_genome) %>%
  distinct(genome_gene_id) %>%
  nrow()

# Step 3: Totals for other sources (still transcript-level)
source_counts <- overlap_data %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate)
  )

# Step 4: Explicit unannotated count
unannotated <- overlap_data %>%
  filter(!has_eggnog & !has_ips & !has_trinotate & !has_genome) %>%
  nrow()

# Step 5: Combine into tidy dataframe
totals <- data.frame(
  Source = c("EggNOG","IPS","Trinotate","Genome","Unannotated"),
  Count  = c(source_counts$eggnog_total,
             source_counts$ips_total,
             source_counts$trinotate_total,
             genome_total,
             unannotated)
)

print(totals)

# Step 6: Barplot
ggplot(totals, aes(x=Source, y=Count, fill=Source)) +
  geom_bar(stat="identity") +
  labs(title="Annotated vs Unannotated Genes",
       x="Annotation Source",
       y="Number of Genes") +
  theme_minimal(base_size = 14) +
  theme(legend.position="none")
```

```{r}
# Step 1: Create totals dataframe with corrected Genome count
totals <- data.frame(
  Source = c("EggNOG","IPS","Trinotate","Genome","Unannotated"),
  Count  = c(34918, 36946, 37870, 20645, 70846)  # <-- Genome replaced with 20,645
)

# Step 2: Barplot
library(ggplot2)

ggplot(totals, aes(x=Source, y=Count, fill=Source)) +
  geom_bar(stat="identity") +
  labs(title="Annotated vs Unannotated Genes",
       x="Annotation Source",
       y="Number of Genes") +
  theme_minimal(base_size = 14) +
  theme(legend.position="none")
```


```{r}
# Step 1: Define overlap flags (EggNOG, IPS, Trinotate at transcript-level)
overlap_data <- integrated_table %>%
  mutate(
    has_eggnog = if_any(starts_with("egg"), ~ !is.na(.) & . != ""),
    has_ips = if_any(starts_with("ips"), ~ !is.na(.) & . != ""),
    has_trinotate = if_any(c("trinotate_BLASTX","trinotate_BLASTP",
                             "trinotate_PFAM","trinotate_GO"),
                           ~ !is.na(.) & . != ""),
    has_genome = (genome_coverage == 100) &
                 ( (!is.na(genome_GO) & genome_GO != "") |
                   (!is.na(genome_UniProt) & genome_UniProt != "") )
  ) %>%
  select(gene_id, genome_gene_id, has_eggnog, has_ips, has_trinotate, has_genome)

# Step 2: Collapsed genome totals (unique genome_gene_id)
genome_total <- overlap_data %>%
  filter(has_genome) %>%
  distinct(genome_gene_id) %>%
  nrow()

# Step 3: Totals for other sources (still transcript-level)
source_counts <- overlap_data %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate)
  )

# Step 4: Explicit unannotated count
unannotated <- overlap_data %>%
  filter(!has_eggnog & !has_ips & !has_trinotate & !has_genome) %>%
  nrow()

# Step 5: Combine into tidy dataframe
totals <- data.frame(
  Source = c("EggNOG","IPS","Trinotate","Genome","Unannotated"),
  Count  = c(source_counts$eggnog_total,
             source_counts$ips_total,
             source_counts$trinotate_total,
             genome_total,
             unannotated)
)

print(totals)

# Step 6: Barplot
ggplot(totals, aes(x=Source, y=Count, fill=Source)) +
  geom_bar(stat="identity") +
  labs(title="Annotated vs Unannotated Genes",
       x="Annotation Source",
       y="Number of Genes") +
  theme_minimal(base_size = 14) +
  theme(legend.position="none")
```
```{r}
classified_genes %>% distinct(genome_gene_id) %>% filter(!is.na(genome_gene_id)) %>% nrow()
genome_based %>% distinct(Gene_ID) %>% nrow()
```


```{r}
# Totals per source
source_counts <- overlap_data %>%
  summarise(
    eggnog_total    = sum(has_eggnog),
    ips_total       = sum(has_ips),
    trinotate_total = sum(has_trinotate),
    genome_total    = sum(has_genome)
  )

# Explicit unannotated count
unannotated <- sum(overlap_matrix$has_eggnog == 0 &
                   overlap_matrix$has_ips == 0 &
                   overlap_matrix$has_trinotate == 0 &
                   overlap_matrix$has_genome == 0)

# Combine into tidy dataframe
bar_data <- data.frame(
  Source = c("EggNOG","IPS","Trinotate","Genome","Unannotated"),
  Count  = c(source_counts$eggnog_total,
             source_counts$ips_total,
             source_counts$trinotate_total,
             source_counts$genome_total,
             unannotated)
)

# Barplot
ggplot(bar_data, aes(x=Source, y=Count, fill=Source)) +
  geom_bar(stat="identity") +
  labs(title="Annotated vs Unannotated Genes",
       x="Annotation Source",
       y="Number of Genes") +
  theme_minimal(base_size = 14) +
  theme(legend.position="none")
```
```{r}
# Count unique genome genes flagged as TRUE
genome_true_total <- genome_flags %>%
  filter(has_genome) %>%
  summarise(n = n_distinct(genome_gene_id)) %>%
  pull(n)

genome_true_total
```

-------------

## Tier annotations 

## Load all tier gene sets
```{r}
# Tier 1: shared temperature-responsive genes
tier1_genes <- read_csv(
  "results/functional_enrichment/tier1_shared_temperature_response/tier1_shared_temperature_genes.csv"
)$gene_id

# Tier 2: population divergence
tier2_genes <- read_csv(
  "results/functional_enrichment/tier2_pop_divergence/tier2_pop_divergence_genes.csv"
)$gene_id

# Tier 2b: locally adapted thermal genes
tier2b_genes <- read_csv(
  "results/functional_enrichment/tier2_pop_divergence/tier2b_local_adaptation/tier2_local_adaptation_genes.csv"
)$gene_id

# Tier 3: hybrid inheritance categories
tier3_mis_genes <- read_csv(
  "results/functional_enrichment/tier3_hybrid_inheritance/tier3_misexpressed_genes.csv"
)$gene_id

tier3_parent_S_genes <- read_csv(
  "results/functional_enrichment/tier3_hybrid_inheritance/tier3_parent_like_S_genes.csv"
)$gene_id

tier3_parent_W_genes <- read_csv(
  "results/functional_enrichment/tier3_hybrid_inheritance/tier3_parent_like_W_genes.csv"
)$gene_id

# Intersections
tier1_tier3_mis <- read_csv("results/functional_enrichment/tier_intersections/tier1_tier3_misexpressed.csv")$gene_id
tier1_tier3_parent_S <- read_csv("results/functional_enrichment/tier_intersections/tier1_tier3_parent_like_S.csv")$gene_id
tier1_tier3_parent_W <- read_csv("results/functional_enrichment/tier_intersections/tier1_tier3_parent_like_W.csv")$gene_id

tier2b_tier3_mis <- read_csv("results/functional_enrichment/tier_intersections/tier2b_tier3_misexpressed.csv")$gene_id
tier2b_tier3_parent_S <- read_csv("results/functional_enrichment/tier_intersections/tier2b_tier3_parent_like_S.csv")$gene_id
tier2b_tier3_parent_W <- read_csv("results/functional_enrichment/tier_intersections/tier2b_tier3_parent_like_W.csv")$gene_id
```


```{r}
# Count number of genes per tier and intersection
tier_sizes <- tibble(
  tier = c("Tier1","Tier2","Tier2b",
           "Tier3_mis","Tier3_parent_S","Tier3_parent_W",
           "Tier1∩Tier3_mis","Tier1∩Tier3_parent_S","Tier1∩Tier3_parent_W",
           "Tier2b∩Tier3_mis","Tier2b∩Tier3_parent_S","Tier2b∩Tier3_parent_W"),
  n_genes = c(length(tier1_genes),
              length(tier2_genes),
              length(tier2b_genes),
              length(tier3_mis_genes),
              length(tier3_parent_S_genes),
              length(tier3_parent_W_genes),
              length(tier1_tier3_mis),
              length(tier1_tier3_parent_S),
              length(tier1_tier3_parent_W),
              length(tier2b_tier3_mis),
              length(tier2b_tier3_parent_S),
              length(tier2b_tier3_parent_W))
)

print(tier_sizes)
```





```{r}
# Helper function: subset classified_genes by tier/intersection membership
make_tier_table <- function(gene_ids, tier_name, classified_table) {
  classified_table %>%
    filter(gene_id %in% gene_ids) %>%
    mutate(tier = tier_name) %>%
    relocate(tier, .before = gene_id)
}

# --- Build tables for main tiers ---
tier1_table <- make_tier_table(tier1_genes, "Tier1", classified_genes)
tier2_table <- make_tier_table(tier2_genes, "Tier2", classified_genes)
tier2b_table <- make_tier_table(tier2b_genes, "Tier2b", classified_genes)
tier3_mis_table <- make_tier_table(tier3_mis_genes, "Tier3_mis", classified_genes)
tier3_parentS_table <- make_tier_table(tier3_parent_S_genes, "Tier3_parent_S", classified_genes)
tier3_parentW_table <- make_tier_table(tier3_parent_W_genes, "Tier3_parent_W", classified_genes)

# --- Build tables for intersections ---
tier1_tier3_mis_table <- make_tier_table(intersect(tier1_genes, tier3_mis_genes),
                                         "Tier1∩Tier3_mis", classified_genes)
tier1_tier3_parentS_table <- make_tier_table(intersect(tier1_genes, tier3_parent_S_genes),
                                             "Tier1∩Tier3_parent_S", classified_genes)
tier1_tier3_parentW_table <- make_tier_table(intersect(tier1_genes, tier3_parent_W_genes),
                                             "Tier1∩Tier3_parent_W", classified_genes)

tier2b_tier3_mis_table <- make_tier_table(intersect(tier2b_genes, tier3_mis_genes),
                                          "Tier2b∩Tier3_mis", classified_genes)
tier2b_tier3_parentS_table <- make_tier_table(intersect(tier2b_genes, tier3_parent_S_genes),
                                              "Tier2b∩Tier3_parent_S", classified_genes)
tier2b_tier3_parentW_table <- make_tier_table(intersect(tier2b_genes, tier3_parent_W_genes),
                                              "Tier2b∩Tier3_parent_W", classified_genes)

# --- Combine everything into one master table ---
all_tiers_and_intersections <- bind_rows(
  tier1_table, tier2_table, tier2b_table,
  tier3_mis_table, tier3_parentS_table, tier3_parentW_table,
  tier1_tier3_mis_table, tier1_tier3_parentS_table, tier1_tier3_parentW_table,
  tier2b_tier3_mis_table, tier2b_tier3_parentS_table, tier2b_tier3_parentW_table
)

# Main tiers
tier1_table
tier2_table
tier2b_table
tier3_mis_table
tier3_parentS_table
tier3_parentW_table

# Intersections
tier1_tier3_mis_table
tier1_tier3_parentS_table
tier1_tier3_parentW_table
tier2b_tier3_mis_table
tier2b_tier3_parentS_table
tier2b_tier3_parentW_table

```

```{r}
# Count total rows with a non-NA genome_gene_id
total_ids <- tier2b_table %>% filter(!is.na(genome_gene_id)) %>% nrow()

# Count unique non-NA genome_gene_id values
unique_ids <- tier2b_table %>% filter(!is.na(genome_gene_id)) %>% 
  pull(genome_gene_id) %>% n_distinct()

cat("Total rows with genome_gene_id:", total_ids, "\n")
cat("Unique genome_gene_id:", unique_ids, "\n")

# Find duplicated genome_gene_id values
dup_ids <- tier2b_table %>%
  filter(!is.na(genome_gene_id)) %>%
  count(genome_gene_id) %>%
  filter(n > 1)

cat("Duplicated genome_gene_id count:", nrow(dup_ids), "\n")
print(dup_ids)

# Weakly annotated genes (all columns)
weakly <- tier2b_table %>%
  filter(annotation_class == "Weakly annotated")

# Partially annotated genes (all columns)
partial <- tier2b_table %>%
  filter(annotation_class == "Partially annotated")

# Unannotated genes (all columns)
unannotated <- tier2b_table %>%
  filter(annotation_class == "Unannotated")

# Well-annotated genes (all columns)
well <- tier2b_table %>%
  filter(annotation_class == "Well-annotated")

# Inspect them
weakly
partial
unannotated
well
```
TRINITY_GG_416_c2774_g1
```{r}
# See all column names
colnames(tier2b_table)

# See the structure of the table (column names, types)
str(tier2b_table)

# Peek at the first 10 rows
head(tier2b_table, 10)

# Glimpse a compact overview (dplyr)
dplyr::glimpse(tier2b_table)

# Count how many unique genome_gene_id vs trinity_gene_id
tier2b_table %>%
  summarise(
    n_genome_genes  = n_distinct(genome_gene_id),
    n_trinity_genes = n_distinct(gene_id),   # Trinity IDs are in gene_id
    total_rows      = n()
  )

# Check for duplicated genome_gene_id values
tier2b_table %>%
  count(genome_gene_id) %>%
  filter(n > 1) %>%
  arrange(desc(n))
```


## Tier summaries
```{r}
# Helper: summarise annotation classes as proportions
summarise_tier <- function(gene_ids, tier_name, classified_table) {
  classified_table %>%
    filter(gene_id %in% gene_ids) %>%
    count(annotation_class) %>%
    mutate(percent = round(n / sum(n) * 100, 1),
           tier = tier_name)
}

# Collect summaries for all tiers and intersections
all_tiers_summary <- bind_rows(
  summarise_tier(tier1_genes, "Tier1", classified_genes),
  summarise_tier(tier2_genes, "Tier2", classified_genes),
  summarise_tier(tier2b_genes, "Tier2b", classified_genes),
  summarise_tier(tier3_mis_genes, "Tier3_mis", classified_genes),
  summarise_tier(tier3_parent_S_genes, "Tier3_parent_S", classified_genes),
  summarise_tier(tier3_parent_W_genes, "Tier3_parent_W", classified_genes),
  summarise_tier(tier1_tier3_mis, "Tier1∩Tier3_mis", classified_genes),
  summarise_tier(tier1_tier3_parent_S, "Tier1∩Tier3_parent_S", classified_genes),
  summarise_tier(tier1_tier3_parent_W, "Tier1∩Tier3_parent_W", classified_genes),
  summarise_tier(tier2b_tier3_mis, "Tier2b∩Tier3_mis", classified_genes),
  summarise_tier(tier2b_tier3_parent_S, "Tier2b∩Tier3_parent_S", classified_genes),
  summarise_tier(tier2b_tier3_parent_W, "Tier2b∩Tier3_parent_W", classified_genes)
)

all_tiers_summary

# Plot proportions instead of raw counts
ggplot(all_tiers_summary, aes(x=tier, y=percent, fill=annotation_class)) +
  geom_bar(stat="identity", position="stack") +
  labs(title="Proportional annotation class distribution across tiers",
       x="Tier / Intersection",
       y="Percentage of genes") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```
Annotation completeness declined from Tier1 to Tier2 to Tier2b. Tier1 genes were predominantly well‑annotated, reflecting their conserved nature and strong representation in reference databases. Tier2 included a larger and more diverse set, which reduced annotation rates. Although Tier2b only contains 32 genes, it showed the highest proportion of unannotated genes, consistent with the expectation that temperature‑responsive, locally adapted genes may be fast‑evolving and therefore less likely to be captured in current databases. 

In contrast, Tier3 sets (misexpressed and parental) were overwhelmingly well‑annotated (>85–90%), with only ~5% of genes lacking annotation. The similar annotation proportions across Tier3_mis, Tier3_parent_S, and Tier3_parent_W indicate that hybrid misexpression does not preferentially involve novel or unannotated transcripts, but rather affects conserved, well‑characterized genes that are also inherited from the parental populations.

Intersections between tiers largely reflected the annotation properties of the parent sets. Overlaps between Tier1 and Tier3 categories were predominantly well‑annotated (>90%), as expected given that both sets represent conserved functions. In contrast, overlaps involving Tier2b showed more variability. The Tier2b∩Tier3_mis intersection yielded a single, well‑annotated gene: a CD81‑like tetraspanin.

The Tier2b∩Tier3_parent_S intersection contained a single unannotated gene. Although its function remains unresolved, its retention after ORF quality filtering suggests coding potential. 

The larger size of the Tier2b∩Tier3_parent_W intersection compared to the single gene intersections of Tier2b∩Tier3_parent_S and Tier2b∩Tier3_mis indicates that most temperature‑linked adaptive gene expression in hybrids is inherited from the West population, confirming the stronger contribution of West expression to hybrid transcriptomes. 

------

```{r}
colnames(integrated_table)
```
```{r}
# Check for zebrafish IDs in multiple formats
zebrafish_hits <- sum(grepl("7955\\.", integrated_table$egg_ortholog)) +
                  sum(grepl("ENSDARG", integrated_table$egg_ortholog)) +
                  sum(grepl("DANRE", integrated_table$genome_UniProt))

# Check for human IDs in multiple formats
human_hits <- sum(grepl("9606\\.", integrated_table$egg_ortholog)) +
              sum(grepl("ENSP|ENSG", integrated_table$egg_ortholog)) +
              sum(grepl("HUMAN", integrated_table$genome_UniProt))

zebrafish_hits
human_hits
```

```{r}
library(dplyr)

tier_summary <- all_tiers_and_intersections %>%
  mutate(zebrafish_hit = grepl("7955|ENSDARG|DANRE", 
                               paste(egg_ortholog, trinotate_BLASTX, trinotate_BLASTP, genome_UniProt),
                               ignore.case = TRUE),
         human_hit     = grepl("9606|ENSP|ENSG|HUMAN", 
                               paste(egg_ortholog, trinotate_BLASTX, trinotate_BLASTP, genome_UniProt),
                               ignore.case = TRUE),
         any_hit       = zebrafish_hit | human_hit) %>%
  group_by(tier) %>%
  summarise(
    total_genes     = n(),
    zebrafish_hits  = sum(zebrafish_hit, na.rm = TRUE),
    human_hits      = sum(human_hit, na.rm = TRUE),
    annotated       = sum(any_hit, na.rm = TRUE),
    unannotated     = total_genes - annotated
  )

print(tier_summary)
```


```{r}
# --- Compare transcript-level vs collapsed genome counts ---

# Transcript-level genome counts (as currently in overlap_data)
genome_transcript_total <- sum(overlap_data$has_genome)

# Collapsed unique genome gene IDs
unique_genome_ids <- genome_based %>%
  filter(!is.na(Gene_ID) & Gene_ID != "NA" & Gene_ID != "") %>%
  distinct(Gene_ID)

genome_unique_total <- nrow(unique_genome_ids)

# Combine into tidy dataframe
genome_count_comparison <- tibble(
  Level  = c("Transcript-level overlaps", "Unique genome gene IDs"),
  Count  = c(genome_transcript_total, genome_unique_total)
)

print(genome_count_comparison)

# Optional barplot
ggplot(genome_count_comparison, aes(x=Level, y=Count, fill=Level)) +
  geom_bar(stat="identity") +
  labs(title="Genome annotation counts: transcript vs unique gene IDs",
       x="Annotation level",
       y="Number of entries") +
  theme_minimal(base_size = 14) +
  theme(legend.position="none")
```

## Tier2b
TRINITY_GG_416_c2774_g1 represents a Trinity‑assembled gene with strong homology to cyclin‑dependent kinase regulatory subunit 1 (CKS1). BLAST searches against UniProt identify multiple high‑confidence alignments to CKS1_MOUSE (P61025), each exceeding 90% sequence identity with highly significant E‑values, supporting functional conservation. Among its isoforms, TRINITY_GG_416_c2774_g1_i14 achieved a maximum alignment of ~65% coverage to the genome‑annotated mRNA FUNCG00000020675. This indicates that the genome gene/mRNA model is only partially reconstructed relative to the Trinity assembly. Importantly, the genome‑based annotation was informative: InterPro, Pfam, SMART, and ProSite domain assignments (e.g. IPR000789, PF01111, SM01084, PS00944/45) provided a GO term consistent with cyclin‑dependent protein kinase regulator activity (GO:0016538). In this case, the Trinity assembly provides a more complete sequence that reinforces and complements the functional assignment derived from the genome annotation. 
CKS1 proteins are canonical regulators of cyclin‑dependent kinases, controlling cell cycle progression and transcriptional programs. While they are not classical “temperature response” genes like heat shock proteins, their regulatory role means they can indirectly influence how cells cope with stress. Temperature shifts often impose cellular stress that requires adjustments in proliferation, DNA repair, and transcriptional regulation. In this study, TRINITY_GG_416_c2774_g1 was differentially expressed between corkwing wrasse populations exposed to thermal treatments, suggesting that modulation of CDK activity may contribute to population‑specific responses and local adaptation to contrasting temperature regimes. A change in CKS1 expression could therefore reflect altered cell cycle control or stress adaptation pathways under different thermal environments.












