---
title: "Comparison and Integration of Annotation Sources"
author: "Carlota_Gouveia"
date: "2026-02-25"
output:
  html_document:
    toc: true
    toc_float: true
---

# Introduction

This report summarises, compares, and integrates the different annotation sources available for the *Symphodus melops* transcriptome, including EggNOG, InterProScan, Trinotate, and genome-based annotations.

## Load necessary packages
```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(readr)
  library(stringr)
  library(dplyr)
  library(stringr)
  library(purrr)
  library(tidyr)
})
```

## Load and Inspect Annotation Sources
```{r}
## EggNOG annotation
# Read file as raw text
raw <- readLines("results/annotation/eggnog/Trinity.emapper.annotations")
# Convert literal "\t" into real tabs
raw <- gsub("\\\\t", "\t", raw)
# Identify the header line (the one starting with "#query")
header_line <- raw[grepl("^#query", raw)]
# Extract column names from the header
col_names <- strsplit(sub("^#", "", header_line), "\t")[[1]]
# Extract all data lines (those NOT starting with "#")
data_lines <- raw[!grepl("^#", raw)]
# Read the data into a table using the extracted header
eggnog <- read.delim(
  textConnection(data_lines),
  header = FALSE,
  sep = "\t",
  quote = "",
  stringsAsFactors = FALSE
)
# Assign the correct column names
colnames(eggnog) <- col_names


## InterProScan output
ips <- read_tsv(
  "results/annotation/ips/IPS_merged.tsv",
  col_names = FALSE
)

## Trinotate full report
trinotate_report <- read_tsv(
  "results/annotation/trinotate/trinotate_final/trinotate_annotation_report.xls"
)

## SignalP predictions
signalp <- read_tsv(
  "results/annotation/trinotate/signalp/signalp_merged.output.gff3",
  col_names = FALSE,
  comment = "##",
  show_col_types = FALSE
)
## Assign the correct column names
colnames(signalp) <- c(
  "seqid", "source", "type", "start", "end",
  "score", "strand", "phase", "attributes"
)

## Keep only transcript ID (remove ORF annotation)
signalp$seqid <- sub(" .*", "", signalp$seqid)

## DeepTMHMM GFF3 
tmhmm <- read_tsv(
  "results/annotation/trinotate/tmhmm/tmhmm_clean.gff3",
  col_names = FALSE,
  comment = "##",
  show_col_types = FALSE
)
colnames(tmhmm) <- c(
  "seqid", "source", "type", "start", "end",
  "score", "strand", "phase", "attributes"
)

## Genome-based annotation
genome_based <- read_tsv(
  "results/annotation/genome_based/trinities_filteringby_cov_summary.tsv"
)

# ---- Quick row-count inspection ----

cat("EggNOG rows:", nrow(eggnog), "\n")
cat("IPS rows:", nrow(ips), "\n")
cat("Trinotate rows:", nrow(trinotate_report), "\n")
cat("SignalP rows:", nrow(signalp), "\n")
cat("TMHMM GFF3 rows:", nrow(tmhmm_gff3), "\n")
cat("Genome-based rows:", nrow(genome_based), "\n")

# ---- Compact structure preview ----

cat("\nEggNOG (head):\n")
head(eggnog)

cat("\nIPS (head):\n")
head(ips)

cat("\nTrinotate report (head):\n")
head(trinotate_report)

cat("\nSignalP (head):\n")
head(signalp)

cat("\nTMHMM GFF3 (head):\n")
head(tmhmm_gff3)

cat("\nGenome-based annotation (head):\n")
head(genome_based)
```

## Standardize Trinity IDs across annotation sources
```{r extract_gene_ids}

# Helper: extract gene-level ID from any Trinity-style ID
extract_gene_id <- function(x) {
  x %>%
    str_remove("\\.p\\d+$") %>%   # remove .p1, .p2, etc.
    str_remove("_i\\d+$") %>%     # remove isoform _i1, _i2, etc.
    str_replace("^#", "")         # remove leading # if present
}

# ---- EggNOG ----
# first column is protein ID like TRINITY_GG_1000_c0_g1_i1.p1
eggnog <- eggnog %>%
  rename(protein_id = query) %>%
  mutate(
    transcript_id = str_remove(protein_id, "\\.p\\d+$"),
    gene_id       = extract_gene_id(transcript_id)
  )

# ---- IPS ----
ips <- ips %>%
  rename(protein_id = X1) %>%
  mutate(
    transcript_id = str_remove(protein_id, "\\.p\\d+$"),
    gene_id       = extract_gene_id(transcript_id)
  )

# ---- Trinotate ----
trinotate_report <- trinotate_report %>%
  rename(gene_id = `#gene_id`) %>%
  mutate(
    gene_id       = extract_gene_id(gene_id),
    transcript_id = transcript_id
  )

# ---- SignalP ----
signalp <- signalp %>%
  mutate(
    # Remove protein suffix (.p1, .p2, ...)
    transcript_id = sub("\\.p[0-9]+$", "", seqid),

    # Derive gene_id (not used for merging)
    gene_id = sub("_i.*", "", transcript_id)
  )


# ---- TMHMM ----
tmhmm <- tmhmm %>%
  mutate(
    transcript_id = sub("\\.p[0-9]+$", "", seqid)
  )

# ---- Genome-based ----
genome_based <- genome_based %>%
  rename(transcript_id = Trinity_ID) %>%
  mutate(
    gene_id = extract_gene_id(transcript_id)
  )

## Identify REAL genome-based annotation (filtered ≥80% coverage)
# IMPORTANT:
# - We filter genome-based annotation BEFORE computing intersections
#   to ensure only high-confidence genome matches are counted.
# - The transcript table later uses UNFILTERED genome_based so that
#   ALL transcripts remain in the universe (important for Tier analysis).
# - Therefore, genome_real is computed here, not in the "Identify real
#   annotation for each tool" section.

# Apply coverage filter (≥80%) for intersection analysis
genome_filtered <- genome_based %>%
  filter(`Coverage_%` >= 80)

genome_real <- genome_filtered %>%
  mutate(real = (!is.na(GO_Terms) & GO_Terms != "") |
               (!is.na(Dbxref) & Dbxref != "") |
               (!is.na(UniProt_ID) & UniProt_ID != "")) %>%
  select(transcript_id, real) %>%
  distinct(transcript_id, .keep_all = TRUE)
```

```{r}
# 3b. Check ID extraction and consistency across datasets

# Helper function to summarize ID status
check_ids <- function(df, name) {
  cat("\n=== Checking:", name, "===\n")
  
  # 1. How many rows?
  cat("Rows:", nrow(df), "\n")
  
  # 2. Missing transcript IDs?
  cat("Missing transcript_id:", sum(is.na(df$transcript_id) | df$transcript_id == ""), "\n")
  
  # 3. Missing gene IDs?
  cat("Missing gene_id:", sum(is.na(df$gene_id) | df$gene_id == ""), "\n")
  
  # 4. Unique gene IDs
  cat("Unique gene_ids:", n_distinct(df$gene_id), "\n")
  
  # 5. Show a few examples
  cat("Example IDs:\n")
  print(head(df %>% select(transcript_id, gene_id)))
}

# Run checks for each dataset
check_ids(eggnog, "EggNOG")
check_ids(ips, "IPS")
check_ids(trinotate_report, "Trinotate")
check_ids(signalp, "SignalP")
check_ids(tmhmm_gff3, "TMHMM")
check_ids(genome_based, "Genome-based annotation")
```



## Identify real annotation for each tool
```{r}
## Identify REAL annotation for functional tools (EggNOG, IPS, Trinotate, SignalP, TMHMM)
# NOTE:
# genome_real is NOT computed here because it requires filtering by coverage.
# The filtered genome_real was computed earlier, right after cleaning genome_based.

eggnog_real <- eggnog %>%
  mutate(real = apply(select(., -protein_id, -transcript_id, -gene_id), 1,
                      function(x) any(x != "-" & !is.na(x)))) %>%
  group_by(transcript_id) %>%                 # <-- FIX: collapse duplicates
  summarise(real = any(real), .groups = "drop") %>%
  distinct(transcript_id, .keep_all = TRUE)

# IPS — already correct
ips_real <- ips %>%
  mutate(real = (X5 != "-" & !is.na(X5)) |
               (X12 != "-" & !is.na(X12)) |
               (X14 != "-" & !is.na(X14))) %>%
  group_by(transcript_id) %>%
  summarise(real = any(real), .groups = "drop") %>%
  distinct(transcript_id, .keep_all = TRUE)

# Trinotate — already correct
trinotate_real <- trinotate_report %>%
  mutate(real = apply(select(., -transcript_id, -gene_id), 1,
                      function(x) any(x != "." & !is.na(x)))) %>%
  select(transcript_id, real) %>%
  distinct(transcript_id, .keep_all = TRUE)

# SignalP — FIXED: collapse duplicates properly
signalp_real <- signalp %>%
  mutate(real = (`SP(Sec/SPI)` > 0) | !is.na(`CS Position`)) %>%
  group_by(transcript_id) %>%                 # <-- FIX: collapse duplicates
  summarise(real = any(real), .groups = "drop") %>%
  distinct(transcript_id, .keep_all = TRUE)

# TMHMM 
tmhmm_real <- tmhmm_gff3 %>%
  mutate(real = X3 == "TMhelix") %>%
  group_by(transcript_id) %>%
  summarise(real = any(real), .groups = "drop") %>%
  distinct(transcript_id, .keep_all = TRUE)

```



## Create transcript table
```{r}
# Extract transcript_id + gene_id from collapsed tables only
eggnog_t    <- eggnog %>% select(transcript_id, gene_id) %>% distinct(transcript_id, .keep_all = TRUE)
ips_t       <- ips %>% select(transcript_id, gene_id) %>% distinct(transcript_id, .keep_all = TRUE)
trinotate_t <- trinotate_report %>% select(transcript_id, gene_id) %>% distinct(transcript_id, .keep_all = TRUE)
signalp_t   <- signalp %>% select(transcript_id, gene_id) %>% distinct(transcript_id, .keep_all = TRUE)
tmhmm_t     <- tmhmm_gff3 %>% select(transcript_id, gene_id) %>% distinct(transcript_id, .keep_all = TRUE)
genome_t    <- genome_based %>% select(transcript_id, gene_id) %>% distinct(transcript_id, .keep_all = TRUE)

# Combine ONLY transcript IDs
all_transcripts <- bind_rows(
  eggnog_t %>% select(transcript_id),
  ips_t %>% select(transcript_id),
  trinotate_t %>% select(transcript_id),
  signalp_t %>% select(transcript_id),
  tmhmm_t %>% select(transcript_id),
  genome_t %>% select(transcript_id)
) %>% distinct()

# Add gene_id from the most complete sources
transcript_level <- all_transcripts %>%
  left_join(trinotate_t, by = "transcript_id") %>%
  left_join(genome_t, by = "transcript_id", suffix = c("_tri", "_gen")) %>%
  mutate(gene_id = coalesce(gene_id_tri, gene_id_gen)) %>%
  select(transcript_id, gene_id)

transcript_level <- transcript_level %>% arrange(transcript_id)
```

## Add REAL annotation flags
```{r}
transcript_flags <- transcript_level %>%
  left_join(eggnog_real, by="transcript_id") %>% rename(eggnog = real) %>%
  left_join(ips_real, by="transcript_id") %>% rename(ips = real) %>%
  left_join(trinotate_real, by="transcript_id") %>% rename(trinotate = real) %>%
  left_join(signalp_real, by="transcript_id") %>% rename(signalp = real) %>%
  left_join(tmhmm_real, by="transcript_id") %>% rename(tmhmm = real) %>%
  left_join(genome_real, by="transcript_id") %>% rename(genome = real)
```

## Compute intersections
```{r}
intersections <- transcript_flags %>%
  count(eggnog, ips, trinotate, signalp, tmhmm, genome) %>%
  arrange(desc(n))

intersections
```

## Load Tier 1 and Tier 2 gene sets
```{r load_genes_tier_1_2}
# Tier 1: shared temperature-responsive genes
tier1_genes <- read.csv(
  "results/functional_enrichment/tier1_shared_temperature_response/tier1_shared_temperature_genes.csv",
  header = TRUE
)$gene_id

# Tier 2: local adaptation candidates
tier2_genes <- read.csv(
  "results/functional_enrichment/tier2_local_adaptation/tier2_local_adaptation_candidates.csv",
  header = TRUE
)$gene_id

tier1_genes
tier2_genes
```

## Create unfiltered genome-based gene-level flag
```{r}
## Create unfiltered genome-based gene-level flag
genome_gene_flag <- genome_based %>%
  group_by(gene_id) %>%
  summarise(
    genome_unfiltered = any(
      (!is.na(GO_Terms) & GO_Terms != "") |
      (!is.na(Dbxref) & Dbxref != "") |
      (!is.na(UniProt_ID) & UniProt_ID != "")
    )
  )

```


## Generate gene-level annotation table
```{r}
## Generate gene-level annotation table (using UNFILTERED genome annotation)
gene_flags <- transcript_flags %>%
  group_by(gene_id) %>%
  summarise(
    eggnog    = any(eggnog, na.rm = TRUE),
    ips       = any(ips, na.rm = TRUE),
    trinotate = any(trinotate, na.rm = TRUE),
    signalp   = any(signalp, na.rm = TRUE),
    tmhmm     = any(tmhmm, na.rm = TRUE)
  ) %>%
  left_join(genome_gene_flag, by = "gene_id") %>%
  rename(genome = genome_unfiltered) %>%
  mutate(genome = if_else(is.na(genome), FALSE, genome)) %>%   # <-- FIX
  arrange(gene_id)
```

## Intersection with tier 1 and 2
```{r}
tier1_summary <- gene_flags %>% filter(gene_id %in% tier1_genes)
tier2_summary <- gene_flags %>% filter(gene_id %in% tier2_genes)

tier1_overview <- tier1_summary %>%
  summarise(
    n_genes   = n(),
    eggnog    = sum(eggnog,    na.rm = TRUE),
    ips       = sum(ips,       na.rm = TRUE),
    trinotate = sum(trinotate, na.rm = TRUE),
    signalp   = sum(signalp,   na.rm = TRUE),
    tmhmm     = sum(tmhmm,     na.rm = TRUE),
    genome    = sum(genome,    na.rm = TRUE)   # <-- FIX
  )
tier1_overview

tier2_overview <- tier2_summary %>%
  summarise(
    n_genes   = n(),
    eggnog    = sum(eggnog,    na.rm = TRUE),
    ips       = sum(ips,       na.rm = TRUE),
    trinotate = sum(trinotate, na.rm = TRUE),
    signalp   = sum(signalp,   na.rm = TRUE),
    tmhmm     = sum(tmhmm,     na.rm = TRUE),
    genome    = sum(genome,    na.rm = TRUE)   # <-- FIX
  )
tier2_overview
```
## Extract annotations for tiers
```{r extract_func}
extract_full_annotations <- function(gene_list) {
  list(
    eggnog     = eggnog %>% filter(gene_id %in% gene_list),
    ips        = ips %>% filter(gene_id %in% gene_list),
    trinotate  = trinotate_report %>% filter(gene_id %in% gene_list),
    signalp    = signalp %>% filter(gene_id %in% gene_list),
    tmhmm      = tmhmm_gff3 %>% filter(gene_id %in% gene_list),
    genome     = genome_based %>% filter(gene_id %in% gene_list)
  )
}
```

## Extarct for tier 2
```{r}
tier2_annotations_full <- extract_full_annotations(tier2_genes)
lapply(tier2_annotations_full, head)
tier2_annotations_full
```

```{r}
## Function to summarize an annotation table
summarize_annotation_table <- function(df, name) {
  cat("\n==============================\n")
  cat("Summary for:", name, "\n")
  cat("==============================\n")
  
  # Basic structure
  cat("Rows:", nrow(df), "\n")
  cat("Columns:", ncol(df), "\n")
  
  # Transcript and gene counts
  if ("transcript_id" %in% colnames(df)) {
    cat("Unique transcripts:", n_distinct(df$transcript_id), "\n")
  }
  if ("gene_id" %in% colnames(df)) {
    cat("Unique genes:", n_distinct(df$gene_id), "\n")
  }
  
  # Rows per gene (if gene_id exists)
  if ("gene_id" %in% colnames(df)) {
    cat("Rows per gene (summary):\n")
    print(df %>% count(gene_id) %>% pull(n) %>% summary())
  }
  
  # Show first few rows
  cat("\nHead of table:\n")
  print(head(df, 3))
}

## Apply to each Tier 2 annotation table
for (nm in names(tier2_annotations_full)) {
  summarize_annotation_table(tier2_annotations_full[[nm]], nm)
}
```


```{r}
extract_annotations <- function(gene_list,
                                eggnog,
                                ips,
                                trinotate,
                                signalp,
                                tmhmm,
                                genome) {
  
  library(dplyr)
  library(tidyr)
  library(stringr)
  
  # Helper to clean collapsed fields
  clean_field <- function(x) {
    x <- gsub("^;|;$", "", x)      # remove leading/trailing ;
    x <- gsub(";;+", ";", x)       # remove double ;;
    x[x == ""] <- NA               # empty string -> NA
    return(x)
  }
  
  # -------------------------------
  # 1. Genome-based: pick highest-coverage transcript
  # -------------------------------
  genome_best <- genome %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    slice_max(`Coverage_%`, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      genome_transcript_id = transcript_id,
      genome_GO            = GO_Terms,
      genome_UniProt       = UniProt_ID,
      genome_Dbxref        = Dbxref,
      genome_coverage      = `Coverage_%`
    ) %>%
    select(gene_id, genome_transcript_id, genome_GO, genome_UniProt,
           genome_Dbxref, genome_coverage)
  
  # -------------------------------
  # 2. IPS: domains + GO
  # -------------------------------
  ips_gene <- ips %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      ips_IPR  = clean_field(paste0(na.omit(unique(X12[X12 != "-" ])), collapse = ";")),
      ips_PFAM = clean_field(paste0(na.omit(unique(X5 [X5  != "-" ])), collapse = ";")),
      ips_GO   = clean_field(paste0(na.omit(unique(X14[X14 != "-" ])), collapse = ";")),
      .groups = "drop"
    )
  
  # -------------------------------
  # 3. Trinotate: BLAST, PFAM, GO, KEGG
  # -------------------------------
  tri_gene <- trinotate %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      tri_BLAST = clean_field(paste0(na.omit(unique(sprot_Top_BLASTX_hit)), collapse = ";")),
      tri_PFAM  = clean_field(paste0(na.omit(unique(Pfam)), collapse = ";")),
      tri_GO    = clean_field(paste0(na.omit(unique(c(
        gene_ontology_BLASTX,
        gene_ontology_BLASTP,
        gene_ontology_Pfam
      ))), collapse = ";")),
      tri_KEGG  = clean_field(paste0(na.omit(unique(Kegg)), collapse = ";")),
      .groups = "drop"
    )
  
  # -------------------------------
  # 4. EggNOG: ortholog, description, KEGG
  # -------------------------------
  egg_gene <- eggnog %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      egg_ortholog = clean_field(paste0(na.omit(unique(`8010.XP_010868907.1`)), collapse = ";")),
      egg_desc     = clean_field(paste0(na.omit(unique(`Odorant receptor, family E, subfamily 120, member 1`)), collapse = ";")),
      egg_KO       = clean_field(paste0(na.omit(unique(`ko:K04257`)), collapse = ";")),
      egg_KEGGmap  = clean_field(paste0(na.omit(unique(`ko04740,map04740`)), collapse = ";")),
      .groups = "drop"
    )
  
  # -------------------------------
  # 5. SignalP: TRUE if any transcript has SP
  # -------------------------------
  sig_gene <- signalp %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      has_signalp = any(`SP(Sec/SPI)` > 0, na.rm = TRUE),
      .groups = "drop"
    )
  
  # -------------------------------
  # 6. TMHMM: TRUE if any transcript has TM helix
  # -------------------------------
  tmh_gene <- tmhmm %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      has_tmhmm = any(X3 == "TMhelix", na.rm = TRUE),
      .groups = "drop"
    )
  
  # -------------------------------
  # 7. Combine everything + dark score
  # -------------------------------
  final <- tibble(gene_id = gene_list) %>%
    left_join(egg_gene,      by = "gene_id") %>%
    left_join(ips_gene,      by = "gene_id") %>%
    left_join(tri_gene,      by = "gene_id") %>%
    left_join(genome_best,   by = "gene_id") %>%
    left_join(sig_gene,      by = "gene_id") %>%
    left_join(tmh_gene,      by = "gene_id") %>%
    mutate(
      has_PFAM   = !is.na(ips_PFAM) | !is.na(tri_PFAM),
      has_IPR    = !is.na(ips_IPR),
      has_GO     = !is.na(ips_GO) | !is.na(tri_GO) | !is.na(genome_GO),
      has_KEGG   = !is.na(tri_KEGG) | !is.na(egg_KO) | !is.na(egg_KEGGmap),
      has_BLAST  = !is.na(tri_BLAST) | !is.na(egg_ortholog),
      has_UniProt= !is.na(genome_UniProt),
      
      # Count how many annotation sources contributed something
      annotation_sources_count =
        (!is.na(ips_IPR)) +
        (!is.na(tri_BLAST)) +
        (!is.na(egg_ortholog)) +
        (!is.na(genome_GO)),
      
      # Dark gene score
      dark_score = (!has_PFAM) + (!has_IPR) + (!has_GO) +
                   (!has_KEGG) + (!has_BLAST) + (!has_UniProt)
    )
  
  return(final)
}
```

```{r}
tier2_gene_annotations <- extract_annotations(
  gene_list = tier2_genes,
  eggnog = eggnog,
  ips = ips,
  trinotate = trinotate_report,
  signalp = signalp,
  tmhmm = tmhmm_gff3,
  genome = genome_based
)
print(head(tier2_gene_annotations, 5), width = Inf)
```
```{r}
sum(tier2_genes %in% genome_based$gene_id)
```
```{r}
extract_genome_annotations <- function(gene_list, genome_based) {
  
  genome_based %>%
    filter(gene_id %in% gene_list) %>%
    
    # keep only the transcript with highest coverage
    group_by(gene_id) %>%
    slice_max(`Coverage_%`, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    
    transmute(
      gene_id,
      transcript       = transcript_id,
      genome_gene_id   = Gene_ID,          # NEW
      genome_coverage  = `Coverage_%`,
      genome_GO        = ifelse(GO_Terms == "" | GO_Terms == "NA", NA, GO_Terms),
      genome_UniProt   = ifelse(UniProt_ID == "" | UniProt_ID == "NA", NA, UniProt_ID),
      genome_Dbxref    = ifelse(Dbxref == "" | Dbxref == "NA", NA, Dbxref)
    )
}
```

```{r}
tier2_genome <- extract_genome_annotations(tier2_genes, genome_based)
tier2_genome
```

```{r}
# ---- Integrate SignalP-6.0 into Trinotate report ----
## SignalP predictions
signalp <- read_tsv(
  "results/annotation/trinotate/signalp/signalp_merged.output.gff3",
  col_names = FALSE,
  comment = "##",
  show_col_types = FALSE
)
## Assign the correct column names
colnames(signalp) <- c(
  "seqid", "source", "type", "start", "end",
  "score", "strand", "phase", "attributes"
)

# ---- SignalP ----
signalp <- signalp %>%
  mutate(
    # Extract the transcript ID from the seqid field
    transcript_id = sub("\\.p[0-9]+.*$", "", seqid),

    # Derive gene_id (not used for merging)
    gene_id = sub("_i.*", "", transcript_id)
  )

# 3. Identify transcripts with a signal peptide (YES)
signalp_hits <- signalp %>%
  filter(type == "signal_peptide") %>%
  distinct(transcript_id) %>%      # <-- ONLY transcript_id
  mutate(SignalP = "YES")

# 4. Remove old Trinotate SignalP column
trinotate_report <- trinotate_report %>% select(-SignalP)

# 5. Build full transcript list from Trinotate
all_transcripts <- trinotate_report %>%
  select(transcript_id) %>%        # <-- ONLY transcript_id
  distinct()

# 6. Create complete YES/NO table (join ONLY on transcript_id)
signalp_full <- all_transcripts %>%
  left_join(signalp_hits, by = "transcript_id") %>%   # <-- ONLY transcript_id
  mutate(SignalP = ifelse(is.na(SignalP), "NO", SignalP))

# 7. Merge into Trinotate report
trinotate_report <- trinotate_report %>%
  left_join(signalp_full, by = "transcript_id")       # <-- ONLY transcript_id
```

```{r}
# ---- Load DeepTMHMM GFF3 ----
tmhmm <- read_tsv(
  "results/annotation/trinotate/tmhmm/tmhmm_clean.gff3",
  col_names = FALSE,
  comment = "##",
  show_col_types = FALSE
)

# Assign column names
colnames(tmhmm) <- c(
  "seqid", "source", "type", "start", "end",
  "score", "strand", "phase", "attributes"
)

# Clean transcript IDs (remove .p1, .p2, ...)
tmhmm <- tmhmm %>%
  mutate(
    transcript_id = sub("\\.p[0-9]+$", "", seqid)
  )

# ---- Identify transcripts with ≥1 TM helix ----
tmhmm_hits <- tmhmm %>%
  filter(type == "TMhelix") %>%
  distinct(transcript_id) %>%
  mutate(TMHMM = "YES")

# ---- Build full transcript list ----
all_transcripts <- trinotate_report %>%
  select(transcript_id) %>%
  distinct()

tmhmm_full <- all_transcripts %>%
  left_join(tmhmm_hits, by = "transcript_id") %>%
  mutate(TMHMM = ifelse(is.na(TMHMM), "NO", TMHMM))

# ---- Merge TMHMM into Trinotate report ----
trinotate_report <- trinotate_report %>%
  { if ("TMHMM" %in% colnames(.)) select(., -TMHMM) else . } %>% 
  left_join(tmhmm_full, by = "transcript_id")
```



```{r}
extract_trinotate_annotations <- function(gene_list, trinotate_report) {
  
  clean_field <- function(x) {
    x <- gsub("^;|;$", "", x)
    x <- gsub(";;+", ";", x)
    x[x == ""] <- NA
    x
  }
  
  trinotate_report %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      BLASTX = clean_field(
        paste0(
          na.omit(unique(sprot_Top_BLASTX_hit[sprot_Top_BLASTX_hit != "."])),
          collapse = ";"
        )
      ),
      
      BLASTP = clean_field(
        paste0(
          na.omit(unique(sprot_Top_BLASTP_hit[sprot_Top_BLASTP_hit != "."])),
          collapse = ";"
        )
      ),
      
      PFAM = clean_field(
        paste0(
          na.omit(unique(Pfam[Pfam != "."])),
          collapse = ";"
        )
      ),
      
      GO = clean_field(
        paste0(
          na.omit(unique(c(
            gene_ontology_BLASTX[gene_ontology_BLASTX != "."],
            gene_ontology_BLASTP[gene_ontology_BLASTP != "."],
            gene_ontology_Pfam[gene_ontology_Pfam != "."]
          ))),
          collapse = ";"
        )
      ),
      
      SignalP = ifelse(any(SignalP == "YES"), "YES", "NO"),
      TMHMM   = ifelse(any(TMHMM == "YES"), "YES", "NO"),
      
      .groups = "drop"
    )
}
```

```{r}
tier2_trinotate <- extract_trinotate_annotations(
  gene_list = tier2_genes,
  trinotate_report = trinotate_report
)

tier2_trinotate
```

```{r}
# ============================================================
# CHECK ALL COMBINATIONS OF TRINOTATE vs GENOME ANNOTATION
# FOR TIER 2 GENES
# ============================================================

# 1. Identify Tier 2 genes with ANY Trinotate annotation
tier2_trinotate_nonempty <- tier2_trinotate %>%
  filter(
    !is.na(BLASTX) |
    !is.na(BLASTP) |
    !is.na(PFAM)   |
    !is.na(GO)     |
    SignalP == "YES"   |
    TMHMM  == "YES"
  )

# 2. Identify Tier 2 genes with NO Trinotate annotation
tier2_trinotate_empty <- tier2_trinotate %>%
  filter(
    is.na(BLASTX) &
    is.na(BLASTP) &
    is.na(PFAM)   &
    is.na(GO)     &
    SignalP == "NO"   &
    TMHMM  == "NO"
  )

# 3. Identify Tier 2 genes with genome annotation
tier2_genome_nonempty <- tier2_genome   # already filtered by gene_id

# 4. Identify Tier 2 genes with NO genome annotation
tier2_genome_empty <- tibble(gene_id = tier2_genes) %>%
  anti_join(tier2_genome, by = "gene_id")

# ============================================================
# COMBINATIONS
# ============================================================

# A) Trinotate ONLY (Trinotate yes, Genome no)
trinotate_only <- tier2_trinotate_nonempty %>%
  anti_join(tier2_genome_nonempty, by = "gene_id")

# B) Genome ONLY (Genome yes, Trinotate no)
genome_only <- tier2_genome_nonempty %>%
  anti_join(tier2_trinotate_nonempty, by = "gene_id")

# C) BOTH Trinotate AND Genome
both_trinotate_genome <- tier2_trinotate_nonempty %>%
  inner_join(tier2_genome_nonempty, by = "gene_id")

# D) NEITHER (true dark genes)
neither_source <- tier2_trinotate_empty %>%
  inner_join(tier2_genome_empty, by = "gene_id")

# ============================================================
# Return everything as a list
# ============================================================

tier2_annotation_diagnostics <- list(
  trinotate_only = trinotate_only,
  genome_only = genome_only,
  both_sources = both_trinotate_genome,
  neither_source = neither_source
)

tier2_annotation_diagnostics
```
```{r}
extract_ips_annotations <- function(gene_list, ips) {
  
  clean_field <- function(x) {
    x <- gsub("[|,]", ";", x)      # normalize | and , to ;
    x <- gsub("^;|;$", "", x)      # trim leading/trailing ;
    x <- gsub(";;+", ";", x)       # collapse multiple ;;
    x[x == ""] <- NA               # empty -> NA
    x
  }
  
  ips %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      ips_IPR  = clean_field(
        paste0(na.omit(unique(X12[X12 != "-"])), collapse = ";")
      ),
      ips_PFAM = clean_field(
        paste0(na.omit(unique(X5[X5 != "-"])), collapse = ";")
      ),
      ips_GO   = clean_field(
        paste0(na.omit(unique(X14[X14 != "-"])), collapse = ";")
      ),
      .groups = "drop"
    )
}
```

```{r}
tier2_ips <- extract_ips_annotations(
  gene_list = tier2_genes,
  ips = ips 
)

tier2_ips
```

```{r}
extract_eggnog_annotations <- function(gene_list, eggnog) {
  
  clean_field <- function(x) {
    x <- unique(na.omit(x))
    x <- x[x != "-" & x != ""]
    if (length(x) == 0) return(NA)
    x <- gsub(",", ";", x)   # normalize EggNOG comma lists
    paste(x, collapse = ";")
  }
  
  eggnog %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      egg_ortholog = clean_field(seed_ortholog),
      egg_desc     = clean_field(Description),
      egg_COG      = clean_field(COG_category),
      egg_KO       = clean_field(KEGG_ko),
      egg_KEGGmap  = clean_field(KEGG_Pathway),
      egg_PFAM     = clean_field(PFAMs),
      egg_GO       = clean_field(GOs),     
      .groups = "drop"
    )
}
```

```{r}
eggnog_ann <- extract_eggnog_annotations(tier2_genes, eggnog)
eggnog_ann
```

```{r}
head(signalp)
colnames(signalp)
head(tmhmm_gff3)
signalp %>% filter(gene_id == "TRINITY_GG_1002_c11_g1")
```

```{r}
extract_signalp_annotations <- function(gene_list, signalp) {
  signalp %>%
    filter(gene_id %in% gene_list) %>%
    group_by(gene_id) %>%
    summarise(
      signalp = ifelse(any(Prediction == "SP(Sec/SPI)"), "YES", "NO"),
      .groups = "drop"
    )
}
```

```{r}
tier2_signalp <- extract_signalp_annotations(
  gene_list = tier2_genes,
  signalp = signalp
)

tier2_signalp
```

```{r}

```

